### 2019-03-30 doc/paper/paper.tex
Chaos section written

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
%\usepackage{lipsum}

\title{Time Series Analysis of Chaotic Dynamical Systems with Recurrent Neural Networks}

\author{
  Lucas Wilson \\
  Undergraduate: Mathematics, Computer Science \\
  Colorado State University\\
  Fort Collins, CO 80523 \\
  \texttt{lkwilson96@gmail.com} \\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

\begin{document}
\maketitle

\begin{abstract}
This is a paper about chaos.
\end{abstract}

%%% INTRODUCTION %%%
\section{Introduction}

This is a paper about chaos.

%%% CHAOS %%%
\section{Chaos}

In 1963, Edward N. Lorenz published an article researching a simplified system of 
ordinary differential equations modeling a convective system \cite{lorenz1963deterministic}. His 
research popularized the possibility of a system being highly sensitive to 
initial conditions, i.e., chaos theory; while he wasn't the first to discover 
this phenomenon, he is considered to be the "official discoverer of chaos 
theory" \cite{oestreicher2007history}.

When forecasting, the goal is to have accurate predictions. However, 
a chaotic system's sensitivity to initial conditions implies that a small 
perturbation as a result of error will produce incorrect solutions.
Further, given a periodic
solution, we hope that a small perturbation, from error in measurement or from 
floating-point errors in numerical calculations, doesn't affect the solution or 
at least produces a quasi-periodic solution close to the periodic one. 
(A solution or trajectory $F$ 
is quasi-periodic if and only if for all $t$ and for some $\tau$, $F(t+\tau)$ 
is arbitrarily close to $F(t)$ \cite{lorenz1963deterministic}.)

An example of this is the trajectories of the sun and planets in our solar 
system. Newton's equations can be used to create a system of equations to model 
the bodies of mass (known as the n-body problem). However, with the existence of
other bodies of mass from other solar systems within the universe, we will have 
small perturbations introduced into the system. If the system is chaotic, then 
over time, the actual trajectories will diverge from our model's 
\cite{oestreicher2007history}. The amount of time for the trajectories to differ
by a factor of 10 is known as the Lyapunov constant 
\cite{oestreicher2007history}.

In terms of forecasting, the Lyapunov time constant can be
used to know how much time
will pass before it's very likely that the model has become incorrect since
errors introduced to the calculated solution will have diverged too far from
the actual solution. 
Researchers have used it to evaluate the effectiveness of forecasting models on
chaotic problems \cite{pathak2018model}. Predicting the solution of
a chaotic model over an interval of time of can be easy if that interval is well
below the Lyapunov constant.

One of the defining characteristics of chaos is that the error from 
perturbations grows exponentially \cite{oestreicher2007history}. At the time of 
Lorenz' paper, unstable 
error was not well understood, and this led to Lorenz' surprising discovery 
\cite{oestreicher2007history}. Lorenz reported that a single iteration in 
calculating the solution to the dynamical system took 
approximately one second \cite{lorenz1963deterministic}. In his paper, he shows 
one of his calculations having about 3000 iterations, which must have taken 
around 50 minutes. Given the amount of time it takes to perform the calculation,
it's tempting to continue the calculation from the output of the previous 
calculation. However, the story of the discover is that while the computer used 
6 digit accuracy, it output only 3 digits, so when the simulation was continued 
by Lorenz with the less accurate measurements, he found the results to be very 
different \cite{oestreicher2007history}.

Chaos can appear in many different situations. Lorenz's paper demonstrates chaos
in a dynamical system useful for forecasting convection in the atmosphere or 
liquids \cite{lorenz1963deterministic}. "Chaos theory has a few applications for
modeling endogenous biological rhythms such as heart rate, brain functioning, 
and biological docks" \cite{oestreicher2007history}. It can also appear in 
solutions to Hamiltonian problems such as the 3 body problem (as mentioned 
before) and the double pendulum problem (as I will show later). Understanding
the predictability of chaos is very useful to many different fields.

\subsection{Lyapunov Constant}

%%% DYNAMICAL SYSTEMS %%%
\section{Dynamical Systems}

\subsection{Lorenz System}
\subsection{Double Pendulum}
\subsection{3-Body Problem}

%%% MODELS %%%
\section{Models}

\subsection{ARIMA}
\subsection{Recurrent Neural Networks}
\subsection{Echo State Networks}
\subsection{LTSM}

%%% ANALYSIS %%%
\section{Analysis}

\subsection{Method}
\subsection{Results}

%% CONCLUSION %%%
\section{Conclusion}

\bibliography{references}
\bibliographystyle{unsrt}

\end{document}

% Useful LaTeX commands

%\section{Headings: first level}
%\subsection{Headings: second level}
%\subsubsection{Headings: third level}

%\keywords{First keyword \and Second keyword \and More}

% \paragraph{Paragraph}

%% \begin{figure}
%%   \centering
%%   \fbox{\rule[-.5cm]{4cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
%%   \caption{Sample figure caption.}
%%   \label{fig:fig1}
%% \end{figure}

### 2019-03-31 src/echonn/sys/double_pendulum_animator.py
modify doub pend after paper instead of wikipedia

@@ -18,7 +18,8 @@ class DoublePendulumAnimator(Animator):
         self.runs = runs
         limit_size = -1
         for run in runs:
-            limit_size = max(limit_size, 2*run['system'].l+limit_margin)
+            l = run['system'].l1 + run['system'].l2
+            limit_size = max(limit_size, l+limit_margin)
         self.limits = [-limit_size, limit_size, -limit_size, limit_size]
         self.speed = speed
 
@@ -38,11 +39,13 @@ class DoublePendulumAnimator(Animator):
 
     def get_endpoints(self, run):
         inner_theta, outer_theta, _, _ = run['results'].y
+        l1 = run['system'].l1
+        l2 = run['system'].l2
         inner_theta = inner_theta - pi/2
         outer_theta = outer_theta - pi/2
-        inner_pos = np.array([np.cos(inner_theta), np.sin(inner_theta)])
-        outer_relative_pos = np.array([np.cos(outer_theta),
-                                       np.sin(outer_theta)])
+        inner_pos = l1*np.array([np.cos(inner_theta), np.sin(inner_theta)])
+        outer_relative_pos = l2*np.array([np.cos(outer_theta),
+                                          np.sin(outer_theta)])
         outer_pos = inner_pos + outer_relative_pos
         return inner_pos, outer_pos
 

### 2019-03-31 src/echonn/sys/double_pendulum_system.py

@@ -13,16 +13,42 @@ class DoublePendulumSystem(DynamicalSystem):
     Equal mass and equal length
     """
 
-    def __init__(self, m=1, l=1):
+    def __init__(self, m1=1, m2=1, l1=1, l2=1):
         super().__init__(4, 'BDF')
-        self.m = m
-        self.l = l
+        self.m1 = m1
+        self.m2 = m2
+        self.l1 = l1
+        self.l2 = l2
 
-        self._theta_pre = 6 / (m*l**2)
-        self._gl = g / l
+        # for wikipedia formula
+        #self._theta_pre = 6 / (m*l**2)
+        #self._gl = g / l
+        #self._moment_pre = -.5*m*l**2
 
-        self._moment_pre = -.5*m*l**2
+    def lagrange(self, v):
+        pass
 
+    def fun(self, t, v):
+        sin = np.sin
+        cos = np.cos
+        m1, m2 = self.m1, self.m2
+        l1, l2 = self.l1, self.l2
+        a1, a2, w1, w2 = v
+
+        a1p = w1
+        a2p = w2
+        w1_num_t1 = sin(a1-a2)*(l1*cos(a1-a2)*w1**2 + w2**2)
+        w1_den_t1 = 2*l2*(1 + m1 - cos(a1-a2)**2)
+        w1_num_t2 = (1 + 2*m1)*sin(a1) + sin(a1-2*a2)
+        w1_den_t2 = l1 * (1 + m1 - cos(a1-a2)**2)
+        w1p = w1_num_t1 / w1_den_t1 - w1_num_t2 / w1_den_t2
+
+        w2_num = (1 + m1)*(cos(a1)+l1*w1**2) + cos(a1-a2)*w2**2
+        w2_den = 1 + m1 - cos(a1-a2)**2
+        w2p = sin(a1 - a2) * w2_num / w2_den
+        return a1p, a2p, w1p, w2p
+
+    """ from wikipedia
     def fun(self, t, v):
         a1, a2, p1, p2 = v
 
@@ -41,3 +67,4 @@ class DoublePendulumSystem(DynamicalSystem):
         p2p = self._moment_pre * (-first_term + self._gl*np.sin(a2))
 
         return a1p, a2p, p1p, p2p
+    """

### 2019-03-31 src/test/sys/test_double_pendulum_animator.py

@@ -17,6 +17,15 @@ class TestDoubPenAnim(unittest.TestCase):
         self.render_long = False  # slows test a lot
         self.render = False  # slows test
 
+    def testWeirdLenSys(self):
+        sys = DoublePendulumSystem(1, 2, .5, 1.5)
+        slv = SystemSolver(sys)
+        tf = 3
+        run1 = slv.run([0, tf], [pi, 1, 1, 0])
+        run2 = slv.run([0, tf], [pi, 1.001, 1, 0])
+        if self.render:
+            self.runRender([run1, run2], 'odd_pend')
+
     def testLongRun(self):
         tf = 50
         run1 = self.pendulum_solver.run([0, tf], [pi, 1, 1, 0])

### 2019-03-31 doc/paper/scripts/DoubPendChaosChart.py
add data generator scripts for paper

@@ -1,0 +1,26 @@
+import numpy as np
+import os
+from echonn.sys import DoublePendulumSystem, SystemSolver
+import matplotlib.pyplot as plt
+from scipy.constants import pi
+
+if __name__ == "__main__":
+    sys = DoublePendulumSystem()
+    slv = SystemSolver(sys)
+    thetas = np.arange(0.001, pi+0.0001, pi/20)
+    out_img = os.path.join('..', 'images', 'chaos_vs_energy_in_doub_pend.png')
+    lces = []
+    for theta in thetas:
+        lce, _ = slv.get_lce(T=200, y0=[theta, 0, 0, 0])
+        lces.append(lce)
+    plt.title('Largest LCE vs Inner Theta IC')
+    plt.xlabel('Inner Theta')
+    plt.ylabel('Largest LCE')
+    plt.plot(180*thetas/pi, lces)
+    try:
+        if os.path.exists(out_img):
+            os.remove(out_img)
+    except:
+        pass  # don't worry about it
+    plt.savefig(out_img)
+    plt.show(True)

### 2019-03-31 doc/paper/scripts/LorenzLCEs.py

@@ -1,0 +1,30 @@
+import numpy as np
+from echonn.sys import LorenzSystem, SystemSolver
+
+if __name__ == "__main__":
+    data = [
+        [16, 45.92, 4, 1.50255],
+        [16, 40, 4, 1.37446],
+        [10, 28, 8/3, 0.90566],
+    ]
+    for sigma, rho, beta, lambda_ in data:
+
+        lces = []
+        for _ in range(10):
+            sys = LorenzSystem(sigma, rho, beta)
+            slv = SystemSolver(sys)
+
+            lce, _ = slv.get_lce(T=150)
+            print('lce:', lce)
+            lces.append(lce)
+        res = {}
+        res['beta'] = beta
+        res['rho'] = rho
+        res['sigma'] = sigma
+        res['lambda'] = lambda_
+        res['mean'] = np.mean(lces)
+        res['std'] = np.std(lces)
+        res['error'] = res['mean'] - lambda_
+        res['relative error'] = res['error'] / lambda_
+        print(res)
+        print()

### 2019-03-31 doc/paper/paper.tex
Updates from Overleaf

@@
 \usepackage{arxiv}
-
 \usepackage[utf8]{inputenc} % allow utf-8 input
@@
 \usepackage{microtype}      % microtypography
-%\usepackage{lipsum}
+\usepackage{amsmath}
+\usepackage{lipsum}
+\usepackage{caption}
+\usepackage{float}
+
+\newcommand{\der}[2][t]{\frac{\mathrm{d}#2}{\mathrm{d}#1}}
 
@@
   \texttt{lkwilson96@gmail.com} \\
-  %% \AND
-  %% Coauthor \\
-  %% Affiliation \\
-  %% Address \\
-  %% \texttt{email} \\
-  %% \And
-  %% Coauthor \\
-  %% Affiliation \\
-  %% Address \\
-  %% \texttt{email} \\
-  %% \And
-  %% Coauthor \\
-  %% Affiliation \\
-  %% Address \\
-  %% \texttt{email} \\
 }
@@
 This is a paper about chaos.
+% TODO: mention \texttt{echonn}
 
@@
 \section{Chaos}
+
+\subsection{History}
 
@@
 over time, the actual trajectories will diverge from our model's 
-\cite{oestreicher2007history}. The amount of time for the trajectories to differ
-by a factor of 10 is known as the Lyapunov constant 
 \cite{oestreicher2007history}.
-
-In terms of forecasting, the Lyapunov time constant can be
-used to know how much time
-will pass before it's very likely that the model has become incorrect since
-errors introduced to the calculated solution will have diverged too far from
-the actual solution. 
-Researchers have used it to evaluate the effectiveness of forecasting models on
-chaotic problems \cite{pathak2018model}. Predicting the solution of
-a chaotic model over an interval of time of can be easy if that interval is well
-below the Lyapunov constant.
 
@@
 perturbations grows exponentially \cite{oestreicher2007history}. At the time of 
-Lorenz' paper, unstable 
-error was not well understood, and this led to Lorenz' surprising discovery 
+Lorenz's paper, unstable 
+error was not well understood, and this led to Lorenz's surprising discovery 
 \cite{oestreicher2007history}. Lorenz reported that a single iteration in 
@@
 
+In terms of forecasting, it's useful to know how chaotic a dynamical system is. 
+To measure this, we will measure the rate at which nearby trajectories 
+diverge. Constants describing the divergence are known as the Lyapunov 
+Characteristic Exponents (LCEs) \cite{sandri1996numerical}. There is an LCE for 
+each dimension, and since they all represent exponential growth. The error 
+growing at the rate of the largest LCE will dominate over the others.
+
+% TODO: talk about attractors
+
+Given a small perturbation $\epsilon$ to a dynamical system in the direction of 
+the largest LCE, we can approximate the growth of the perturbation by the 
+equation $\epsilon(\Delta t) = \epsilon_0 e^{\lambda \Delta t}$ where $\lambda$ 
+is the largest LCE
+\cite{bezruchko2010extracting}. For an LCE of zero, then the error doesn't grow
+exponentially, and thus, there is no chaos. Error created
+from smaller LCEs is negligible. The approximate time scale where 
+perturbations become large is going to be proportional to $1 / \lambda$. This is
+known as the Lyapunov time \cite{bezruchko2010extracting}. For our solar
+system, the Lyapunov time is 10,000,000 years \cite{oestreicher2007history}.
+Researchers have used it to evaluate the effectiveness of forecasting models on
+chaotic problems \cite{pathak2018model}. Predicting the solution of
+a chaotic model over an interval of time which is multiple times the size of the
+Lyapunov time shows the predictability of a model.
+
+In order to find the Lyapunov time, we only need to solve for the largest LCE. 
+To find this, we will use a method detailed in \cite{viswanath1998lyapunov}.
+
+% TODO: Lyapunov calculation
+
 %%% DYNAMICAL SYSTEMS %%%
@@
 
+Three dynamical systems will be used to demonstrate the predictability of 
+chaotic systems: the Lorenz system, Poincaré's 3-body 
+problem solution, and the double pendulum system.
+The Poincaré system is the one of the first (possibly the first), and Lorenz
+is a classical example. The double pendulum is commonly referred to as chaotic, 
+and I will show this later.
+
 \subsection{Lorenz System}
+
+The Lorenz system was originally based on a system of equations modeling 
+convection created by Saltzman \cite{lorenz1963deterministic} 
+\cite{saltzman1962finite}. Simplified by Lorenz, it is defined as follows
+\cite{lorenz1963deterministic}:
+
+\begin{align}
+    \der{x} &= \sigma (y - x), \nonumber \\
+    \der{y} &= x (\rho - z) - y, \nonumber \\
+    \der{z} &= x y - \beta z. \label{eq:lorenz_equation}
+\end{align}
+
+Not all parameters for the Lorenz equation will produce chaotic behavior. In 
+Lorenz's article, he uses the parameter values $\sigma=10$, $\beta=8/3$, and
+$\rho=28$ \cite{lorenz1963deterministic}, and it has a maximum LCE of $0.90566$ 
+\cite{viswanath1998lyapunov}. There are other parameters which
+produce chaotic results outlined in Table \ref{table:lorenz_params}. Also shown 
+in Table \ref{table:lorenz_params} are the values calculated by my Python
+module \texttt{echonn} are listed. These values are not as accurate, but 
+demonstrate the sufficient accuracy of \texttt{echonn}.
+
+\begin{table}[H]
+    \centering
+    \begin{tabular}{|l|l|l|l|l|l|l|}
+         % TODO: \lambda_1?
+         % TODO: calculated values
+        \hline
+        $\sigma$ & $\rho$ & $\beta$ & actual $\lambda$ & calculated $\lambda$ & relative error of calculated $\lambda$ \\
+        \hline \hline
+        16 & 45.92 & 4 & 1.50255 & & \\
+        16 & 40 & 4 & 1.37446 & & \\
+        10 & 28 & 8/3 & 0.90566 & & \\
+        \hline
+    \end{tabular}
+    \caption{
+        Lorenz Parameters and Largest Lyapunov Exponent
+        \cite{viswanath1998lyapunov}
+    }
+    \label{table:lorenz_params}
+\end{table}
+
+\subsection{3-Body Problem}
+
 \subsection{Double Pendulum}
-\subsection{3-Body Problem}
+
+The double pendulum is commonly cited to be chaotic 
+\cite{stachowiak2006numerical} \cite{levien1993double}. Using the system 
+of equations derived by Stachowiak \cite{stachowiak2006numerical}, the system is
+defined as follows:
+
+\begin{align}
+    \der{\theta_1} &= \omega_1, \nonumber \\
+    \der{\theta_2} &= \omega_2, \nonumber \\
+    \der{\omega_1} &= 
+    \frac{
+        \sin(\theta_1 - \theta_2) \lbrack
+            l_1 \cos(\theta_1 - \theta_2) \omega_1^2 + \omega_2^2
+        \rbrack
+    }{
+        2 l_1 \lbrack
+            1 + m_1 - \cos^2(\theta_1 - \theta_2)
+        \rbrack
+    }
+    -
+    \frac{
+        (1 + 2 m_1) \sin \theta_1 + \sin(\theta_1 - 2 \theta_2)
+    }{
+        l_1 \lbrack
+            1 + m_1 - \cos^2(\theta_1 - \theta_2)
+        \rbrack
+    }
+    , \nonumber \\
+    \der{\omega_2} &= \sin (\theta_1 - \theta_2) 
+    \frac{
+        (1+m_1) (\cos \theta_1 + l_1 \omega_1^2)
+        +
+        \cos(\theta_1 - \theta_2) \omega_2^2
+    }{
+        1 + m_1 - \cos^2(\theta_1 - \theta_2)
+    }. \label{eq:doub_pen}
+\end{align}
+
+While the system is a Hamiltonian system, Stachowiak decided to define the first order 
+derivative components as angular velocity as opposed to momentum. This won't affect
+anything regarding the purpose of this paper.
+
+Interestingly, the 
+degree of chaos is dependent on the initial conditions \cite{levien1993double}. 
+In the experiment, the initial conditions had angular velocities of zero, the outer 
+pendulum was straight down, and the inner pendulum varied from 0 degrees to 180 degrees 
+\cite{levien1993double}; the results showed that the LCE increased with theta.
+
+Repeating the same experiment, but with different parameters, the results are similar (see 
+Figure \ref{fig:doub_pend_energy}).
+
+\begin{figure}[H]
+    \centering
+    %\includegraphics[.5\linewidth]{images/chaos_vs_energy_in_doub_pend.png}
+    \caption{Largest Lyapunov Characteristic Exponent vs Inner Theta Initial Condition}
+    \label{fig:doub_pend_energy}
+\end{figure}
 
@@
 \subsection{ARIMA}
-\subsection{Recurrent Neural Networks}
+%\subsection{Recurrent Neural Networks}
 \subsection{Echo State Networks}
-\subsection{LTSM}
+%\subsection{LTSM}
 
@@
 %% \end{figure}
+
+%%\author{
+  %%Lucas Wilson \\
+  %%Undergraduate: Mathematics, Computer Science \\
+  %%Colorado State University\\
+  %%Fort Collins, CO 80523 \\
+  %%\texttt{lkwilson96@gmail.com} \\
+  %% \AND
+  %% Coauthor \\
+  %% Affiliation \\
+  %% Address \\
+  %% \texttt{email} \\
+  %% \And
+  %% Coauthor \\
+  %% Affiliation \\
+  %% Address \\
+  %% \texttt{email} \\
+  %% \And
+  %% Coauthor \\
+  %% Affiliation \\
+  %% Address \\
+  %% \texttt{email} \\
+%%}

### 2019-04-02 doc/paper/scripts/RenderR3Bod.py
add restricted 3 body prob

@@ -1,0 +1,26 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from echonn.sys import SystemSolver, RestrictedCircular3Body, LyapunovSystem
+
+if __name__ == "__main__":
+    mu = 0.5
+    sys = RestrictedCircular3Body(body_ratio=mu)
+    lce = LyapunovSystem(sys)
+
+    #init = [1, -1, .1, .1]
+    init = np.array(1000*np.random.rand(4), dtype=int)/1000
+    slv = SystemSolver(sys)
+    #run = slv.run([0, 10], init, max_step=0.001)
+    T = 50
+    lce, run = slv.get_lce(T=T, y0=init)
+    #lce, run = slv.get_lce(T=T)
+    mat = run['results'].y[4:, -1].reshape(4, 4)
+    run['results'].y = run['results'].y[[0, 2], :]
+    slv.plot2d(run)
+    plt.scatter([-sys.alpha, sys.mu], [0, 0])
+    print(init)
+    print(mat)
+    print('lce:', lce)
+    #plt.xlim((-15, 15))
+    #plt.ylim((-15, 15))
+    plt.show(True)

### 2019-04-02 src/echonn/sys/__init__.py

@@ -6,3 +6,4 @@ from .double_pendulum_system import DoublePendulumSystem
 from .n_body_system import NBodySystem
 from .animator import Animator
 from .lyapunov_system import LyapunovSystem
+from .restricted_circular_3_body import RestrictedCircular3Body

### 2019-04-02 src/echonn/sys/restricted_circular_3_body.py

@@ -1,0 +1,34 @@
+import numpy as np
+from .system import DynamicalSystem
+
+
+class RestrictedCircular3Body(DynamicalSystem):
+    def __init__(self, body_ratio=0.5, method='BDF'):
+        super().__init__(4, method)
+        if body_ratio < 0 or 1 < body_ratio:
+            raise ValueError('body_ratio should be in range [0, 1]')
+        self.mu = body_ratio
+        self.alpha = 1 - body_ratio
+
+    def r1(self, t, v):
+        mu = self.mu
+        x1, _, y1, _ = v
+        return np.sqrt((x1-mu)**2 + y1**2)
+
+    def r2(self, t, v):
+        a = self.alpha
+        x1, _, y1, _ = v
+        return np.sqrt((x1+a)**2 + y1**2)
+
+    def fun(self, t, v):
+        a = self.alpha
+        mu = self.mu
+        r1 = self.r1(t, v)
+        r2 = self.r2(t, v)
+
+        x1, x2, y1, y2 = v
+        x1p, y1p = x2, y2
+        x2p = 2*y2 + x1 - a/r1**3*(x1-mu) - mu/r2**3*(x1+a)
+        y2p = -2*x2 + (1 - a/r1**3 - mu/r2**3)*y1
+
+        return x1p, x2p, y1p, y2p

### 2019-04-02 src/echonn/sys/system_solver.py

@@ -100,7 +100,7 @@ class SystemSolver:
         run = lce_solver.run([tspan[-1], T], v, **kwargs)
         # if failed, raise exception
         if run['results'].status != 0:
-            raise Exception(run['results'].message)
+            print('Warning:', run['results'].message)
         vf = run['results'].y[:, -1]
         Df_y0 = v[sys_dim:].reshape(sys_dim, sys_dim)
         lce = self.calc_lce(Df_y0, T)

### 2019-04-02 src/test/sys/test_lce_calc.py

@@ -1,6 +1,6 @@
 import unittest
 import numpy as np
-from echonn.sys import LorenzSystem, DoublePendulumSystem, SystemSolver
+from echonn.sys import LorenzSystem, DoublePendulumSystem, SystemSolver, CircleSystem
 import matplotlib.pyplot as plt
 from scipy.constants import pi
 from .sys_util import clearFigs
@@ -51,6 +51,12 @@ class TestLCECalc(unittest.TestCase):
         sys = DoublePendulumSystem()
         slv = SystemSolver(sys)
         slv.get_lce(T=2, y0=[1.8, 1.8, 0, 0])
+
+    def testCircleLCE(self):
+        sys = SystemSolver(CircleSystem())
+        lce, run = sys.get_lce()
+        self.assertAlmostEqual(lce, 0, places=4)
+        # print(lce)
 
     @unittest.skip
     def testLCEPendulumLong(self):

### 2019-04-02 src/test/sys/test_restrict_3body.py

@@ -1,0 +1,15 @@
+import unittest
+import numpy as np
+import matplotlib.pyplot as plt
+from echonn.sys import SystemSolver, RestrictedCircular3Body
+
+
+class TestModel(unittest.TestCase):
+    def testExample(self):
+        sys = RestrictedCircular3Body()
+        slv = SystemSolver(sys)
+        init = [1, -1, 1, -1]
+        run = slv.run([0, 5], init)
+        run['results'].y = run['results'].y[:2, :]
+        slv.plot2d(run)
+        # plt.show(True)

### 2019-04-02 doc/paper/paper.tex
Update more sections of paper
Updates from Overleaf

@@
 \usepackage{float}
+\usepackage{graphicx}
+
+% the comment 'DATA POINT' will mark where data was calculated and then inserted into
+% the paper.
 
@@
 % TODO: Lyapunov calculation
+% TODO: Test with Circle yielding zero
 
@@
 
+% TODO: numerical solutions section
+
 \subsection{Lorenz System}
+
+% TODO: Image of Lorenz System
 
@@
         \hline \hline
-        16 & 45.92 & 4 & 1.50255 & & \\
+        16 & 45.92 & 4 & 1.50255 & & \\ % DATA POINTS
         16 & 40 & 4 & 1.37446 & & \\
@@
 
+The 3-body problem, or the generalized n-body problem, is a famous problem which many have 
+analyzed; most notably, Herni Poincaré wrote many papers developing the area significantly
+\cite{chenciner2000remarkable}. One of Poincaré's solutions, known as the circular 
+restricted 3-body problem, demonstrates chaotic properties \cite{oestreicher2007history}. 
+Poincaré performs several reductions of the original 3-body problem in order to come to a 
+much more useful form of the problem. The mass of one of the objects is assumed to be 
+negligible. This is the case for Sun-Earth-Moon orbits. The moon has little effect on the 
+orbit of the sun.
+Another aspect is the rotating reference frame.
+Since the mass of one of the bodies is negligible, the problem simplifies to first solving a two body problem.
+While much easier to solver, Poincaré went further to stop the motion of the two planets by using a rotating reference frame.
+Rotating with the planets, the problem can be modeled seemingly by two static planets and a third mass-less planet orbiting around those in a two dimensional plane.
+
+% norm from https://tex.stackexchange.com/questions/107186/how-to-write-norm-which-adjusts-its-size
+\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
+\newcommand{\rv}{\mathbf{r}}
+The N body problem is defined as follows \cite{chenciner2000remarkable}:
+
+Let $\rv_i$ be the position of the $i^\text{th}$ body in the N body problem.
+Then, the force of gravity is the sum of gravity from every other body in the system.
+That is,
+
+\begin{equation}
+   F_g^{(i)} = m_i \frac{\mathrm{d}^2 \rv_i}{\mathrm{dt}^2} =
+   \sum_{j \neq i} G \frac{m_i m_j \hat{\rv}_{ij}}{\norm{\rv_{ij}}^3}
+   \label{eq:nbody}
+\end{equation}
+
+where $\rv_{ij} = \rv_j - \rv_i$, $m_i$ is the mass of body $i$, and $G$ is the gravitational constant.
+
+For our purposes, the reduced system is more useful since it is much easier to calculate than numerically solving the original 3-body problem.
+With three bodies in three dimension, their motion is described by 9 second order differential equations.
+Therefore, reducing the system to first order differential equations yields 18 first order equations.
+The reduced problem only requires 4 first order differential equations since the position is defined by two second order differential equations.
+
+Derived in \cite{eberle2007case}, the system of equations defined with dimensionless rotating (synodic) coordinates:
+
+\begin{align}
+    x'' - 2 y' &= x
+        - \frac{\alpha}{r_1^3} (x - \mu)
+        - \frac{\mu}{r_2^3} (x + \alpha) \nonumber \\
+    y'' + 2 x' &= \left(
+        1 - \frac{\alpha}{r_1^3}
+        - \frac{\mu}{r_2^3}
+    \right) y \nonumber \\
+\end{align}
+
+where
+
+\begin{align}
+    \mu &= \frac{m_1}{m_1 + m_2} \nonumber \\
+    \alpha &= 1 - \mu \nonumber \\
+    r_1(t) &= \left[(x(t) - \mu)^2 + y(t)^2\right]^{\frac{1}{2}} \nonumber \\
+    r_2(t) &= \left[(x(t) + \alpha)^2 + y(t)^2\right]^{\frac{1}{2}} \label{eq:reduc_3body_cor}
+\end{align}
+
+and where mass body 1 and two are located at $-\alpha$ and $\mu$.
+
+If we define the following system:
+
+\begin{align}
+    x_1 &= x \nonumber \\
+    x_2 &= x' \nonumber \\
+    y_1 &= y \nonumber \\
+    y_2 &= y', \label{eq:first_order_def}
+\end{align}
+
+then, we can take the derivative of each:
+
+\begin{align*}
+    x_1' &= x' \\
+    &= x_2,
+\end{align*}
+and
+\begin{align*}
+    x_2' &= x'' \\
+    &= 2 y' + x
+        - \frac{\alpha}{r_1^3} (x - \mu)
+        - \frac{\mu}{r_2^3} (x + \alpha) \\
+    &= 2 y_2 + x_1 - \frac{\alpha}{r_1^3} (x_1 - \mu)
+        - \frac{\mu}{r_2^3} (x_1 + \alpha),
+\end{align*}
+and
+\begin{align*}
+    y_1' &= y' \\
+    &= y_2,
+\end{align*}
+and
+\begin{align*}
+    y_2' &= y'' \\
+    &= - 2 x' + \left(
+        1 - \frac{\alpha}{r_1^3}
+        - \frac{\mu}{r_2^3}
+    \right) y \\
+    &= - 2 x_2 + \left(
+        1 - \frac{\alpha}{r_1^3}
+        - \frac{\mu}{r_2^3}
+    \right) y_1.
+\end{align*}
+
+Then, we can represent the second order system of motion as four first order equations:
+
+\begin{align}
+    x_1' &= x_2 \nonumber \\
+    x_2' &= 2 y_2 + x_1 - \frac{\alpha}{r_1^3} (x_1 - \mu)
+        - \frac{\mu}{r_2^3} (x_1 + \alpha) \nonumber \\
+    y_1' &= y_2 \nonumber \\
+    y_2' &= - 2 x_2 + \left(
+        1 - \frac{\alpha}{r_1^3}
+        - \frac{\mu}{r_2^3}
+    \right) y_1, \label{eq:reduce_3_body_prog_sys} \\
+\end{align}
+
+where
+
+\begin{align}
+    r_1 &= \left[(x_1 - \mu)^2 + y_1^2\right]^{\frac{1}{2}} \nonumber \\
+    r_2 &= \left[(x_1 + \alpha)^2 + y_1^2\right]^{\frac{1}{2}}.
+\end{align}
+
+Here, $x_1$ and $y_1$ represent the position of the third body relative to the other two,
+but not in the standard 
+Cartesian coordinates of 3D space. Outlined in the paper \cite{eberle2007case}, the 
+transformation back to Cartesian coordinates is given, but we don't need the exact 
+position of the bodies since we are only concerned with the chaotic motion, so that isn't 
+shown here.
+
+There are however two issues when integrating this problem. As objects approach each
+other, the distance between them 
+approaches zero. If we have a collision, then this approaches infinity. There is no 
+collision detection in this model as we assume falsely that each body has a radius of zero,
+so this case is not handled. Instead, we will get a 
+divide by zero error, and the model cannot be calculated further. Further, near collisions 
+are also an issue. Detailed in \cite{chambers1999hybrid}, with the distance between bodies 
+approaching zero, their potential energy becomes significantly smaller than other values, 
+such as those describing kinetic energy. This 
+produces a stiff system of equations, and introduces error. Since energy is conserved in
+this system (as it is a Hamiltonian system), the error introduced represents energy being 
+created. This causes the singularity property seen in Figure \ref{fig:singularity}
+
+\begin{figure}[h]
+    \centering
+    \begin{subfigure}[b]{0.4\textwidth}
+        \includegraphics[width=\textwidth]{images/r3b_collision.png}
+        \caption{The restricted 3 body problem with collision}
+        \label{fig:r3b_collision}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.4\textwidth}
+        \includegraphics[width=\textwidth]{images/r3b_collision_whole.png}
+        \caption{The resultant divergence of the body}
+        \label{fig:r3b_collision_whole}
+    \end{subfigure}
+    \caption{The Double Pendulum}
+    \label{fig:doub_pend}
+\end{figure}
+
+While it is possible to solve this issue with symplectic integrators \cite{chambers1999hybrid},
+
 \subsection{Double Pendulum}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=.5\linewidth]{images/example_doub_pend.png}
+    \caption{The Double Pendulum}
+    \label{fig:doub_pend}
+\end{figure}
 
@@
 
+$\theta_1$ and $\omega_1$ represent the angle and angular velocity of the inner pendulum 
+where $\theta_1=0$ implies the pendulum is straight down, and $\theta_1=\epsilon$ implies 
+the pendulum is rotated $\epsilon$ radians counter-clockwise. The same is true for the 
+outer pendulum relative to its pivot point, the end of the first pendulum. $l_1$ and $l_2$ 
+are the lengths of the inner and outer pendulums, respectively. $m_1$ and $m_2$ are the 
+masses of the balls on the end points of the inner and outer pendulums, respectively.
+Then, the positions of the endpoints $(x_1$, $y_1)$ and $(x_2, y_2)$ of the pendulum, with 
+the center pivot at the origin, $(0, 0)$, can be calculated as follows:
+
+\begin{align}
+    x_1 &= l_1 \sin(\theta_1), \nonumber \\
+    y_1 &= - l_1 \cos(\theta_1), \nonumber \\
+    x_2 &= x_1 + l_2 \sin(\theta_2), \nonumber \\
+    y_2 &= y_1 - l_2 \cos(\theta_2), \label{eq:doub_pend_endpoints}
+\end{align}
+
+% TODO: position of the pendulum end points
+
 While the system is a Hamiltonian system, Stachowiak decided to define the first order 
@@
 
-Repeating the same experiment, but with different parameters, the results are similar (see 
-Figure \ref{fig:doub_pend_energy}).
+Repeating the same experiment, but with different parameters ($l_1 = l_2 = 1$, and
+$m_1 = m_2 = 1$), the results are similar (see Figure \ref{fig:doub_pend_energy}).
+As inner theta increases, the largest LCE of the system seems to
+approach 1.4. % DATA POINT
 
@@
     \centering
-    %\includegraphics[.5\linewidth]{images/chaos_vs_energy_in_doub_pend.png}
+    \includegraphics[width=.5\linewidth]{images/chaos_vs_energy_in_doub_pend.png}
     \caption{Largest Lyapunov Characteristic Exponent vs Inner Theta Initial Condition}

### 2019-04-02 doc/paper/paper.tex
3 body problem paper written

@@
 \usepackage{graphicx}
+\usepackage{subcaption}
 
@@
 Further, given a periodic
-solution, we hope that a small perturbation, from error in measurement or from 
+solution, the hope is that a small perturbation, from error in measurement or from 
 floating-point errors in numerical calculations, doesn't affect the solution or 
@@
 the bodies of mass (known as the n-body problem). However, with the existence of
-other bodies of mass from other solar systems within the universe, we will have 
-small perturbations introduced into the system. If the system is chaotic, then 
+other bodies of mass from other solar systems within the universe, 
+small perturbations are introduced into the system. If the system is chaotic, then 
 over time, the actual trajectories will diverge from our model's 
@@
 In terms of forecasting, it's useful to know how chaotic a dynamical system is. 
-To measure this, we will measure the rate at which nearby trajectories 
+Chaos will be measured by the exponential rate at which nearby trajectories 
 diverge. Constants describing the divergence are known as the Lyapunov 
@@
 problem solution, and the double pendulum system.
-The Poincaré system is the one of the first (possibly the first), and Lorenz
-is a classical example. The double pendulum is commonly referred to as chaotic, 
-and I will show this later.
+The Poincaré system is the one of the first, and the Lorenz system
+is a classic example. The double pendulum is commonly referred to as chaotic, 
+and I will show this later. Both Poincaré's restricted 3 body problem solution
+and 
+the double pendulum are Hamiltonian systems representing physical systems
+within our universe.
 
@@
 other, the distance between them 
-approaches zero. If we have a collision, then this approaches infinity. There is no 
-collision detection in this model as we assume falsely that each body has a radius of zero,
-so this case is not handled. Instead, we will get a 
-divide by zero error, and the model cannot be calculated further. Further, near collisions 
-are also an issue. Detailed in \cite{chambers1999hybrid}, with the distance between bodies 
+approaches zero. If collision occurs, then this approaches infinity. There is no 
+collision detection in this model, and each body has a radius of zero,
+so this case is not handled. Instead, a divide by zero error occurs,
+and the model cannot be calculated further.
+
+Further, near collisions 
+are also an issue. With the distance between bodies 
 approaching zero, their potential energy becomes significantly smaller than other values, 
-such as those describing kinetic energy. This 
+such as those describing kinetic energy \cite{chambers1999hybrid}. This 
 produces a stiff system of equations, and introduces error. Since energy is conserved in
-this system (as it is a Hamiltonian system), the error introduced represents energy being 
-created. This causes the singularity property seen in Figure \ref{fig:singularity}
+this Hamiltonian system, the error introduced represents energy being 
+created. This can cause the singularity property seen in Figure \ref{fig:singularity}.
+This problem can be solved by using a symplectic integrator \cite{chambers1999hybrid}.
 
-\begin{figure}[h]
+\begin{figure}[H]
     \centering
-    \begin{subfigure}[b]{0.4\textwidth}
+    \begin{subfigure}[b]{0.3\textwidth}
         \includegraphics[width=\textwidth]{images/r3b_collision.png}
@@
     ~
-    \begin{subfigure}[b]{0.4\textwidth}
+    \begin{subfigure}[b]{0.3\textwidth}
         \includegraphics[width=\textwidth]{images/r3b_collision_whole.png}
@@
     \end{subfigure}
-    \caption{The Double Pendulum}
-    \label{fig:doub_pend}
+    \caption{The Restricted 3 Body Problem Error from Collision}
+    \label{fig:singularity}
 \end{figure}
 
-While it is possible to solve this issue with symplectic integrators \cite{chambers1999hybrid},
+While it is possible to use a symplectic integrator, it is not necessary. Under the right 
+initial conditions, the third body will diverge regardless. The simple divergent spiral
+isn't chaotic, so instead orbital and collision less initial conditions will be used.
+Without divergence and without collisions, the symplectic integrator is unnecessary.
 
 \subsection{Double Pendulum}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=.5\linewidth]{images/example_doub_pend.png}
-    \caption{The Double Pendulum}
-    \label{fig:doub_pend}
-\end{figure}
 
@@
 \end{align}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.5\linewidth]{images/example_doub_pend.png}
+    \caption{The Double Pendulum}
+    \label{fig:doub_pend}
+\end{figure}
 

### 2019-04-02 doc/paper/paper.tex
formatting

@@
 
-In 1963, Edward N. Lorenz published an article researching a simplified system of 
-ordinary differential equations modeling a convective system \cite{lorenz1963deterministic}. His 
-research popularized the possibility of a system being highly sensitive to 
-initial conditions, i.e., chaos theory; while he wasn't the first to discover 
-this phenomenon, he is considered to be the "official discoverer of chaos 
-theory" \cite{oestreicher2007history}.
+In 1963, Edward N. Lorenz published an article researching a simplified
+system of ordinary differential equations modeling a convective system
+\cite{lorenz1963deterministic}. His research popularized the possibility of a
+system being highly sensitive to initial conditions, i.e., chaos theory;
+while he wasn't the first to discover this phenomenon, he is considered to be
+the "official discoverer of chaos theory" \cite{oestreicher2007history}.
 
-When forecasting, the goal is to have accurate predictions. However, 
-a chaotic system's sensitivity to initial conditions implies that a small 
-perturbation as a result of error will produce incorrect solutions.
-Further, given a periodic
-solution, the hope is that a small perturbation, from error in measurement or from 
-floating-point errors in numerical calculations, doesn't affect the solution or 
-at least produces a quasi-periodic solution close to the periodic one. 
-(A solution or trajectory $F$ 
-is quasi-periodic if and only if for all $t$ and for some $\tau$, $F(t+\tau)$ 
-is arbitrarily close to $F(t)$ \cite{lorenz1963deterministic}.)
+When forecasting, the goal is to have accurate predictions. However, a
+chaotic system's sensitivity to initial conditions implies that a small
+perturbation as a result of error will produce incorrect solutions. Further,
+given a periodic solution, the hope is that a small perturbation, from error
+in measurement or from floating-point errors in numerical calculations,
+doesn't affect the solution or at least produces a quasi-periodic solution
+close to the periodic one. (A solution or trajectory $F$ is quasi-periodic if
+and only if for all $t$ and for some $\tau$, $F(t+\tau)$ is arbitrarily close
+to $F(t)$ \cite{lorenz1963deterministic}.)
 
 An example of this is the trajectories of the sun and planets in our solar
-system. Newton's equations can be used to create a system of equations to model 
-the bodies of mass (known as the n-body problem). However, with the existence of
-other bodies of mass from other solar systems within the universe, 
-small perturbations are introduced into the system. If the system is chaotic, then 
-over time, the actual trajectories will diverge from our model's 
-\cite{oestreicher2007history}.
+system. Newton's equations can be used to create a system of equations to
+model the bodies of mass (known as the n-body problem). However, with the
+existence of other bodies of mass from other solar systems within the
+universe, small perturbations are introduced into the system. If the system
+is chaotic, then over time, the actual trajectories will diverge from our
+model's \cite{oestreicher2007history}.
 
 One of the defining characteristics of chaos is that the error from
-perturbations grows exponentially \cite{oestreicher2007history}. At the time of 
-Lorenz's paper, unstable 
-error was not well understood, and this led to Lorenz's surprising discovery 
-\cite{oestreicher2007history}. Lorenz reported that a single iteration in 
-calculating the solution to the dynamical system took 
-approximately one second \cite{lorenz1963deterministic}. In his paper, he shows 
-one of his calculations having about 3000 iterations, which must have taken 
-around 50 minutes. Given the amount of time it takes to perform the calculation,
-it's tempting to continue the calculation from the output of the previous 
-calculation. However, the story of the discover is that while the computer used 
-6 digit accuracy, it output only 3 digits, so when the simulation was continued 
-by Lorenz with the less accurate measurements, he found the results to be very 
-different \cite{oestreicher2007history}.
+perturbations grows exponentially \cite{oestreicher2007history}. At the time
+of Lorenz's paper, unstable error was not well understood, and this led to
+Lorenz's surprising discovery \cite{oestreicher2007history}. Lorenz reported
+that a single iteration in calculating the solution to the dynamical system
+took approximately one second \cite{lorenz1963deterministic}. In his paper,
+he shows one of his calculations having about 3000 iterations, which must
+have taken around 50 minutes. Given the amount of time it takes to perform
+the calculation, it's tempting to continue the calculation from the output of
+the previous calculation. However, the story of the discover is that while
+the computer used 6 digit accuracy, it output only 3 digits, so when the
+simulation was continued by Lorenz with the less accurate measurements, he
+found the results to be very different \cite{oestreicher2007history}.
 
-Chaos can appear in many different situations. Lorenz's paper demonstrates chaos
-in a dynamical system useful for forecasting convection in the atmosphere or 
-liquids \cite{lorenz1963deterministic}. "Chaos theory has a few applications for
-modeling endogenous biological rhythms such as heart rate, brain functioning, 
-and biological docks" \cite{oestreicher2007history}. It can also appear in 
-solutions to Hamiltonian problems such as the 3 body problem (as mentioned 
-before) and the double pendulum problem (as I will show later). Understanding
-the predictability of chaos is very useful to many different fields.
+Chaos can appear in many different situations. Lorenz's paper demonstrates
+chaos in a dynamical system useful for forecasting convection in the
+atmosphere or liquids \cite{lorenz1963deterministic}. "Chaos theory has a few
+applications for modeling endogenous biological rhythms such as heart rate,
+brain functioning, and biological docks" \cite{oestreicher2007history}. It
+can also appear in solutions to Hamiltonian problems such as the 3 body
+problem (as mentioned before) and the double pendulum problem (as I will show
+later). Understanding the predictability of chaos is very useful to many
+different fields.
 
@@
 
-In terms of forecasting, it's useful to know how chaotic a dynamical system is. 
-Chaos will be measured by the exponential rate at which nearby trajectories 
-diverge. Constants describing the divergence are known as the Lyapunov 
-Characteristic Exponents (LCEs) \cite{sandri1996numerical}. There is an LCE for 
-each dimension, and since they all represent exponential growth. The error 
-growing at the rate of the largest LCE will dominate over the others.
+In terms of forecasting, it's useful to know how chaotic a dynamical system
+is. Chaos will be measured by the exponential rate at which nearby
+trajectories diverge. Constants describing the divergence are known as the
+Lyapunov Characteristic Exponents (LCEs) \cite{sandri1996numerical}. There is
+an LCE for each dimension, and since they all represent exponential growth.
+The error growing at the rate of the largest LCE will dominate over the
+others.
 
@@
 
-Given a small perturbation $\epsilon$ to a dynamical system in the direction of 
-the largest LCE, we can approximate the growth of the perturbation by the 
-equation $\epsilon(\Delta t) = \epsilon_0 e^{\lambda \Delta t}$ where $\lambda$ 
-is the largest LCE
-\cite{bezruchko2010extracting}. For an LCE of zero, then the error doesn't grow
-exponentially, and thus, there is no chaos. Error created
-from smaller LCEs is negligible. The approximate time scale where 
-perturbations become large is going to be proportional to $1 / \lambda$. This is
-known as the Lyapunov time \cite{bezruchko2010extracting}. For our solar
-system, the Lyapunov time is 10,000,000 years \cite{oestreicher2007history}.
-Researchers have used it to evaluate the effectiveness of forecasting models on
-chaotic problems \cite{pathak2018model}. Predicting the solution of
-a chaotic model over an interval of time which is multiple times the size of the
-Lyapunov time shows the predictability of a model.
+Given a small perturbation $\epsilon$ to a dynamical system in the direction
+of the largest LCE, we can approximate the growth of the perturbation by the
+equation $\epsilon(\Delta t) = \epsilon_0 e^{\lambda \Delta t}$ where
+$\lambda$ is the largest LCE \cite{bezruchko2010extracting}. For an LCE of
+zero, then the error doesn't grow exponentially, and thus, there is no chaos.
+Error created from smaller LCEs is negligible. The approximate time scale
+where perturbations become large is going to be proportional to $1 /
+\lambda$. This is known as the Lyapunov time \cite{bezruchko2010extracting}.
+For our solar system, the Lyapunov time is 10,000,000 years
+\cite{oestreicher2007history}. Researchers have used it to evaluate the
+effectiveness of forecasting models on chaotic problems
+\cite{pathak2018model}. Predicting the solution of a chaotic model over an
+interval of time which is multiple times the size of the Lyapunov time shows
+the predictability of a model.
 
-In order to find the Lyapunov time, we only need to solve for the largest LCE. 
-To find this, we will use a method detailed in \cite{viswanath1998lyapunov}.
+In order to find the Lyapunov time, we only need to solve for the largest
+LCE. To find this, we will use a method detailed in
+\cite{viswanath1998lyapunov}.
 
@@
 Three dynamical systems will be used to demonstrate the predictability of
-chaotic systems: the Lorenz system, Poincaré's 3-body 
-problem solution, and the double pendulum system.
-The Poincaré system is the one of the first, and the Lorenz system
-is a classic example. The double pendulum is commonly referred to as chaotic, 
-and I will show this later. Both Poincaré's restricted 3 body problem solution
-and 
-the double pendulum are Hamiltonian systems representing physical systems
-within our universe.
+chaotic systems: the Lorenz system, Poincaré's 3-body problem solution, and
+the double pendulum system. The Poincaré system is the one of the first, and
+the Lorenz system is a classic example. The double pendulum is commonly
+referred to as chaotic, and I will show this later. Both Poincaré's
+restricted 3 body problem solution and the double pendulum are Hamiltonian
+systems representing physical systems within our universe.
 
@@
 Lorenz's article, he uses the parameter values $\sigma=10$, $\beta=8/3$, and
-$\rho=28$ \cite{lorenz1963deterministic}, and it has a maximum LCE of $0.90566$ 
-\cite{viswanath1998lyapunov}. There are other parameters which
-produce chaotic results outlined in Table \ref{table:lorenz_params}. Also shown 
-in Table \ref{table:lorenz_params} are the values calculated by my Python
-module \texttt{echonn} are listed. These values are not as accurate, but 
-demonstrate the sufficient accuracy of \texttt{echonn}.
+$\rho=28$ \cite{lorenz1963deterministic}, and it has a maximum LCE of
+$0.90566$ \cite{viswanath1998lyapunov}. There are other parameters which
+produce chaotic results outlined in Table \ref{table:lorenz_params}. Also
+shown in Table \ref{table:lorenz_params} are the values calculated by my
+Python module \texttt{echonn} are listed. These values are not as accurate,
+but demonstrate the sufficient accuracy of \texttt{echonn}.
 
@@
 
-The 3-body problem, or the generalized n-body problem, is a famous problem which many have 
-analyzed; most notably, Herni Poincaré wrote many papers developing the area significantly
-\cite{chenciner2000remarkable}. One of Poincaré's solutions, known as the circular 
-restricted 3-body problem, demonstrates chaotic properties \cite{oestreicher2007history}. 
-Poincaré performs several reductions of the original 3-body problem in order to come to a 
-much more useful form of the problem. The mass of one of the objects is assumed to be 
-negligible. This is the case for Sun-Earth-Moon orbits. The moon has little effect on the 
-orbit of the sun.
-Another aspect is the rotating reference frame.
-Since the mass of one of the bodies is negligible, the problem simplifies to first solving a two body problem.
-While much easier to solver, Poincaré went further to stop the motion of the two planets by using a rotating reference frame.
-Rotating with the planets, the problem can be modeled seemingly by two static planets and a third mass-less planet orbiting around those in a two dimensional plane.
+The 3-body problem, or the generalized n-body problem, is a famous problem
+which many have analyzed; most notably, Herni Poincaré wrote many papers
+developing the area significantly \cite{chenciner2000remarkable}. One of
+Poincaré's solutions, known as the circular restricted 3-body problem,
+demonstrates chaotic properties \cite{oestreicher2007history}. Poincaré
+performs several reductions of the original 3-body problem in order to come
+to a much more useful form of the problem. The mass of one of the objects is
+assumed to be negligible. This is the case for Sun-Earth-Moon orbits. The
+moon has little effect on the orbit of the sun. Another aspect is the
+rotating reference frame. Since the mass of one of the bodies is negligible,
+the problem simplifies to first solving a two body problem. While much easier
+to solver, Poincaré went further to stop the motion of the two planets by
+using a rotating reference frame. Rotating with the planets, the problem can
+be modeled seemingly by two static planets and a third mass-less planet
+orbiting around those in a two dimensional plane.
 
@@
 \newcommand{\rv}{\mathbf{r}}
+
 The N body problem is defined as follows \cite{chenciner2000remarkable}:
@@
 Let $\rv_i$ be the position of the $i^\text{th}$ body in the N body problem.
-Then, the force of gravity is the sum of gravity from every other body in the system.
-That is,
+Then, the force of gravity is the sum of gravity from every other body in the
+system. That is,
 
@@
 
-where $\rv_{ij} = \rv_j - \rv_i$, $m_i$ is the mass of body $i$, and $G$ is the gravitational constant.
+where $\rv_{ij} = \rv_j - \rv_i$, $m_i$ is the mass of body $i$, and $G$ is
+the gravitational constant.
 
-For our purposes, the reduced system is more useful since it is much easier to calculate than numerically solving the original 3-body problem.
-With three bodies in three dimension, their motion is described by 9 second order differential equations.
-Therefore, reducing the system to first order differential equations yields 18 first order equations.
-The reduced problem only requires 4 first order differential equations since the position is defined by two second order differential equations.
+For our purposes, the reduced system is more useful since it is much easier
+to calculate than numerically solving the original 3-body problem. With three
+bodies in three dimension, their motion is described by 9 second order
+differential equations. Therefore, reducing the system to first order
+differential equations yields 18 first order equations. The reduced problem
+only requires 4 first order differential equations since the position is
+defined by two second order differential equations.
 
-Derived in \cite{eberle2007case}, the system of equations defined with dimensionless rotating (synodic) coordinates:
+Derived in \cite{eberle2007case}, the system of equations defined with
+dimensionless rotating (synodic) coordinates:
 
@@
 
-Then, we can represent the second order system of motion as four first order equations:
+Then, we can represent the second order system of motion as four first order
+equations:
 
@@
 
-Here, $x_1$ and $y_1$ represent the position of the third body relative to the other two,
-but not in the standard 
-Cartesian coordinates of 3D space. Outlined in the paper \cite{eberle2007case}, the 
-transformation back to Cartesian coordinates is given, but we don't need the exact 
-position of the bodies since we are only concerned with the chaotic motion, so that isn't 
+Here, $x_1$ and $y_1$ represent the position of the third body relative to
+the other two, but not in the standard Cartesian coordinates of 3D space.
+Outlined in the paper \cite{eberle2007case}, the transformation back to
+Cartesian coordinates is given, but we don't need the exact position of the
+bodies since we are only concerned with the chaotic motion, so that isn't
 shown here.
 
-There are however two issues when integrating this problem. As objects approach each
-other, the distance between them 
-approaches zero. If collision occurs, then this approaches infinity. There is no 
-collision detection in this model, and each body has a radius of zero,
-so this case is not handled. Instead, a divide by zero error occurs,
-and the model cannot be calculated further.
+There are however two issues when integrating this problem. As objects
+approach each other, the distance between them approaches zero. If collision
+occurs, then this approaches infinity. There is no collision detection in
+this model, and each body has a radius of zero, so this case is not handled.
+Instead, a divide by zero error occurs, and the model cannot be calculated
+further.
 
-Further, near collisions 
-are also an issue. With the distance between bodies 
-approaching zero, their potential energy becomes significantly smaller than other values, 
-such as those describing kinetic energy \cite{chambers1999hybrid}. This 
-produces a stiff system of equations, and introduces error. Since energy is conserved in
-this Hamiltonian system, the error introduced represents energy being 
-created. This can cause the singularity property seen in Figure \ref{fig:singularity}.
-This problem can be solved by using a symplectic integrator \cite{chambers1999hybrid}.
+Further, near collisions are also an issue. With the distance between bodies
+approaching zero, their potential energy becomes significantly smaller than
+other values, such as those describing kinetic energy
+\cite{chambers1999hybrid}. This produces a stiff system of equations, and
+introduces error. Since energy is conserved in this Hamiltonian system, the
+error introduced represents energy being created. This can cause the
+singularity property seen in Figure \ref{fig:singularity}. This problem can
+be solved by using a symplectic integrator \cite{chambers1999hybrid}.
 
@@
 
-While it is possible to use a symplectic integrator, it is not necessary. Under the right 
-initial conditions, the third body will diverge regardless. The simple divergent spiral
-isn't chaotic, so instead orbital and collision less initial conditions will be used.
-Without divergence and without collisions, the symplectic integrator is unnecessary.
+While it is possible to use a symplectic integrator, it is not necessary.
+Under the right initial conditions, the third body will diverge regardless.
+The simple divergent spiral isn't chaotic, so instead orbital and collision
+less initial conditions will be used. Without divergence and without
+collisions, the symplectic integrator is unnecessary.
 
@@
 The double pendulum is commonly cited to be chaotic
-\cite{stachowiak2006numerical} \cite{levien1993double}. Using the system 
-of equations derived by Stachowiak \cite{stachowiak2006numerical}, the system is
+\cite{stachowiak2006numerical} \cite{levien1993double}. Using the system of
+equations derived by Stachowiak \cite{stachowiak2006numerical}, the system is
 defined as follows:
@@
 
-$\theta_1$ and $\omega_1$ represent the angle and angular velocity of the inner pendulum 
-where $\theta_1=0$ implies the pendulum is straight down, and $\theta_1=\epsilon$ implies 
-the pendulum is rotated $\epsilon$ radians counter-clockwise. The same is true for the 
-outer pendulum relative to its pivot point, the end of the first pendulum. $l_1$ and $l_2$ 
-are the lengths of the inner and outer pendulums, respectively. $m_1$ and $m_2$ are the 
-masses of the balls on the end points of the inner and outer pendulums, respectively.
-Then, the positions of the endpoints $(x_1$, $y_1)$ and $(x_2, y_2)$ of the pendulum, with 
-the center pivot at the origin, $(0, 0)$, can be calculated as follows:
+$\theta_1$ and $\omega_1$ represent the angle and angular velocity of the
+inner pendulum where $\theta_1=0$ implies the pendulum is straight down, and
+$\theta_1=\epsilon$ implies the pendulum is rotated $\epsilon$ radians
+counter-clockwise. The same is true for the outer pendulum relative to its
+pivot point, the end of the first pendulum. $l_1$ and $l_2$ are the lengths
+of the inner and outer pendulums, respectively. $m_1$ and $m_2$ are the
+masses of the balls on the end points of the inner and outer pendulums,
+respectively. Then, the positions of the endpoints $(x_1$, $y_1)$ and $(x_2,
+y_2)$ of the pendulum, with the center pivot at the origin, $(0, 0)$, can be
+calculated as follows:
 
@@
 
-While the system is a Hamiltonian system, Stachowiak decided to define the first order 
-derivative components as angular velocity as opposed to momentum. This won't affect
-anything regarding the purpose of this paper.
+While the system is a Hamiltonian system, Stachowiak decided to define the
+first order derivative components as angular velocity as opposed to momentum.
+This won't affect anything regarding the purpose of this paper.
 
-Interestingly, the 
-degree of chaos is dependent on the initial conditions \cite{levien1993double}. 
-In the experiment, the initial conditions had angular velocities of zero, the outer 
-pendulum was straight down, and the inner pendulum varied from 0 degrees to 180 degrees 
-\cite{levien1993double}; the results showed that the LCE increased with theta.
+Interestingly, the degree of chaos is dependent on the initial conditions
+\cite{levien1993double}. In the experiment, the initial conditions had
+angular velocities of zero, the outer pendulum was straight down, and the
+inner pendulum varied from 0 degrees to 180 degrees \cite{levien1993double};
+the results showed that the LCE increased with theta.
 
-Repeating the same experiment, but with different parameters ($l_1 = l_2 = 1$, and
-$m_1 = m_2 = 1$), the results are similar (see Figure \ref{fig:doub_pend_energy}).
-As inner theta increases, the largest LCE of the system seems to
-approach 1.4. % DATA POINT
+Repeating the same experiment, but with different parameters ($l_1 = l_2 =
+1$, and $m_1 = m_2 = 1$), the results are similar (see Figure
+\ref{fig:doub_pend_energy}). As inner theta increases, the largest LCE of the
+system seems to approach 1.4. % DATA POINT
 

### 2019-04-02 doc/paper/scripts/LorenzLCEs.py
add lorenz system pictures

@@ -10,12 +10,13 @@ if __name__ == "__main__":
     for sigma, rho, beta, lambda_ in data:
 
         lces = []
-        for _ in range(10):
+        print('calculating lces...')
+        for i in range(10):
             sys = LorenzSystem(sigma, rho, beta)
             slv = SystemSolver(sys)
 
             lce, _ = slv.get_lce(T=150)
-            print('lce:', lce)
+            print('\t{}:'.format(i), lce)
             lces.append(lce)
         res = {}
         res['beta'] = beta

### 2019-04-02 doc/paper/scripts/LorenzPicture.py

@@ -1,0 +1,15 @@
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+from echonn.sys import LorenzSystem, SystemSolver
+
+if __name__ == "__main__":
+    lorenz_3d_plot = os.path.join('..', 'images', 'lorenz_3d_plot.png')
+    lorenz_nd_plot = os.path.join('..', 'images', 'lorenz_nd_plot.png')
+    slv = SystemSolver(LorenzSystem())
+    res = slv.run([0, 50], [10, 20, 30])
+    slv.plotnd(res, dims=['x', 'y', 'z'])
+    plt.savefig(lorenz_nd_plot)
+    slv.plot3d(res)
+    plt.savefig(lorenz_3d_plot)
+    plt.show(True)

### 2019-04-02 doc/paper/scripts/RenderR3Bod.py

@@ -23,4 +23,5 @@ if __name__ == "__main__":
     print('lce:', lce)
     #plt.xlim((-15, 15))
     #plt.ylim((-15, 15))
+    # Saved manually since it can't be scripted
     plt.show(True)

### 2019-04-02 src/echonn/sys/system_solver.py

@@ -170,7 +170,7 @@ class SystemSolver:
         plt.tight_layout()
         return fig
 
-    def plotnd(self, run=None, fig=None, figsize=None, legend=False):
+    def plotnd(self, run=None, fig=None, figsize=None, legend=False, dims=None):
         """ warning legend=True is going to mess up the layout. The legend can be seen by the side of the mini graph. """
         run = self.get_last_run(run)
         sys_num = run['index']
@@ -195,7 +195,10 @@ class SystemSolver:
             for _ in range(i+1):
                 plots[i].plot(res.t, y)  # TODO: fix this hack
                 # This makes graphs match color in overlay
-            plots[i].set_ylabel('dim {}'.format(i))
+            if dims is None:
+                plots[i].set_ylabel('dim {}'.format(i))
+            else:
+                plots[i].set_ylabel(dims[i])
             label = 'sys {} dim {}'.format(sys_num, i)
             overlay.plot(res.t, y, label=label)
         if legend:

### 2019-04-02 src/test/sys/test_system_solver.py

@@ -11,7 +11,6 @@ class TestModel(unittest.TestCase):
     def setUp(self):
         # test data
         self.circle_solver = SystemSolver(CircleSystem())
-        self.lorenz_solver = SystemSolver(LorenzSystem())
 
     def testSomething(self):
         res = self.circle_solver.run([0, 7], [0, 1])['results']

### 2019-04-03 doc/paper/scripts/CircleLCE.py
lorenz values reporting, lyapunov circle report

@@ -1,0 +1,7 @@
+from echonn.sys import CircleSystem, SystemSolver
+
+if __name__ == "__main__":
+    sys = SystemSolver(CircleSystem())
+    lce, run = sys.get_lce()
+    print('lce:', lce)
+    print(run['results'].y[2:, -1].reshape(2, 2))

### 2019-04-03 doc/paper/scripts/LorenzLCEs.py

@@ -7,6 +7,7 @@ if __name__ == "__main__":
         [16, 40, 4, 1.37446],
         [10, 28, 8/3, 0.90566],
     ]
+    results = []
     for sigma, rho, beta, lambda_ in data:
 
         lces = []
@@ -29,3 +30,10 @@ if __name__ == "__main__":
         res['relative error'] = res['error'] / lambda_
         print(res)
         print()
+        results.append(res)
+
+    print('results:')
+    for result in results:
+        toprint = [res['beta'], res['rho'], res['sigma'], res['lambda'],
+                   res['mean'], res['error'], res['relative error']]
+        print('        ' + ' & '.join(toprint))

### 2019-04-03 doc/paper/paper.tex

@@
 
+\newcommand{\echonn}{\texttt{echonn}}
 \newcommand{\der}[2][t]{\frac{\mathrm{d}#2}{\mathrm{d}#1}}
@@
 This is a paper about chaos.
+% TODO: Intro
 % TODO: mention \texttt{echonn}
@@
 
+\subsection{Calculating the Largest Lyapunov Characteristic Exponent}
+
 In order to find the Lyapunov time, we only need to solve for the largest
-LCE. To find this, we will use a method detailed in
+LCE. To find this, we will use the following method detailed in
 \cite{viswanath1998lyapunov}.
 
-% TODO: Lyapunov calculation
-% TODO: Test with Circle yielding zero
+\newcommand{\xv}{\mathbf{x}}
+\newcommand{\fv}{\mathbf{f}}
+\newcommand{\Mv}{\mathbf{M}}
+
+The dynamical system will be defined by $\der{\xv}=\fv(\xv)$ where $\fv, \xv \in 
+\mathbb{R}^d$. Given an initial condition $\xv_0$, the largest LCE, denoted 
+$\lambda_1$, is calculated as follows:
+
+\begin{equation}
+    \lambda_1 = \lim_{t \to \infty} \frac{1}{t}\log \sigma_1(\frac{\partial \xv(t)}{\partial \xv_0}) \label{eq:max_lce}
+\end{equation}
+
+where $\sigma_1$ is the first singular value of the matrix $\frac{\partial 
+\xv(t)}{\partial \xv_0}$ and $\log$ is the natural logarithm. Since we can only 
+numerically approximate the limit for $\lambda_1$, we will need to use a
+sufficiently large $t$ until the limit converges to the accuracy we need.
+
+In order to calculate $\frac{\partial \xv(t)}{\partial \xv_0}$, we need to integrate the matrix differential equation:
+
+\begin{equation}
+    \der{\Mv(t)} = \frac{\partial \fv}{\partial \xv} \Mv(t)
+    \label{eq:lce_mat_diffeq}
+\end{equation}
+
+where $\Mv(t) \in \mathbb{R}^{d \times d}$ and $\frac{\partial \fv}{\partial \xv}$
+is the $d \times d$ Jacobian matrix of $\fv(\xv)$ at $\xv(t)$.
+
+Since values for $\xv(t)$ are needed as well, the systems will be integrated 
+together. If the dynamical system has dimensions $d$, then the new system of
+differential equations will have dimension $d + d \times d$. For larger dynamical 
+systems, this calculation becomes costly. The dynamical systems analyzed are
+small enough (having dimensions ranging from 3-4) that the calculation can be 
+performed timely and accurately.
+
+There are several issues with calculating $\lambda_1$ with this algorithm.
+Already mentioned, we need to choose a sufficiently
+large $t$ for the limit to converge. 
+Unfortunately, for chaotic systems, $\Mv$ will increase rapidly with time. The 
+large floating point values will eventually overflow and $\lambda_1$ will not
+be accurate. I have found empirically that if values of $\Mv$ don't exceed 
+$10^{100}$, then $\lambda_1$ will be accurate.
+
+Implemented in \echonn, the algorithm was tested on the Lorenz system for various 
+parameters and initial conditions. The LCEs match the expected values very 
+accurately (see the Lorenz System section: \ref{sec:lorenz}).
+The 3 dimensional 3-body problem is an 18 dimensional dynamical system, so while 
+the dynamical system is implemented in \echonn, its max LCE can't be calculated 
+accurately since the resultant system has 342 dimensions. The algorithm was 
+also tested on the differential equation which produces a circle:
+
+\begin{align*}
+    x' &= y \\
+    y' &= -x
+\end{align*}
+
+with the solution
+
+\begin{align*}
+    x(t) &= r\sin(t) \\
+    y(t) &= r\cos(t)
+\end{align*}
+
+for any $r \in \mathbb{R}$.
+
+Given an initial condition $(x_0, y_0)$, it produces the trajectory of a circle
+with radius $r=\sqrt{x_0^2 + y_0^2}$. If perturbed, the radius will change, 
+$r_\epsilon=\sqrt{(x_0+\epsilon_x)^2 + (y_0+\epsilon_y)^2}$, so the new 
+trajectory is
+
+\begin{align*}
+    x_\epsilon(t) &= r_\epsilon\sin(t) \\
+    y_\epsilon(t) &= r_\epsilon\cos(t).
+\end{align*}
+
+Now, the $L^2$ norm of error is 
+\begin{align*}
+    \sqrt{(x_\epsilon - x)^2 + (y_\epsilon - y)^2} &= \sqrt{(r_\epsilon - r)^2\sin^2(t) + (r_\epsilon - r)^2\cos^2(t)} \\
+    &= \sqrt{(r_\epsilon - r)^2(\sin^2(t) + \cos^2(t))^2} \\
+    &= \sqrt{(r_\epsilon - r)^2} \\
+    &= \vert r_\epsilon - r \vert.
+\end{align*}
+
+The error doesn't grow at all over time, so the LCE should be less than or equal to zero as the system is not chaotic. The algorithm reports an LCE of $-1.15954 \times 10^{-06}$ as expected.
+
+Empirically shown, \echonn can be used to identify the max LCE of smaller 
+dynamical systems, and special precaution has to be taken when choosing a 
+time interval. Using a larger time will yield a more accurate $\lambda_1$, but 
+it may also cause an overflow if the system is chaotic. Thus, 
+maximum time values need to be chosen such that $\vert \vert \Mv\vert 
+\vert_{\infty}$ doesn't exceed $10^{100}$ (where $\vert \vert \cdot \vert 
+\vert_{\infty}$ is the $L^\infty$ norm).
 
@@
 \subsection{Lorenz System}
-
-% TODO: Image of Lorenz System
+\label{sec:lorenz}
 
@@
 shown in Table \ref{table:lorenz_params} are the values calculated by my
-Python module \texttt{echonn} are listed. These values are not as accurate,
-but demonstrate the sufficient accuracy of \texttt{echonn}.
+Python module \echonn are listed. These values are not as accurate,
+but demonstrate the sufficient accuracy of \echonn.
 
@@
         \hline
-        $\sigma$ & $\rho$ & $\beta$ & actual $\lambda$ & calculated $\lambda$ & relative error of calculated $\lambda$ \\
+        $\sigma$ & $\rho$ & $\beta$ & actual $\lambda_1$ & calculated $\lambda_1$ & relative error of calculated $\lambda_1$ \\
         \hline \hline
@@
 \end{table}
+
+The Lorenz system is named after its creator, but it is also known as the
+Lorenz butterfly due to the shape of the attractor of the system. The Lorenz
+Butterfly can be seen in Figure \ref{fig:lorenz_butter}.
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.6\textwidth}
+        \includegraphics[width=\textwidth]{images/lorenz_nd_plot.png}
+        \caption{Plot of each dimension of Lorenz System}
+        \label{fig:lorenz_nd}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.6\textwidth}
+        \includegraphics[width=\textwidth]{images/lorenz_3d_plot.png}
+        \caption{The Lorenz Butterfly}
+        \label{fig:lorenz_3d}
+    \end{subfigure}
+    \caption{The Lorenz System Visuals}
+    \label{fig:lorenz_butter}
+\end{figure}
 
@@
 
-Derived in \cite{eberle2007case}, the system of equations defined with
-dimensionless rotating (synodic) coordinates:
+Derived in two independent papers \cite{eberle2007case} and \cite{frnkacase},
+the system of equations defined with dimensionless rotating (synodic)
+coordinates:
 
@@
 and where mass body 1 and two are located at $-\alpha$ and $\mu$.
+
+Although the notation in both papers was different, the equations were the same. 
+It's also worth mentioning that Eberle's paper was for his Masters Thesis, and 
+Frnka's paper has no details associated with it, but I was able to find a 
+mathematician with the exact name who achieved his doctorate. With that said, 
+I believe the math is reliable, and the simulations seem to work and at least 
+demonstrate chaotic properties (the only requirement for this project). With my
+lacking physics knowledge, I was unable to derive these equations myself from
+the assumptions made by Henri Poincaré outlined in 
+\cite{chenciner2000remarkable}.
 
@@
 the other two, but not in the standard Cartesian coordinates of 3D space.
-Outlined in the paper \cite{eberle2007case}, the transformation back to
+Outlined in the both papers, \cite{eberle2007case} and \cite{frnkacase}, the transformation back to
 Cartesian coordinates is given, but we don't need the exact position of the
@@
     \centering
-    \begin{subfigure}[b]{0.3\textwidth}
+    \begin{subfigure}[b]{0.4\textwidth}
         \includegraphics[width=\textwidth]{images/r3b_collision.png}
-        \caption{The restricted 3 body problem with collision}
+        \caption{Restricted 3 Body Problem with collision}
         \label{fig:r3b_collision}
@@
     ~
-    \begin{subfigure}[b]{0.3\textwidth}
+    \begin{subfigure}[b]{0.4\textwidth}
         \includegraphics[width=\textwidth]{images/r3b_collision_whole.png}
@@
 \end{figure}
-
-% TODO: position of the pendulum end points
 

### 2019-04-12 doc/paper/paper.tex
numerical methods section written

@@
 
+Neural Networks are known to be able to approximate any function given 
+enough data and time to train \cite{cybenko1989approximation}.
+In light of this, the hope is that a deep learning based model,
+the ESN, will outperform ARIMA.
+
 This is a paper about chaos.
@@
 
-\subsection{History}
+%\subsection{History}
 
@@
 others.
-
-% TODO: talk about attractors
 
@@
 
-Now, the $L^2$ norm of error is 
+Now, the $\ell^2$ norm of error is 
 \begin{align*}
@@
 \vert_{\infty}$ doesn't exceed $10^{100}$ (where $\vert \vert \cdot \vert 
-\vert_{\infty}$ is the $L^\infty$ norm).
+\vert_{\infty}$ is the $\ell^\infty$ norm).
 
@@
 
-% TODO: numerical solutions section
+\subsection{Numerical Integration Methods}
+
+In order to generate the data to be analyized, the solutions to these 
+dynamical systems need to be computed. Each of the systems are defined by 
+a system of differential equations which can be numerically solved. 
+However, due to the chaotic nature, error introduced by the numerical 
+solution will grow exponentially, so precaution need to be taken to avoid 
+having this sort of error introduced into the collected data. When 
+numerically solving chaotic problems, it's possible for the chaotic 
+behavior to be suppressed, or where there shouldn't be chaotic behavior,
+it could be created \cite{corless1994good}.
+
+However, it's been shown that the Fourth Order Runge-Kutta Method (RK4)
+outperforms other numerical methods such as Euler's Method and the
+Midpoint Method for chaotic systems \cite{zidan2011effect}.
+Not only does RK4 perform better than other methods, it serves as a good
+method for producing accurate numerical solutions to choatic systems with 
+small enough time steps: the Lü chaotic system \cite{mehdi2017using} and
+the Zhou system \cite{roslan2013solving}, for example. One article points
+out that a larger concern 
+regarding these solutions is the accuracy of the initial conditions 
+measured from the physical problem which the mathematical model is based,
+but otherwise, they echo the strength of correctly implemented 
+numerical solutions \cite{corless1994good}. Further, my initial conditions
+are randomly generated, so this is not a concern.
+
+I will be using an improved version of RK4 which
+incorporates the fifth order Runge-Kutta: RK4(5)
+\cite{dormand1980family}. This is the method implemented in
+the Python function \texttt{scipy.integrate.solve\_ivp}, and my
+numerical solver
+wraps this function to integrate the chaotic systems. Using small time
+steps will yield accurate samples from the chaotic systems.
 
@@
 The Lorenz system is named after its creator, but it is also known as the
-Lorenz butterfly due to the shape of the attractor of the system. The Lorenz
+Lorenz butterfly due to the shape of the system. The Lorenz
 Butterfly can be seen in Figure \ref{fig:lorenz_butter}.
@@
 
-and where mass body 1 and two are located at $-\alpha$ and $\mu$.
+and where mass body one and two are located at $-\alpha$ and $\mu$.
 
@@
 
-%%% MODELS %%%
-\section{Models}
+\section{The Echo State Network}
 
-\subsection{ARIMA}
-%\subsection{Recurrent Neural Networks}
-\subsection{Echo State Networks}
+\newcommand{\Tv}{\mathbf{T}}
+\newcommand{\Uv}{\mathbf{U}}
+\newcommand{\Dv}{\mathbf{D}}
+\newcommand{\tv}{\mathbf{t}}
+\newcommand{\uv}{\mathbf{u}}
+\newcommand{\dv}{\mathbf{d}}
+\newcommand{\yv}{\mathbf{y}}
+\newcommand{\Fv}{\mathbf{F}}
+\newcommand{\wv}{\mathbf{w}}
+
+The Echo State Network (ESN) is a type of Recurrent Neural Network (RNN)
+used for time series analysis, but it is trained differently than most
+RNNs and is 
+characterized by having the echo state property. I implement the design
+detailed by Jaeger in is 2002 book \cite{jaeger2002tutorial} and
+further detailed in his 2007 article \cite{jaeger2007echo}.
+
+A time series data set will contain samples of data which change
+over time. Part of the data at each time point is measured or known
+before, called input data and denoted by $\uv(t)$ with $K$ dimensions.
+The rest of the data will be the goal for prediction, called output data and denoted $\dv(t)$ with $L$ dimensions.
+
+For $T$ time samples, $t_1, t_2, \dots, t_T$, we have a $T \times K$ 
+matrix, $\Uv$, with sample input vectors as rows. We also
+have a $T \times L$ matrix,
+$\Dv$, with sample output vectors as rows. A model will predict the 
+values of $\dv$, and this prediction will be denoted $\yv$.
+
+\subsection{Recurrent Neural Networks and Traditional Training Method}
+\label{sec:rnn}
+RNNs are usually designed by having $K$ input units, $L$ output units, 
+and some internal units organized by some network structure. These units
+have evolving values for every time point. The
+structure of networks describes how units are connected. 
+The input units at time $t_i$ are set to $\uv(t_i)$. However, output
+units and
+internal units are calculated from the values of other units. These other 
+units are said to be connected to the resultant unit. Output
+units are calculated from current input units, current internal units,
+and from previous output units. Internal units 
+are calculated from the values of the current input units, 
+previous internal units, and previous output units.
+
+Although, it's rare that the output units are calculated from previous 
+output units \cite{jaeger2002tutorial}. Sometimes for training, the output 
+units are set to the value of $dv$, called teacher forcing, so that the 
+model trains on actual values as opposed to its accumulated errors 
+\cite{jaeger2002tutorial}.
+
+To calculate the value of a unit, an activation function is applied to a 
+linear combination of the units connected to it \cite{svozil1997introduction}. The constants for the 
+linear combination are called weights, and they are parameters of the 
+model. Which activation function is used is also a parameter of the 
+function. Typically, some non-linear function is used, such as $\tanh$, 
+sigmoid, or rectified linear unit.
+
+During training, we search for
+weights which maximize the model's performance on data
+which the model hasn't seen. A common training method for RNNs
+is to use a form of backpropagation \cite{jaeger2002tutorial}. 
+Backpropagation for RNNs is slightly different from traditional Neural Networks, but the idea is the same.
+
+For example, given some neural network model $\yv(\xv; \wv)$, where
+$\xv$ is known input data, $\wv$ is the model's parameters, and $\yv$ 
+approximates some values, $\tv$, then we define an objective 
+function as the sum of squared error:
+
+\newcommand{\Ev}{\mathbf{E}}
+
+\begin{equation}
+    \Ev(\xv; \wv) = \vert \vert \yv(\xv; \wv) - \tv \vert \vert_2^2.
+    \label{eq:error_eq_rnn}
+\end{equation}
+
+Using steepest descent to solve this minimization problem (as shown in 
+\cite{svozil1997introduction}), we can
+reduce $\Ev$ in (\ref{eq:error_eq_rnn}) by
+updating the weights as follows:
+
+\begin{equation}
+    \wv = \wv - \lambda \nabla_\wv \Ev
+\end{equation}
+
+for some learning rate $\lambda$.
+
+\subsection{Echo State Network Implementation}
+
+The implementation design used for this paper is detailed by Herbert Jaeger
+in \cite{jaeger2002tutorial} and \cite{jaeger2007echo}. I explain it here 
+to clarify my exact process (and with a slight change of notation to be 
+consistent with the rest of this paper).
+
+Like RNNs, ESNs have an input layer with $K$ units and with values 
+$\uv(t_1), \uv(t_2), \dots,  \uv(t_T)$ where $\uv(t) \in \mathbb{R}^K$.
+$K$ is zero if there are no input units. There is an internal layer
+with $N$ units, denoted by $\xv(t_i) \in \mathbb{R}^N$ for
+$i = 1, 2, \dots, T$, and 
+there is an output 
+layer with $L$ units, denoted by $\yv(t_i) \in \mathbb{R}^L$ for all
+$i = 1, 2, \dots, T$.
+
+\newcommand{\W}{\mathbf{W}}
+\newcommand{\Win}{\mathbf{W}_{in}}
+\newcommand{\Wback}{\mathbf{W}_{back}}
+\newcommand{\Wout}{\mathbf{W}_{out}}
+
+How the units are connected can vary, but the suggested way 
+\cite{jaeger2002tutorial} is to connect the internal layer to the current 
+input layer, the previous internal layer, and the previous output layer. 
+These connections will have weights described in \ref{sec:rnn}. 
+Respectively, these weights will be denoted by the matrices:
+$\Win$, $\W$, and $\Wback$. The activation function for the internal
+layer is $\tanh$.
+
+The output layer is connected to the current input layer and current 
+internal layer. The weight for these connections will be denoted $\Wout$ 
+(we only have one matrix since the connected layers are concatenated into a
+single vector, per Jaeger's notation \cite{jaeger2002tutorial}). The 
+activation function for these connections is linear. Although, the 
+recommended choice
+is $\tanh$, the output layer doesn't necessarily lie in the range of 
+$\tanh$, so it can't be used without some form of normalization.
+
+In mathematical notation, the model is defined as follows:
+
+\begin{align}
+    \xv(t_{i}) &= \tanh \left(
+            \Win \uv(t_i)
+            + \W \xv(t_{i-1})
+            + \Wback \yv(t_{i-1})
+        \right) \\
+    \yv(t_i) &= \Wout (\uv(t_i), \xv(t_i))_{\text{concat}}.
+\end{align}
+
+$\xv$ and $\yv$ are calculated from previous unit values but for the 
+first time step, there are none, so we need to initialize the
+units to zeros \cite{jaeger2002tutorial}. 
+This is done by defining $\xv(t_0)$ and $\yv(t_0)$ to zeros.
+
+The weights are initialized such that the model has the defining 
+characteristic of echo state networks: the echo state property
+(Section \ref{sec:eprop}).
+During the training phase (Section \ref{sec:etrain}), $\Wout$ is fit to
+the training data.
+
+Once the network has been initialized and trained, it can be used to 
+perform time series analysis.
+
+\subsection{The Echo State Property}
+\label{sec:eprop}
+
+The echo state property is defined by the following \cite{jaeger2002tutorial}:
+
+\begin{quote}
+    Assume an untrained network with weights $\Win$, $\W$, and $\Wback$ is 
+    driven by teacher input $\uv(n)$ and teacher-forced by teacher output 
+    $\dv(n)$ from compact intervals $\Uv$ and $\Dv$. The network ($\Win$, 
+    $\W$, and $\Wback$) has echo states w.r.t. $\Uv$ and $\Dv$, if for 
+    every left-infinite input/output sequence $(\uv(n), \dv(n-1))$, where 
+    $n=\dots, -2, -1, 0$, and for all state sequences $\xv(n)$,
+    $\xv'(n)$ compatible with the teacher sequence, i.e. with
+    \begin{align}
+        \xv(n) &= \tanh \left(
+                \Win \uv(n)
+                + \W \xv(n-1)
+                + \Wback \yv(n-1)
+            \right) \\
+        \xv'(n) &= \tanh \left(
+                \Win \uv(n)
+                + \W \xv'(n-1)
+                + \Wback \yv(n-1)
+            \right)
+    \end{align}
+    it holds that $\xv(n) = \xv'(n)$ for all $n \leq 0$.
+\end{quote}
+
+The idea is that if the network is run long enough, it will, in a sense, 
+forget long past input and focus only on the most recent input. If two 
+networks with differing internal states are run with the same input and 
+teacher-forced output, they will eventually converge to each other.
+It's name suggests data fed to the model will echo around but eventually 
+fade as echos do.
+
+Jaeger outlines a process to create weights which will likely have
+echo state properties. The process is posed as conjecture and 
+doesn't guarantee the echo state property. However, the echo state is very 
+likely and can be
+assumed confidently \cite{jaeger2002tutorial} \cite{jaeger2007echo}.
+
+To demonstrate the echo state property, initialize the network with random 
+values. Set user input (if any) and teacher-forced output units to zero,
+After enough iterations, the internal state should converge to zero.
+
+The process to initialize the echo state weights, $\W$, is as follows \cite{jaeger2002tutorial}:
+
+\newcommand{\specrad}{\vert \lambda_\text{max} \vert}
+\begin{enumerate}
+    \item Initialize an $N \times N$ matrix, $\W_0$, with values from a 
+    uniform distribution in $[-1, 1]$.
+    \item Calculate the spectral radius, $\specrad$, of $\W_0$, i.e., the 
+    maximum absolute value of the eigenvalues of $\W_0$.
+    \item Normalize $\W_0$ to $\W_1$ so that it has unit spectral radius:
+    $\W_1 = \frac{1}{\specrad} \W_0$
+    \item Choose parameter $\alpha < 1$. Typically, $\alpha$ lies in
+    the range $(0.7, 0.98)$ \cite{jaeger2002tutorial}. This number will 
+    represent the new spectral radius of $\W$.
+    \item Scale $\W_1$ to $\W_2$ to have a spectral radius of
+    $\alpha$: $\W_2 = \alpha \W_1$. Now, let $\W = \W_2$, and the model
+    will have the echo state property \cite{jaeger2002tutorial}.
+\end{enumerate}
+
+If the spectral radius of $\W$ is found to be greater than 1, then it can 
+be shown that the model doesn't demonstrate the echo state property; 
+Jaeger's process uses this fact by ensuring that this isn't the case 
+\cite{jaeger2002tutorial}. While it can't guarantee an echo state property,
+the resultant matrix does have the property frequently enough that it's a
+reliable process. In his words, the model "has always been found to be an 
+echo state network" \cite{jaeger2002tutorial}.
+
+\newcommand{\vv}{\mathbf{v}}
+
+Applying linear algebra, this makes a lot of sense. Let $A$ be an
+$N \times N$ matrix with spectral radius $r = \rho(A) > 0$ and denote its
+eigenvalues $\lambda_i$ and eigenvectors $\vv_i$ for
+$i = 1, 2, \dots, N$.  Then, define $B = \frac{\alpha}{r} A$.
+
+By definition, $\vv_i A = \lambda_i \vv_i$, so 
+\begin{align*}
+   \vv_i B &= \frac{\alpha}{r} \vv_i A \\
+   &= \frac{\alpha}{r} \lambda_i \vv_i
+\end{align*}
+
+Therefore, $B$ has eigenvalues $\frac{\alpha}{r} \lambda_i$ for all $i$. 
+Further, its spectral radius is then
+
+\begin{align*}
+    \vert\vert \frac{\alpha}{r} \lambda \vert\vert_\infty
+    &= \frac{\alpha}{r} \vert\vert \lambda \vert\vert_\infty \\
+    &= \frac{\alpha}{r} r \\
+    &= \alpha
+\end{align*}
+
+where $\lambda$ is a vector of eigvenvalues and
+$\vert \vert \cdot \vert \vert_\text{max}$ is the $\ell^\infty$ norm.
+
+As the model iterates, the matrix transformation of $B$ is repeatedly 
+applied. With $\uv$ and teacher-forced $\dv$ as zeros, the internal state
+has the value $\xv(t_{k+1}) = \tanh(B \xv_k)$ for iteration $k$.
+
+Since the spectral radius of $B$ is less than 1, 
+\begin{align*}
+    \lim_{k \to \infty}B^k \vv_i &= \lim_{k \to \infty} \lambda_i^k \vv_i & \text{ (where $\lambda_i$ and $\vv_i$ is an eigvenvalue-eigenvector pair of $B$)} \\
+    &= 0 \vv_i & \text{ (since $\vert\lambda_i\vert < \alpha < 1$ and $\lim_{k \to \infty}\alpha^k = 0$)} \\
+    &= 0
+\end{align*}
+
+This suggests that as the model iterates, the internal state converges to 
+zero.
+
+The rest of weights can be initialized however desired, so we will use a 
+common technique.
+
+\begin{enumerate}
+    \item Initialize an $N \times M$ matrix, $\W_0$, with values from a 
+    uniform distribution in $[-1, 1]$. $M$ and $N$ here are not the 
+    same values from before for the sake for generality. 
+    \item Scale the matrix by $\frac{1}{\sqrt{M}}$ to calculate the weight matrix.
+\end{enumerate}
+
+Then, $\Win$ and $\Wback$ can be calculated as follows:
+
+\begin{align}
+    \Win &= \frac{\W_0}{\sqrt{K}} \nonumber \\
+    \Wback &= \frac{\W_0}{\sqrt{L}} \nonumber
+\end{align}
+
+Different from most training techniques, $\Win$, $\W$, and $\Wback$ are
+never changed after initialization.
+
+\subsection{Echo State Network Training}
+\label{sec:etrain}
+
+In order to train the echo state network, we need to find good 
+weight values for $\Wout$. Training the echo state network is different
+from the backpropagation approach in that we can solve for
+and calculate directly the weights matrix; more modern methods use
+an iterative approach which converges to good weight values. Part of the
+reason is due to the calculation cost of the iterative methods. Today, that
+it is no longer an issue, but historically, it was a strong quality of
+the echo state network \cite{jaeger2007echo}.
+
+First, the data (input, $\uv$, and output, $\dv$, values) is partitioned into three groups:
+\begin{enumerate}
+    \item Initialization values at time points
+    $t_0, t_1, \dots, t_{T_0-1}$.
+    \item Training values from
+    $t_{T_0}, \dots, t_T$
+    \item Testing values from
+    $t_{T+1}, \dots, t_{T_f}$
+\end{enumerate}
+
+Second, the initialization data is fed into the model. This initializes the
+inner state of the model, and the initial state (zeros) is lost due to the 
+echo state property. $T_0$ should be chosen such that this is truly the 
+case. Like $\alpha$, $T_0$ is a parameter of the echo state network and
+differs depending on the problem. Good values for $T_0$ are 10 to 500 
+depending on the problem complexity.
+
+Third, the training values are fed into the model to sample the network.
+We can store the input unit values, the internal state values, and the 
+desired output values as matrices $M$ and $T$. $M$ will store the input 
+unit values and internal state value pairs
+$(\uv(t_{T_0}), \xv(t_{T_0})), \dots, (\uv(t_T), \xv(t_T))$
+in its rows. $T$ will store the output unit values
+$\dv(t_{T_0}), \dots, \dv(t_T)$.
+
+To train the model, we want $\Wout M$ to approximate accurately
+$g^{-1}(T)$, where $g$ is the activation function for the output layer.
+Since we use an idempotent output layer activation function,
+$g^{-1}(T) = T$.
+
+Therefore, $\Wout$ should be found such that $\Wout M = T$ 
+\cite{jaeger2002tutorial}.
+Solving for $\Wout$,
+
+\begin{align*}
+    \Wout M &= T \\
+    (\Wout M)^T &= T^T & \text{ (where $A^T$ is the transpose of $A$)} \\
+    M^T \Wout^T &= T^T \\
+    (M^T)^{-1} M^T \Wout^T &= (M^T)^{-1} T^T & \text{ (where $A^{-1}$ is the Moore-Penrose pseudoinverse \cite{ben2003generalized} of $A$)} \\
+    \Wout^T &= (M^T)^{-1} T^T \\
+    \Wout &= \left((M^T)^{-1} T^T\right)^T \\
+    \Wout &= T \left((M^T)^{-1}\right)^T
+\end{align*}
+
+Using packages like \texttt{numpy.linalg.eig} and 
+\texttt{numpy.linalg.pinv}, solving for eigenvalues and
+pseudoinverses for $\Wout$ is very 
+easy. After solving for $\Wout$, the echo state network is trained.
+
+To evaluate the network, the testing values are fed into the model. 
+However, this time, the output layer is not teacher-forced since 
+in real application, the output values are unknown. 
+To measure how well the echo state network approximates the desired output
+we define the error to be the root mean squared error (RMSE) of the
+actual output values $\dv$ and
+predicted values $\yv$: $\sqrt{\sum_{i=T+1}^{n} (\dv(t_i) - \yv(t_i))^2}$ 
+where $n \in [T+1 \dots T_f]$. The error represents the predictive power
+of the model for $n$ time steps in the future. If RMSE is low for larger 
+values of $n$, the model has a strong long term predictive power.
+
+%noise: 0.0001 to 0.01 \cite{jaeger2002tutorial}
+
 %\subsection{LTSM}

### 2019-04-12 src/echonn/ml/__init__.py
add analytics code

@@ -1,1 +1,2 @@
 from .echo_state_network import EchoStateNetwork
+from .esn_experiment import TSData, TSAnalysis, ESNExperiment

### 2019-04-12 src/echonn/ml/echo_state_network.py

@@ -8,7 +8,7 @@ class EchoStateNetwork(TimeSeriesForecaster):
     def __init__(self, K, N, L, T0=50, alpha=.8, noise=0.0, f=None, g=None, g_inv=None):
         """
         K - input units, u
-        N - internal units, x
+        N - internal units, x (should not exceed Tf/10 to Tf/2 at risk of overfitting
         L - output units, y
         T0 - how long does it take for transient dynamics to wash out (10 for
         small, fast nets to 500 for large, slow nets). T is the sampling range, set by the train call.
@@ -16,6 +16,7 @@ class EchoStateNetwork(TimeSeriesForecaster):
         Specifies whether the algorithm should feed noise into the model
         during sampling. Uniform distribution ranging from [-noise, noise].
         """
+        # TODO: implement bias
         self.shapes = [
             (N, K),
             (N, N),
@@ -96,7 +97,9 @@ class EchoStateNetwork(TimeSeriesForecaster):
 
     def score(self, ds, ys, T, Tf=None, T0=None):
         """
+        ds is be what ys should be (not what what passed to predict but more)
         ys should be what was returned by predict
+        T is 1 indexed
         """
         if T0 is None:
             T0 = self.T0
@@ -105,6 +108,7 @@ class EchoStateNetwork(TimeSeriesForecaster):
         train_d = ds[T0:T]
         train_y = ys[T0:T]
         train_rmse = self.rmse(train_d, train_y)
+        # Tf as None means index to the end.
         test_d = ds[T:Tf]
         test_y = ys[T:Tf]
         test_rmse = self.rmse(test_d, test_y)

### 2019-04-12 src/echonn/ml/esn_experiment.py

@@ -1,16 +1,246 @@
-from .time_series_forecasting import TSData
+from copy import deepcopy
+import numpy as np
+from ..sys import SystemSolver
+from . import EchoStateNetwork
 
 
-class EchoStateExperiment:
-    def __init__(self, ts_data):
+class ESNExperiment:
+    def __init__(self, Model, lce_T, time_steps_per_lce_time=1000):
+        """ Model is the dynamical system to test with """
+        self.ModelClass = Model  # assign the class not initialize it
+        self.time_steps_per_lce_time = time_steps_per_lce_time
+
+    # TODO: visualization tools?
+
+    def run(self):
+        """
+        build model, extract data and perform analysis
+
+        Return
+            system integration results: run
+            LCE results: (lce, run)
+            trained model: esn
+            data set: ts_data
+            param performance: {pair: RMSE vector per each network trial}
+        """
+        model_run, lce, lce_run = self.build_model()
+        esn, ts_data, param_performance = self.build_esn(model_run)
+        return model_run, (lce, lce_run), ts_data, esn, param_performance
+
+    def build_model(self):
+        """
+        Generate data
+            Build Model with default params
+            Generate random IC
+            Calc Max LCE from IC
+            Calc Time is 20 times 1 / LCE
+            Calc Step Size is 1 / LCE / 1000 (10,000?)
+            Integrate model
+        """
+        model = self.ModelClass()
+        y0 = model.get_rnd_ic()
+        model_solver = SystemSolver(model)
+        lce_T = model.get_best_lce_T()
+        lce, lce_run = model_solver.get_lce(lce_T, y0)
+        lce_time = 1 / lce
+        Tf = lce_time * 50
+        time_step = lce_time / self.time_steps_per_lce_time
+        T_range = np.arange(0, Tf, time_step)
+        model_run = model_solver.run([0, Tf], y0,
+                                     dense_output=True,
+                                     t_eval=T_range)
+        return model_run, lce, lce_run
+
+    def build_esn(self, model_run):
+        """
+        Analyze data
+            Automate different ESN params and pairs of those params
+            Initialize N models with a pair (recommended ranges from paper)
+            Train all with train data
+            Test with CV many runs
+            performance is end RMSE (long term forecast)
+            Collect AVG performance and best model
+            Use param pair with best average performance
+            Use model with best performance, retrain on train and cv
[…]

### 2019-04-12 src/echonn/ml/time_series_forecaster.py

@@ -33,42 +33,3 @@ class TimeSeriesForecaster(ABC):
 
     def predict(self, *args, **kwargs):
         pass
-
-
-class TSData:
-    def __init__(self, data=None, run=None, split=0.9):
-        """
-        data - is a tuple of t and y of the following format:
-            t[time point]
-            y[dimension, time point]
-        run - is what's returned by the system solver
-        """
-        if run is not None:
-            res = run['results']
-            data = res.t, res.y
-        self.run = run
-        self.t, self.y = data
-        self.N = self.t.shape[0]
-        self.test_index = int(self.N * split)
-        self.cv_index = int(self.N * split**2)
-
-        self.train_t = self.t[:self.cv_index]
-        self.train_y = self.y[:, :self.cv_index]
-
-        self.validation_t = self.t[self.cv_index:self.test_index]
-        self.validation_y = self.y[:, self.cv_index:self.test_index]
-
-        self.test_t = self.t[self.test_index:]
-        self.test_y = self.y[:, self.test_index:]
-
-    @property
-    def train(self):
-        return self.train_t, self.train_y
-
-    @property
-    def validation(self):
-        return self.validation_t, self.validation_y
-
-    @property
-    def test(self):
-        return self.test_t, self.test_y

### 2019-04-12 src/echonn/ml/tools.py

@@ -1,6 +1,7 @@
 import operator
 from functools import reduce
 import numpy as np
+import itertools
 
 
 def product(args):

### 2019-04-12 src/echonn/sys/lorenz_system.py

@@ -2,7 +2,7 @@ from .system import DynamicalSystem
 
 
 class LorenzSystem(DynamicalSystem):
-    def __init__(self, sigma=10, rho=28, beta=8/3, method='BDF'):
+    def __init__(self, sigma=10, rho=28, beta=8/3, method='RK45'):
         super().__init__(3, method)
         self.sigma = sigma
         self.rho = rho

### 2019-04-12 src/echonn/sys/system.py

@@ -11,6 +11,9 @@ class DynamicalSystem(ABC):
     def fun(self, t, v):
         return v
 
+    # TODO get lce T value from empirical results
+    # TODO get random IC
+
     def Dfun(self, t, v):
         ''' This ignores t '''
         eps = np.sqrt(np.finfo(np.float64).eps)

### 2019-04-12 src/echonn/sys/system_solver.py

@@ -58,7 +58,7 @@ class SystemSolver:
         if svd <= 0:
             return -np.inf
         # calculate lyapunov exponent
-        lce = np.log(svds[0])/T
+        lce = np.log(svd)/T
         return lce
 
     def quick_lce(self, T=1000, y0=None, partition=None, **kwargs):

### 2019-04-12 src/echonn/ml/esn_experiment.py
small debugging

@@ -2,10 +2,11 @@ from copy import deepcopy
 import numpy as np
 from ..sys import SystemSolver
 from . import EchoStateNetwork
+import itertools
 
 
 class ESNExperiment:
-    def __init__(self, Model, lce_T, time_steps_per_lce_time=1000):
+    def __init__(self, Model, time_steps_per_lce_time=1000):
         """ Model is the dynamical system to test with """
         self.ModelClass = Model  # assign the class not initialize it
         self.time_steps_per_lce_time = time_steps_per_lce_time

### 2019-04-12 src/echonn/ml/tools.py

@@ -1,7 +1,6 @@
 import operator
 from functools import reduce
 import numpy as np
-import itertools
 
 
 def product(args):

### 2019-04-12 src/echonn/sys/double_pendulum_system.py

@@ -14,7 +14,7 @@ class DoublePendulumSystem(DynamicalSystem):
     """
 
     def __init__(self, m1=1, m2=1, l1=1, l2=1):
-        super().__init__(4, 'BDF')
+        super().__init__(4, 'BDF', 100)
         self.m1 = m1
         self.m2 = m2
         self.l1 = l1

### 2019-04-12 src/echonn/sys/lorenz_system.py

@@ -3,7 +3,7 @@ from .system import DynamicalSystem
 
 class LorenzSystem(DynamicalSystem):
     def __init__(self, sigma=10, rho=28, beta=8/3, method='RK45'):
-        super().__init__(3, method)
+        super().__init__(3, method, 150)
         self.sigma = sigma
         self.rho = rho
         self.beta = beta

### 2019-04-12 src/echonn/sys/restricted_circular_3_body.py

@@ -4,7 +4,7 @@ from .system import DynamicalSystem
 
 class RestrictedCircular3Body(DynamicalSystem):
     def __init__(self, body_ratio=0.5, method='BDF'):
-        super().__init__(4, method)
+        super().__init__(4, method, 100)  # TODO: verify this
         if body_ratio < 0 or 1 < body_ratio:
             raise ValueError('body_ratio should be in range [0, 1]')
         self.mu = body_ratio

### 2019-04-12 src/echonn/sys/system.py

@@ -3,9 +3,10 @@ import numpy as np
 
 
 class DynamicalSystem(ABC):
-    def __init__(self, dim=1, method='RK45'):
+    def __init__(self, dim=1, method='RK45', best_lce_T=1):
         self.dim = dim
         self._method = method
+        self.best_lce_T = best_lce_T
 
     @abstractmethod
     def fun(self, t, v):

### 2019-04-12 src/echonn/sys/system_solver.py

@@ -61,7 +61,7 @@ class SystemSolver:
         lce = np.log(svd)/T
         return lce
 
-    def quick_lce(self, T=1000, y0=None, partition=None, **kwargs):
+    def quick_lce(self, T=None, y0=None, partition=None, **kwargs):
         """
         y0 - will be set randomly and then run through the system for a
         little
@@ -70,6 +70,8 @@ class SystemSolver:
         computation) (defaults to T)
         kwargs - are passed to the run parameters of SystemSolver
         """
+        if T is None:
+            T = self.system.best_lce_T
         if partition is None:
             partition = T
         sys_dim = self.system.dim
@@ -110,13 +112,15 @@ class SystemSolver:
         tspan[-1] = T
         return tspan, lces
 
-    def get_lce(self, T=1000, y0=None, **kwargs):
+    def get_lce(self, T=None, y0=None, **kwargs):
         """
         T - should be set manually to a reasonable value.
         y0 - will be set randomly and then run through the system for a
         little
         kwargs - are passed to the run parameters of SystemSolver
         """
+        if T is None:
+            T = self.system.best_lce_T
         sys_dim = self.system.dim
         # create the lyapunov system
         lce_system = LyapunovSystem(self.system)

### 2019-04-12 src/test/ml/test_experiment.py

@@ -1,5 +1,17 @@
 import unittest
+from echonn.ml import ESNExperiment
+from echonn.sys import LorenzSystem, DoublePendulumSystem, RestrictedCircular3Body
 
 
 class TestExperiment(unittest.TestCase):
-    pass
+    def testLorenz(self):
+        experiment = ESNExperiment(LorenzSystem, 1000)
+        # experiment.run()
+
+    def testDoubPend(self):
+        experiment = ESNExperiment(DoublePendulumSystem, 1000)
+        # experiment.run()
+
+    def test3BodyProblem(self):
+        experiment = ESNExperiment(RestrictedCircular3Body, 1000)
+        # experiment.run()

### 2019-04-12 src/test/sys/test_restrict_3body.py

@@ -8,8 +8,29 @@ class TestModel(unittest.TestCase):
     def testExample(self):
         sys = RestrictedCircular3Body()
         slv = SystemSolver(sys)
+        lce, lce_run = slv.get_lce()
+        slv.calc_lce
+        # print(lce)
         init = [1, -1, 1, -1]
         run = slv.run([0, 5], init)
         run['results'].y = run['results'].y[:2, :]
         slv.plot2d(run)
         # plt.show(True)
+
+    @unittest.skip
+    def testLCE(self):
+        # slv.calc_lce(Df_y0, T)
+        sys = RestrictedCircular3Body()
+        slv = SystemSolver(sys)
+        lce, lce_run = slv.get_lce(T=100)
+        t = lce_run['results'].t
+        y = lce_run['results'].y
+        sys_dim = sys.dim
+        lces = []
+        start_t = 10
+        for i in range(start_t, t.shape[0]):
+            Df_y0 = y[sys_dim:, i].reshape(sys_dim, sys_dim)
+            lce = slv.calc_lce(Df_y0, t[i])
+            lces.append(lce)
+        plt.plot(t[start_t:], lces)
+        plt.show(True)

### 2019-04-12 doc/paper/scripts/LorenzLCEs.py
generate lorenz lce script

@@ -16,7 +16,7 @@ if __name__ == "__main__":
             sys = LorenzSystem(sigma, rho, beta)
             slv = SystemSolver(sys)
 
-            lce, _ = slv.get_lce(T=150)
+            lce, _ = slv.get_lce()
             print('\t{}:'.format(i), lce)
             lces.append(lce)
         res = {}
@@ -33,7 +33,8 @@ if __name__ == "__main__":
         results.append(res)
 
     print('results:')
-    for result in results:
+    for res in results:
         toprint = [res['beta'], res['rho'], res['sigma'], res['lambda'],
-                   res['mean'], res['error'], res['relative error']]
+                   res['mean'], res['relative error']]
+        toprint = [str(v) for v in toprint]
         print('        ' + ' & '.join(toprint))

### 2019-04-13 src/echonn/ml/echo_state_network.py
performance of echo state network is now collectable

@@ -28,7 +28,7 @@ class EchoStateNetwork(TimeSeriesForecaster):
         self.L = L
         self.T0 = T0  # at T0, x[T0] has stabilized
         self.alpha = alpha
-        self.noise = 0.0001 if noise is None else noise
+        self.noise = noise
         self.w = None
         self.Ws = None
         self.Win = None
@@ -79,7 +79,10 @@ class EchoStateNetwork(TimeSeriesForecaster):
         x = self.x
         y = self.y
 
-        noise = np.random.uniform(-self.noise, self.noise)
+        if self.noise is None:
+            noise = 0
+        else:
+            noise = np.random.uniform(-self.noise, self.noise)
         return f(Win @ u[n] + W @ x[n-1] + Wback @ y[n-1] + noise)
 
     def calc_y(self, n):
@@ -91,9 +94,7 @@ class EchoStateNetwork(TimeSeriesForecaster):
         Wout = self.Wout
         u = self.u
         x = self.x
-
-        system_state = np.concatenate((u[n], x[n]))
-        return g(Wout @ system_state)
+        return g(Wout[:, :self.K]@u[n] + Wout[:, self.K:]@x[n])
 
     def score(self, ds, ys, T, Tf=None, T0=None):
         """
@@ -167,6 +168,8 @@ class EchoStateNetwork(TimeSeriesForecaster):
 
     @staticmethod
     def fill_with_zeros(vs, dim):
+        if isinstance(np.array([1, 2, 3]), np.ndarray):
+            return vs.reshape(vs.shape[0], dim)
         vs = [np.zeros(dim) if v is None else np.array(v) for v in vs]
         return np.array(vs).reshape(len(vs), dim)
 

### 2019-04-13 src/echonn/ml/esn_experiment.py

@@ -3,17 +3,22 @@ import numpy as np
 from ..sys import SystemSolver
 from . import EchoStateNetwork
 import itertools
+import time
+import matplotlib.pyplot as plt
 
 
 class ESNExperiment:
-    def __init__(self, Model, time_steps_per_lce_time=1000):
-        """ Model is the dynamical system to test with """
-        self.ModelClass = Model  # assign the class not initialize it
+    def __init__(self, model, alpha=None, N=None, T0=None, trials=2, time_steps_per_lce_time=1000):
+        self.model = model
+        self.alpha = alpha
+        self.N = N
+        self.T0 = T0
+        self.trials = trials
         self.time_steps_per_lce_time = time_steps_per_lce_time
 
     # TODO: visualization tools?
 
-    def run(self):
+    def run(self, verbose=False):
         """
         build model, extract data and perform analysis
 
@@ -24,9 +29,10 @@ class ESNExperiment:
             data set: ts_data
             param performance: {pair: RMSE vector per each network trial}
         """
+        self.verbose = verbose
         model_run, lce, lce_run = self.build_model()
-        esn, ts_data, param_performance = self.build_esn(model_run)
-        return model_run, (lce, lce_run), ts_data, esn, param_performance
+        ts_data, results = self.build_esn(model_run)
+        return model_run, (lce, lce_run), ts_data, results
 
     def build_model(self):
         """
@@ -38,10 +44,12 @@ class ESNExperiment:
             Calc Step Size is 1 / LCE / 1000 (10,000?)
             Integrate model
         """
-        model = self.ModelClass()
+        if self.verbose:
+            print('Collecting data...')
+        model = self.model
         y0 = model.get_rnd_ic()
         model_solver = SystemSolver(model)
-        lce_T = model.get_best_lce_T()
+        lce_T = model.best_lce_T
         lce, lce_run = model_solver.get_lce(lce_T, y0)
         lce_time = 1 / lce
         Tf = lce_time * 50
@@ -66,13 +74,14 @@ class ESNExperiment:
             Run model on testing set and calculate RMSE for every time point
         """
         ts_data = TSData(run=model_run)
-        ts_analysis = TSAnalysis(ts_data)
-        results = ts_analysis.run()
+        ts_analysis = TSAnalysis(
+            ts_data, self.alpha, self.N, self.T0, self.trials)
[…]

### 2019-04-13 src/echonn/sys/double_pendulum_system.py

@@ -25,8 +25,13 @@ class DoublePendulumSystem(DynamicalSystem):
         #self._gl = g / l
         #self._moment_pre = -.5*m*l**2
 
-    def lagrange(self, v):
-        pass
+    def get_rnd_ic(self, **kwargs):
+        a1 = (0.5 + np.random.rand())*pi  # []
+        a2 = 2*np.random.rand()*pi  # [0, 2*pi]
+        w1 = np.random.rand() / 100  # [0, .01)
+        w2 = np.random.rand() / 100  # [0, .01)
+        y0 = np.array([a1, a2, w1, w2])
+        return y0
 
     def fun(self, t, v):
         sin = np.sin
@@ -48,7 +53,7 @@ class DoublePendulumSystem(DynamicalSystem):
         w2p = sin(a1 - a2) * w2_num / w2_den
         return a1p, a2p, w1p, w2p
 
-    """ from wikipedia
+    """ formula from wikipedia
     def fun(self, t, v):
         a1, a2, p1, p2 = v
 

### 2019-04-13 src/echonn/sys/lorenz_system.py

@@ -1,4 +1,6 @@
 from .system import DynamicalSystem
+from .system_solver import SystemSolver
+import numpy as np
 
 
 class LorenzSystem(DynamicalSystem):
@@ -7,6 +9,15 @@ class LorenzSystem(DynamicalSystem):
         self.sigma = sigma
         self.rho = rho
         self.beta = beta
+
+    def get_rnd_ic(self, **kwargs):
+        T = 30
+        y0 = np.random.rand(self.dim)
+        slvr = SystemSolver(self)
+        run = slvr.run([0, T], y0, **kwargs)
+        # tend it towards the attractor
+        y0 = run['results'].y[:, -1]
+        return y0
 
     def fun(self, t, v):
         x, y, z = v

### 2019-04-13 src/echonn/sys/system.py

@@ -12,8 +12,8 @@ class DynamicalSystem(ABC):
     def fun(self, t, v):
         return v
 
-    # TODO get lce T value from empirical results
-    # TODO get random IC
+    def get_rnd_ic(self, **kwargs):
+        return np.random.rand(self.system.dim)
 
     def Dfun(self, t, v):
         ''' This ignores t '''

### 2019-04-13 src/test/ml/test_experiment.py

@@ -1,17 +1,66 @@
+import os
 import unittest
 from echonn.ml import ESNExperiment
 from echonn.sys import LorenzSystem, DoublePendulumSystem, RestrictedCircular3Body
+import pickle
 
 
 class TestExperiment(unittest.TestCase):
+    def setUp(self):
+        self.file_io = True
+        self.verbose = True
+        self.alpha = [.7, .8]
+        self.N = [3]
+        self.T0 = [500]
+        self.trials = 2
+
+    @unittest.skip
     def testLorenz(self):
-        experiment = ESNExperiment(LorenzSystem, 1000)
-        # experiment.run()
+        if self.verbose:
+            print()
+        fn = os.path.join('src', 'test', 'test_data', 'lorenz.p')
+        if self.verbose:
+            print('Testing Lorenz Model')
+        experiment = ESNExperiment(
+            LorenzSystem(), self.alpha,
+            self.N, self.T0, self.trials, 1000)
+        res = experiment.run(self.verbose)
+        if self.file_io:
+            with open(fn, 'wb') as f:
+                pickle.dump(res, f)
+            with open(fn, 'rb') as f:
+                res = pickle.load(f)
+        print(res)
 
+    @unittest.skip
     def testDoubPend(self):
-        experiment = ESNExperiment(DoublePendulumSystem, 1000)
-        # experiment.run()
+        if self.verbose:
+            print()
+        fn = os.path.join('src', 'test', 'test_data', 'doub_pend.p')
+        if self.verbose:
+            print('Testing Double Pendulum Model')
+        experiment = ESNExperiment(
+            DoublePendulumSystem(), self.alpha,
+            self.N, self.T0, self.trials, 1000)
+        res = experiment.run(self.verbose)
+        if self.file_io:
+            with open(fn, 'wb') as f:
+                pickle.dump(res, f)
+            with open(fn, 'rb') as f:
+                res = pickle.load(f)
+        print(res)
 
+    @unittest.skip
     def test3BodyProblem(self):
-        experiment = ESNExperiment(RestrictedCircular3Body, 1000)
-        # experiment.run()
+        if self.verbose:
+            print()
+        fn = os.path.join('src', 'test', 'test_data', '3b.p')
+        if self.verbose:
+            print('Testing 3 Body Problem Model')
+        experiment = ESNExperiment(
+            RestrictedCircular3Body(), self.alpha,
+            self.N, self.T0, self.trials, 1000)
[…]

### 2019-04-13 doc/paper/scripts/AnalyzeDoubPend.py
add doub pend and lorenz explorer

@@ -1,0 +1,13 @@
+import os
+import pickle
+from echonn.sys import DoublePendulumSystem
+from echonn.ml import ESNExperiment
+
+
+fn = os.path.join('.', 'lorenz_results.p')
+print('Testing Lorenz Model')
+experiment = ESNExperiment(DoublePendulumSystem())
+res = experiment.run()
+with open(fn, 'wb') as f:
+    pickle.dump(res, f)
+print(res)

### 2019-04-13 doc/paper/scripts/AnalyzeLorenz.py

@@ -1,0 +1,13 @@
+import os
+import pickle
+from echonn.sys import LorenzSystem
+from echonn.ml import ESNExperiment
+
+
+fn = os.path.join('.', 'lorenz_results.p')
+print('Testing Lorenz Model')
+experiment = ESNExperiment(LorenzSystem())
+res = experiment.run()
+with open(fn, 'wb') as f:
+    pickle.dump(res, f)
+print(res)

### 2019-04-13 src/echonn/ml/echo_state_network.py
implement bias in esn

@@ -5,7 +5,7 @@ from .time_series_forecaster import TimeSeriesForecaster
 
 
 class EchoStateNetwork(TimeSeriesForecaster):
-    def __init__(self, K, N, L, T0=50, alpha=.8, noise=0.0, f=None, g=None, g_inv=None):
+    def __init__(self, K, N, L, T0=50, alpha=.8, noise=0.0, bias=True, f=None, g=None, g_inv=None):
         """
         K - input units, u
         N - internal units, x (should not exceed Tf/10 to Tf/2 at risk of overfitting
@@ -17,10 +17,13 @@ class EchoStateNetwork(TimeSeriesForecaster):
         during sampling. Uniform distribution ranging from [-noise, noise].
         """
         # TODO: implement bias
+        bias_K = K
+        if bias:
+            bias_K += 1
         self.shapes = [
             (N, K),
             (N, N),
-            (L, K+N),
+            (L, bias_K+N),
             (N, L)
         ]
         self.K = K
@@ -35,6 +38,7 @@ class EchoStateNetwork(TimeSeriesForecaster):
         self.W = None
         self.Wout = None
         self.Wback = None
+        self.bias = bias
         self.f = np.tanh if f is None else f
         if g is None:
             self.g = lambda x: x
@@ -75,7 +79,8 @@ class EchoStateNetwork(TimeSeriesForecaster):
         Win = self.Win
         W = self.W
         Wback = self.Wback
-        u = self.u
+        u = self.u[:, :self.K]
+
         x = self.x
         y = self.y
 
@@ -94,7 +99,10 @@ class EchoStateNetwork(TimeSeriesForecaster):
         Wout = self.Wout
         u = self.u
         x = self.x
-        return g(Wout[:, :self.K]@u[n] + Wout[:, self.K:]@x[n])
+        K = self.K
+        if self.bias:
+            K += 1
+        return g(Wout[:, :K]@u[n] + Wout[:, K:]@x[n])
 
     def score(self, ds, ys, T, Tf=None, T0=None):
         """
@@ -114,6 +122,11 @@ class EchoStateNetwork(TimeSeriesForecaster):
         test_y = ys[T:Tf]
         test_rmse = self.rmse(test_d, test_y)
         return (train_d, train_y, train_rmse), (test_d, test_y, test_rmse)
+
+    def add_bias(self):
+        if self.bias:
+            ones_vec = np.ones((self.u.shape[0], 1))
[…]

### 2019-04-13 src/echonn/ml/esn_experiment.py

@@ -8,11 +8,9 @@ import matplotlib.pyplot as plt
 
 
 class ESNExperiment:
-    def __init__(self, model, alpha=None, N=None, T0=None, trials=2, time_steps_per_lce_time=1000):
+    def __init__(self, model, params=None, trials=2, time_steps_per_lce_time=1000):
         self.model = model
-        self.alpha = alpha
-        self.N = N
-        self.T0 = T0
+        self.params = params
         self.trials = trials
         self.time_steps_per_lce_time = time_steps_per_lce_time
 
@@ -74,21 +72,18 @@ class ESNExperiment:
             Run model on testing set and calculate RMSE for every time point
         """
         ts_data = TSData(run=model_run)
-        ts_analysis = TSAnalysis(
-            ts_data, self.alpha, self.N, self.T0, self.trials)
+        ts_analysis = TSAnalysis(ts_data, self.params, self.trials)
         results = ts_analysis.run(self.verbose)
         return ts_data, results
 
 
 class TSAnalysis:
-    def __init__(self, ts_data, alpha=None, N=None, T0=None, trials=2):
+    def __init__(self, ts_data, params, trials=2):
         """
         ts_data is a TSDAta object
         """
         self.ts_data = ts_data
-        self.alpha = alpha
-        self.N = N
-        self.T0 = T0
+        self.params = params
         self.trials = trials
 
     def run_params(self, alpha, N, T0):
@@ -201,10 +196,9 @@ class TSAnalysis:
         std_dev_rmses = []
         if self.verbose:
             self.times = []
-            self.num_pairs = len(
-                list(self.build_param_pairs(self.alpha, self.N, self.T0)))
+            self.num_pairs = len(self.build_param_pairs())
             print('Total Param Pairs:', self.num_pairs)
-        for pair in self.build_param_pairs(self.alpha, self.N, self.T0):
+        for pair in self.build_param_pairs():
             if verbose:
                 self.times.append(time.time())
                 print('Testing Params:', pair)
@@ -260,7 +254,7 @@ class TSAnalysis:
 
         return (bm_train, bm_cv, bm_full_train, bm_test)
 
-    def build_param_pairs(self, alpha=None, N=None, T0=None):
+    def build_param_pairs(self):
         # affects computation
         if alpha is None:
             alpha = [.7, .75, .8, .85, .9, .98]
@@ -277,13 +271,23 @@ class TSAnalysis:
[…]

### 2019-04-13 src/test/ml/test_experiment.py

@@ -1,3 +1,4 @@
+import itertools
 import os
 import unittest
 from echonn.ml import ESNExperiment
@@ -12,6 +13,7 @@ class TestExperiment(unittest.TestCase):
         self.alpha = [.7, .8]
         self.N = [3]
         self.T0 = [500]
+        self.params = itertools.product(self.alpha, self.N, self.T0)
         self.trials = 2
 
     @unittest.skip
@@ -22,8 +24,7 @@ class TestExperiment(unittest.TestCase):
         if self.verbose:
             print('Testing Lorenz Model')
         experiment = ESNExperiment(
-            LorenzSystem(), self.alpha,
-            self.N, self.T0, self.trials, 1000)
+            LorenzSystem(), self.params, self.trials, 1000)
         res = experiment.run(self.verbose)
         if self.file_io:
             with open(fn, 'wb') as f:
@@ -40,8 +41,7 @@ class TestExperiment(unittest.TestCase):
         if self.verbose:
             print('Testing Double Pendulum Model')
         experiment = ESNExperiment(
-            DoublePendulumSystem(), self.alpha,
-            self.N, self.T0, self.trials, 1000)
+            DoublePendulumSystem(), self.params, self.trials, 1000)
         res = experiment.run(self.verbose)
         if self.file_io:
             with open(fn, 'wb') as f:
@@ -58,8 +58,7 @@ class TestExperiment(unittest.TestCase):
         if self.verbose:
             print('Testing 3 Body Problem Model')
         experiment = ESNExperiment(
-            RestrictedCircular3Body(), self.alpha,
-            self.N, self.T0, self.trials, 1000)
+            RestrictedCircular3Body(), self.params, self.trials, 1000)
         res = experiment.run(self.verbose)
         if self.file_io:
             with open(fn, 'wb') as f:

### 2019-04-15 doc/paper/scripts/AnalyzeLorenz.py
generate results

@@ -6,7 +6,18 @@ from echonn.ml import ESNExperiment
 
 fn = os.path.join('.', 'lorenz_results.p')
 print('Testing Lorenz Model')
-experiment = ESNExperiment(LorenzSystem())
+params = [
+    (.7, 300, 300),
+    (.8, 300, 300),
+    (.9, 300, 300),
+    (.98, 300, 300),
+    (.7, 100, 300),
+    (.8, 100, 300),
+    (.9, 100, 300),
+    (.98, 100, 300),
+]
+#params = None
+experiment = ESNExperiment(LorenzSystem(), params=params)
 res = experiment.run(verbose=True)
 with open(fn, 'wb') as f:
     pickle.dump(res, f)

### 2019-04-15 doc/paper/scripts/AnalyzeResultsLorenz.py

@@ -1,0 +1,48 @@
+import pickle
+import numpy as np
+import matplotlib.pyplot as plt
+from copy import deepcopy
+
+from echonn.sys import SystemSolver
+
+res = pickle.load(open('lorenz_results.p', 'rb'))
+run, lce, ts_data, results = res
+
+ds_test, ys_test, total_rmse = results['best model rmse'][0][3]
+
+plt.figure()
+plt.title('Param Config {} RMSE: {}'.format((.7, 500, 300), total_rmse))
+slvr = SystemSolver(run['system'])
+runt = deepcopy(run)
+runt['results'].t = ts_data.test_t
+runt['results'].y = ds_test.T
+fig = slvr.plot3d(runt)
+runt['results'].t = ts_data.test_t
+runt['results'].y = ys_test.T
+slvr.plot3d(runt, fig=fig)
+plt.show(True)
+
+# for i, rmse in enumerate(results['best model rmse']):
+#test_rmse = rmse[3]
+#ds_test, ys_test, total_rmse = test_rmse
+#
+# plt.figure()
+#plt.title('Param Config {} RMSE: {}'.format(i, total_rmse))
+#slvr = SystemSolver(run['system'])
+#runt = deepcopy(run)
+#runt['results'].t = ts_data.test_t
+#runt['results'].y = ds_test.T
+#fig = slvr.plot3d(runt)
+#runt['results'].t = ts_data.test_t
+#runt['results'].y = ys_test.T
+#slvr.plot3d(runt, fig=fig)
+# print()
+#rerr = np.mean((ys_test - ds_test) / ds_test, axis=1)
+# print(rerr)
+# print(np.mean(rerr))
+# print(total_rmse)
+
+# plt.show(True)
+#plt.plot(ts_data.test_t, ds_test)
+#plt.plot(ts_data.test_t, np.mean((ds_test - ys_test) / ds_test, axis=1))
+# plt.show(True)

### 2019-04-15 src/echonn/ml/echo_state_network.py

@@ -40,14 +40,6 @@ class EchoStateNetwork(TimeSeriesForecaster):
         self.Wback = None
         self.bias = bias
         self.f = np.tanh if f is None else f
-        if g is None:
-            self.g = lambda x: x
-            self.g_inv = self.g
-        else:
-            self.g = g
-            if g_inv is None:
-                raise Exception('g_inv must be specified with g')
-            self.g_inv = g_inv
 
         # set on train
         self.n = 0
@@ -56,6 +48,12 @@ class EchoStateNetwork(TimeSeriesForecaster):
         self.u = [np.zeros(self.K)]
         self.x = [np.zeros(self.N)]
         self.y = [np.zeros(self.L)]
+
+    def g(self, x):
+        return x
+
+    def g_inv(self, x):
+        return x
 
     def init_weights(self):
         self.w, self.Ws = init_weights(self.shapes)

### 2019-04-15 src/echonn/ml/esn_experiment.py

@@ -8,8 +8,9 @@ import matplotlib.pyplot as plt
 
 
 class ESNExperiment:
-    def __init__(self, model, params=None, trials=2, time_steps_per_lce_time=1000):
+    def __init__(self, model, data=None, params=None, trials=2, time_steps_per_lce_time=100):
         self.model = model
+        self.data = data
         self.params = params
         self.trials = trials
         self.time_steps_per_lce_time = time_steps_per_lce_time
@@ -23,12 +24,21 @@ class ESNExperiment:
         Return
             system integration results: run
             LCE results: (lce, run)
-            trained model: esn
             data set: ts_data
-            param performance: {pair: RMSE vector per each network trial}
+            param performance results:
+                {
+                    'params': [ param pairs, ... ]
+                    'best model': [ best model for param (based on cv), ... ]
+                    'best model rmse': [ (train, cv, full_train, test), ... ]
+                    'avg rmse': [ (train, cv), ... ]
+                    'std dev rmse': [ (train, cv), ... ]
+                }
         """
         self.verbose = verbose
-        model_run, lce, lce_run = self.build_model()
+        if self.data is None:
+            model_run, lce, lce_run = self.build_model()
+        else:
+            model_run, lce, lce_run = self.data
         ts_data, results = self.build_esn(model_run)
         return model_run, (lce, lce_run), ts_data, results
 
@@ -94,7 +104,6 @@ class TSAnalysis:
         t_validation, y_validation = self.ts_data.validation
         esns = []
         rmses = []
-        param_timer = time.time()
         for i in range(self.trials):
             if self.verbose:
                 print('\tRunning Trial {}/{} ... '.format(i+1, self.trials))
@@ -183,22 +192,22 @@ class TSAnalysis:
             {
                 'params': [ param pairs, ... ]
                 'best model': [ best model for param (based on cv), ... ]
-                'best model rmse': [ (train, cv, full_train, test), ... ]
+                'best model rmse': [ (train, cv, full_train, (ds_test, ys_test, rmse_test)), ... ]
                 'avg rmse': [ (train, cv), ... ]
                 'std dev rmse': [ (train, cv), ... ]
             }
         """
         self.verbose = verbose
-        params = []
+        params = self.build_param_pairs()
         best_models = []
         best_model_rmses = []
         avg_rmses = []
         std_dev_rmses = []
         if self.verbose:
             self.times = []
[…]

### 2019-04-15 doc/paper/scripts/AnalyzeLorenz.py
udpate

@@ -7,6 +7,10 @@ from echonn.ml import ESNExperiment
 fn = os.path.join('.', 'lorenz_results.p')
 print('Testing Lorenz Model')
 params = [
+    (.7, 500, 300),
+    (.8, 500, 300),
+    (.9, 500, 300),
+    (.98, 500, 300),
     (.7, 300, 300),
     (.8, 300, 300),
     (.9, 300, 300),

### 2019-04-15 src/echonn/ml/esn_experiment.py
use differential not output

@@ -8,8 +8,9 @@ import matplotlib.pyplot as plt
 
 
 class ESNExperiment:
-    def __init__(self, model, data=None, params=None, trials=2, time_steps_per_lce_time=100):
+    def __init__(self, model, data=None, params=None, trials=2, t_len=50, time_steps_per_lce_time=100):
         self.model = model
+        self.t_len = t_len
         self.data = data
         self.params = params
         self.trials = trials
@@ -60,7 +61,7 @@ class ESNExperiment:
         lce_T = model.best_lce_T
         lce, lce_run = model_solver.get_lce(lce_T, y0)
         lce_time = 1 / lce
-        Tf = lce_time * 50
+        Tf = lce_time * self.t_len
         time_step = lce_time / self.time_steps_per_lce_time
         T_range = np.arange(0, Tf, time_step)
         model_run = model_solver.run([0, Tf], y0,
@@ -314,7 +315,8 @@ class TSData:
             data = res.t, res.y.T
         self.run = run
         self.t, self.y = data
-        self.y = self.y
+        self.t = self.t[:-1]
+        self.y = self.y[:-1] - self.y[1:]
         self.N, self.dim = self.y.shape
         self.test_index = int(self.N * split)
         self.cv_index_end = self.test_index

### 2019-04-15 src/test/ml/test_esn.py

@@ -1,4 +1,7 @@
 from echonn.ml import EchoStateNetwork
+import itertools
+from echonn.ml import ESNExperiment
+from echonn.sys import LorenzSystem
 import numpy as np
 import unittest
 
@@ -193,5 +196,23 @@ class TestEchoStateNetwork(unittest.TestCase):
                 return
         raise Exception('noise paper experiment failed')
 
+    @unittest.skip
     def testBiasWorks(self):
-        pass  # TODO enable bias in esn
+        class Run:
+            def __init__(self, t, y):
+                self.t = t
+                self.y = y
+        lce = 1
+        t = np.arange(0, 50, 1/100, dtype=np.float64)
+        y = np.array([np.sin(t), np.cos(t), np.tanh(t)])
+        run = {'results': Run(t, y)}
+        data = (run, lce, None)
+        ESNExperiment
+        alphas = [.7, .75, .8, .85]
+        Ns = [10, 20, 30, 40]
+        T0s = [50, 100, 150]
+        params = list(itertools.product(alphas, Ns, T0s))
+        experiment = ESNExperiment(
+            LorenzSystem(), data=data, params=params, trials=3)
+        res = experiment.run(verbose=True)
+        print(res)

### 2019-04-16 src/echonn/ml/esn_experiment.py
default split to .9

@@ -302,7 +302,7 @@ class TSAnalysis:
 
 
 class TSData:
-    def __init__(self, data=None, run=None, split=0.7):
+    def __init__(self, data=None, run=None, split=0.9):
         """
         data - is a tuple of t and y of the following format:
             t[time point]
@@ -314,9 +314,9 @@ class TSData:
             res = run['results']
             data = res.t, res.y.T
         self.run = run
-        self.t, self.y = data
-        self.t = self.t[:-1]
-        self.y = self.y[:-1] - self.y[1:]
+        self.t_orig, self.y_orig = data
+        self.t = self.t_orig[:-1]
+        self.y = self.y_orig[:-1] - self.y_orig[1:]
         self.N, self.dim = self.y.shape
         self.test_index = int(self.N * split)
         self.cv_index_end = self.test_index

### 2019-04-16 doc/paper/scripts/AnalyzeDoubPend.py
lorenz result photos

@@ -1,3 +1,4 @@
+import itertools
 import os
 import pickle
 from echonn.sys import DoublePendulumSystem
@@ -6,7 +7,12 @@ from echonn.ml import ESNExperiment
 
 fn = os.path.join('.', 'doub_pend_results.p')
 print('Testing Double Pendulum Model')
-experiment = ESNExperiment(DoublePendulumSystem())
+alphas = [.7, .75, .8, .85, .9, .98]
+Ns = [10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 150, 200, 300]
+T0s = [50, 100, 150, 200, 300, 500]
+params = list(itertools.product(alphas, Ns, T0s))
+experiment = ESNExperiment(DoublePendulumSystem(), params=params,
+                           t_len=50, trials=3, time_steps_per_lce_time=100)
 res = experiment.run(verbose=True)
 with open(fn, 'wb') as f:
     pickle.dump(res, f)

### 2019-04-16 doc/paper/scripts/AnalyzeLorenz.py

@@ -1,3 +1,4 @@
+import itertools
 import os
 import pickle
 from echonn.sys import LorenzSystem
@@ -6,22 +7,15 @@ from echonn.ml import ESNExperiment
 
 fn = os.path.join('.', 'lorenz_results.p')
 print('Testing Lorenz Model')
-params = [
-    (.7, 500, 300),
-    (.8, 500, 300),
-    (.9, 500, 300),
-    (.98, 500, 300),
-    (.7, 300, 300),
-    (.8, 300, 300),
-    (.9, 300, 300),
-    (.98, 300, 300),
-    (.7, 100, 300),
-    (.8, 100, 300),
-    (.9, 100, 300),
-    (.98, 100, 300),
-]
-#params = None
-experiment = ESNExperiment(LorenzSystem(), params=params)
+alphas = [.7, .75, .8, .85, .9, .98]
+#alphas = [.7, .75, .8]
+Ns = [10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 150, 200, 300]
+#Ns = [100, 150, 200]
+T0s = [50, 100, 150, 200, 300, 500]
+#T0s = [300]
+params = list(itertools.product(alphas, Ns, T0s))
+experiment = ESNExperiment(LorenzSystem(), params=params,
+                           t_len=50, trials=3, time_steps_per_lce_time=100)
 res = experiment.run(verbose=True)
 with open(fn, 'wb') as f:
     pickle.dump(res, f)

### 2019-04-16 doc/paper/scripts/AnalyzeResultsDoubPend.py

@@ -1,12 +1,11 @@
-import os
 import pickle
+import os
 import numpy as np
 import matplotlib.pyplot as plt
 from copy import deepcopy
+from echonn.sys import SystemSolver, DoublePendulumAnimator
 
-from echonn.sys import SystemSolver
-
-res = pickle.load(open('lorenz_results.p', 'rb'))
+res = pickle.load(open('doub_pend_results.p', 'rb'))
 run, lce, ts_data, results = res
 
 test_rmse = [rmse for _, _, _, (_, _, rmse) in results['best model rmse']]
@@ -19,30 +18,22 @@ def rmse(d, y):
 
 
 score = []
-higher_score = []
 for rmse_res in results['best model rmse']:
     ds_test, ys_test, total_rmse = rmse_res[3]
     for sub in range(ds_test.shape[0]):
         err = rmse(ds_test[:sub+1], ys_test[:sub+1])
-        if err > .05:
+        if err > .2:
             score.append(sub)
             break
-    for sub in range(ds_test.shape[0]):
-        err = rmse(ds_test[:sub+1], ys_test[:sub+1])
-        if err > 1:
-            higher_score.append(sub)
-            break
-dir_pre = os.path.join('..', 'images', 'lorenz')
-slvr = SystemSolver(run['system'])
-runt = deepcopy(run)
-runt['results'].t = ts_data.t
-runt['results'].y = ts_data.y.T
-fig = slvr.plot3d(runt)
-plt.savefig(os.path.join(dir_pre, 'full_differential.png'))
+
+# doub_pend_anim = os.path.join('..', 'images', 'doub_pend_anim')
+# anim = DoublePendulumAnimator(run)
+# anim.render()
+# anim.save(doub_pend_anim)
 
 sorter = np.flip(np.argsort(score))
 how_many = 10
-for rank, i in enumerate(sorter[:how_many]):
+for i in sorter[:how_many]:
     ds_test, ys_test, total_rmse = results['best model rmse'][i][3]
     print(i, total_rmse, results['params'][i])
 
@@ -50,33 +41,23 @@ for rank, i in enumerate(sorter[:how_many]):
                    for sub in range(ds_test.shape[0])]
 
     plt.figure()
-    plt.title('RMSE vs Lyapunov Time\nParam {} ; RMSE {}'.format(
+    plt.title('Param Config {} RMSE: {}'.format(
[…]

### 2019-04-16 src/echonn/sys/__init__.py

@@ -7,3 +7,4 @@ from .n_body_system import NBodySystem
 from .animator import Animator
 from .lyapunov_system import LyapunovSystem
 from .restricted_circular_3_body import RestrictedCircular3Body
+from .double_pendulum_animator import DoublePendulumAnimator

### 2019-04-16 doc/paper/paper.tex

@@
 the previous calculation. However, the story of the discover is that while
-the computer used 6 digit accuracy, it output only 3 digits, so when the
+the computer used 6 digit accuracy, it only output 3 digits, so when the
 simulation was continued by Lorenz with the less accurate measurements, he
@@
 
+\label{sec:calc_lya}
 In order to find the Lyapunov time, we only need to solve for the largest
@@
     \begin{tabular}{|l|l|l|l|l|l|l|}
-         % TODO: \lambda_1?
-         % TODO: calculated values
         \hline
@@
         \hline \hline
-        16 & 45.92 & 4 & 1.50255 & & \\ % DATA POINTS
-        16 & 40 & 4 & 1.37446 & & \\
-        10 & 28 & 8/3 & 0.90566 & & \\
+        4 & 45.92 & 16 & 1.50255 & 1.49869 & -2.5694 $\times 10^{-3}$ \\ % DATA POINTS
+        4 & 40 & 16 & 1.37446 & 1.37297 & -1.0876 $\times 10^{-3}$ \\
+        8/3 & 28 & 10 & 0.90566 & 0.91188 & 6.8694 $\times 10^{-3}$ \\
         \hline
@@
 While it is possible to use a symplectic integrator, it is not necessary.
-Under the right initial conditions, the third body will diverge regardless.
-The simple divergent spiral isn't chaotic, so instead orbital and collision
-less initial conditions will be used. Without divergence and without
-collisions, the symplectic integrator is unnecessary.
+Under the right initial conditions, the third body can diverge regardless
+(fast initial velocity away from the bodies). The simple divergent spiral
+isn't chaotic, so instead orbital and collision less initial conditions
+will be used. Without divergence and without collisions, the symplectic
+integrator is unnecessary.
 
@@
 The output layer is connected to the current input layer and current 
-internal layer. The weight for these connections will be denoted $\Wout$ 
+internal layer. Additionally, a bias input is connected to this layer.
+The weight for these connections will be denoted $\Wout$ 
 (we only have one matrix since the connected layers are concatenated into a
@@
         \right) \\
-    \yv(t_i) &= \Wout (\uv(t_i), \xv(t_i))_{\text{concat}}.
+    \yv(t_i) &= \Wout (\uv(t_i), 1, \xv(t_i))_{\text{concat}}.
 \end{align}
@@
 unit values and internal state value pairs
-$(\uv(t_{T_0}), \xv(t_{T_0})), \dots, (\uv(t_T), \xv(t_T))$
+$(\uv(t_{T_0}), 1, \xv(t_{T_0})), \dots, (\uv(t_T), 1, \xv(t_T))$
 in its rows. $T$ will store the output unit values
@@
 \subsection{Method}
+
+First, data from the models is collected. Using the process described
+above (Section \ref{sec:calc_lya}), the Largest Lyapunov Exponent is
+calculated, $\lambda_1$. Then, the Lyapunov Time is calculated by
+$\frac{1}{\lambda_1}$. We will then integrate the model with an algorithm
+called RK4(5) over the interval
+$[0, \frac{50}{\lambda_1}]$ with time steps of $\frac{1}{100 \lambda_1}$. 
+This will result in 5,000 data points.
+
+Then, the data is partitioned into a training set, a validation set, and 
+a testing set. I use a 90\% split, so testing is the last 10\% of the
+data. The training and validation set are the first 90\%. Of the 90\%,
+training is the first 90\%, and validation is the final 10\%. In total,
+it's a 81\%-9\%-10\% split.
+
+The training set is used to train the model. The validation
+set is used to tune the parameters of the model ($\alpha, N, T_0$). The
+testing set is used to evaluate the performance.
+
+The first data point of the training set is $t_0 = 0$. The parameter
+$T_0$ marks where the model begins training on the data (see Section
+\ref{sec:etrain}). Using slightly different notation to account for the
+extra validation set, we will denote the first validation time point by
+$t = T$. We will denote the first testing time point by $t = T_f$
+
+Performance of the model is calculated by RMSE for a given time interval
+$[t_0, t_f]$:
+
+\begin{align}
+    \text{RMSE} = \sqrt{\sum^{t_f}_{t=t_0} (\dv(t) - \yv(t))^2}
+\end{align}
+
+where $(\cdot)^2$ is the dot product of the error with itself,
+$\dv(t)$ is the true value at time $t$,
+and $\yv(t)$ is the model's approximation at time $t$.
+
+First, we generate a list of parameter combinations to try. Usually, this
+is the Cartesian product of typical parameter values for the model. 
+For each combination, we train several models on the training data. Then,
+we predict the validation data to measure the general performance of the
+model under those parameters.
+
+$\text{RMSE}_\text{train}$ is calculated on the interval $[T_0, T)$,
+and the prediction uses teacher-forced values on $[0, T_0)$.
+$\text{RMSE}_\text{validation}$ is calculated on the interval $[T, T_f)$,
+and the prediction uses teacher-forced values on $[0, T)$.
+
+Then, the models which performed best on the validation set should also
+perform well on the testing set. The models are retrained on the testing
+and validation set, and the error is calculated:
+$\text{RMSE}_\text{full train}$ is calculated on the interval $[T, T_f)$,
+and the prediction uses teacher-forced values on $[0, T)$.
+
+Finally, the error for the testing set is calculated. 
+The performance on the testing set is 
+an unbias representation of how well the ESN can fit the model.
+
+However, the $\text{RMSE}_\text{test}$ is calculated slightly different
+since we want to know how far the model can predict before deviating too
+far. The $\text{RMSE}_\text{test}(t_f)$ will be calculated on the interval
+$[T_f, t_f)$ and the prediction uses teacher-forced values on $[0, T_f)$.
+Lastly, $\text{RMSE}_\text{test}$ will notate the RMSE for the entire test
+set.
+
+We will look at how far past $T_f$ the ESN model's
+$\text{RMSE}_\text{test}$ grows too large.
+
 \subsection{Results}
+\subsubsection{Lorenz}
+
+One of the issues run into while forecasting with the Lorenz is that the
+continuous nature of the equation led to the ESN just outputing the
+previous value. For initial values far from the center of the Lorenz
+Butterfly, it would tend towards the center and stop when it does arrive
+there. Most models trained poorly and gave $\text{RMSE}_\text{test}$
+values of 16 or well above 10,000, implying the model diverged.
+
+To fix this issue, we instead have the ESN train on differences between
+time points. This reduces our data set to 4,999 values, and it creates
+a new shape of the data. The Lorenz Butterfly had a similar shape, two rings, but they intersected differently (See figure \ref{fig:lorenz_diff}).
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.5\linewidth]{images/lorenz/test.png}
+    \caption{Largest Lyapunov Characteristic Exponent vs Inner Theta Initial Condition}
+    \label{fig:lorenz_diff}
+\end{figure}
+
+\subsubsection{Double Pendulum}
+\subsubsection{Restricted Circular 3 Body}
 

### 2019-04-17 src/echonn/ml/esn_experiment.py
enable disable diff

@@ -8,10 +8,11 @@ import matplotlib.pyplot as plt
 
 
 class ESNExperiment:
-    def __init__(self, model, data=None, params=None, trials=2, t_len=50, time_steps_per_lce_time=100):
+    def __init__(self, model, data=None, params=None, use_diff=True, trials=2, t_len=50, time_steps_per_lce_time=100):
         self.model = model
         self.t_len = t_len
         self.data = data
+        self.use_diff = use_diff
         self.params = params
         self.trials = trials
         self.time_steps_per_lce_time = time_steps_per_lce_time
@@ -82,7 +83,7 @@ class ESNExperiment:
             Use model with best performance, retrain on train and cv
             Run model on testing set and calculate RMSE for every time point
         """
-        ts_data = TSData(run=model_run)
+        ts_data = TSData(run=model_run, use_diff=self.use_diff)
         ts_analysis = TSAnalysis(ts_data, self.params, self.trials)
         results = ts_analysis.run(self.verbose)
         return ts_data, results
@@ -302,7 +303,7 @@ class TSAnalysis:
 
 
 class TSData:
-    def __init__(self, data=None, run=None, split=0.9):
+    def __init__(self, data=None, run=None, use_diff=True, split=0.9):
         """
         data - is a tuple of t and y of the following format:
             t[time point]
@@ -315,8 +316,12 @@ class TSData:
             data = res.t, res.y.T
         self.run = run
         self.t_orig, self.y_orig = data
-        self.t = self.t_orig[:-1]
-        self.y = self.y_orig[:-1] - self.y_orig[1:]
+        if use_diff:
+            self.t = self.t_orig[:-1]
+            self.y = self.y_orig[:-1] - self.y_orig[1:]
+        else:
+            self.t = self.t_orig
+            self.y = self.y_orig
         self.N, self.dim = self.y.shape
         self.test_index = int(self.N * split)
         self.cv_index_end = self.test_index

### 2019-04-17 doc/paper/paper.tex

@@
 We will look at how far past $T_f$ the ESN model's
-$\text{RMSE}_\text{test}$ grows too large.
+$\text{RMSE}_\text{test}$ grows too large. The best model will be the 
+model with the lowest Testing RMSE for the longest time.
 
@@
 time points. This reduces our data set to 4,999 values, and it creates
-a new shape of the data. The Lorenz Butterfly had a similar shape, two rings, but they intersected differently (See figure \ref{fig:lorenz_diff}).
+a new shape of the data. The Lorenz Butterfly had a similar shape, two rings, but in this case, the rings intersect (See figure \ref{fig:lorenz_diff}).
 
@@
     \centering
-    \includegraphics[width=.5\linewidth]{images/lorenz/test.png}
-    \caption{Largest Lyapunov Characteristic Exponent vs Inner Theta Initial Condition}
+    \includegraphics[width=.5\linewidth]{doc/paper/images/lorenz/full_differential.png}
+    \caption{A plot of the differences between time steps of an approximation of the Lorenz Butterfly}
     \label{fig:lorenz_diff}
+\end{figure}
+
+After this transformation of the data, the model performed significantly
+better. $\text{RMSE}_\text{test}$ dropped to ranges of 1-2, the
+model exploded (diverged to infinity) less frequently, and it was no
+longer outputting previous values.
+
+Most well performing models couldn't predict past one Lyapunov constant. 
+From the graphs, we can see that most models deviated once reaching the
+intersection of the butterfly wings.
+
+Once reaching the intersection,
+several things can happen. First, it can deviate completely and stop
+following either wings. Second, it can follow the wrong wing. Which wing
+the trajectory follows seems almost random, so it makes sense if it
+chooses the wrong one. Third, it will continue to follow the trajectory,
+but slightly perturbed. With every passing of the intersection, any of
+these can happen. Eventually, even the perturbations causes problems. 
+The model was trained on data that existed in a specific range of space. 
+When the perturbations lead the model into space not explored, 
+it performs even worse, and this can cause very sporastic behavior.
+
+The chaotic behavior of the Lorenz model dictates which wing the
+trajectory will follow. This is difficult for the ESN to be able to
+model. However, it is possible that I am not using enough data to train
+the ESN, and due to resource limitations, I may have not been able to find
+the correct parameters to achieve a more perfect performance.
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_0_param_225_fit.png}
+        \caption{The True Figure vs ESN Prediction}
+        \label{fig:lorenz_r0_fit}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_0_param_225_rmse.png}
+        \caption{RMSE Over Time}
+        \label{fig:lorenz_r0_rmse}
+    \end{subfigure}
+    \caption{The model diverges when entering never seen region.}
+    \label{fig:lorenz_r0}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_1_param_230_fit.png}
+        \caption{The True Figure vs ESN Prediction}
+        \label{fig:lorenz_r1_fit}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_1_param_230_rmse.png}
+        \caption{RMSE Over Time}
+        \label{fig:lorenz_r1_rmse}
+    \end{subfigure}
+    \caption{The model continues in wrong direction after intersection.}
+    \label{fig:lorenz_r1}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_2_param_458_fit.png}
+        \caption{The True Figure vs ESN Prediction}
+        \label{fig:lorenz_r2_fit}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_2_param_458_rmse.png}
+        \caption{RMSE Over Time}
+        \label{fig:lorenz_r2_rmse}
+    \end{subfigure}
+    \caption{The model has completely deviated after an intersection.}
+    \label{fig:lorenz_r2}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_3_param_151_fit.png}
+        \caption{The True Figure vs ESN Prediction}
+        \label{fig:lorenz_r3_fit}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_3_param_151_rmse.png}
+        \caption{RMSE Over Time}
+        \label{fig:lorenz_r3_rmse}
+    \end{subfigure}
+    \caption{Fits model really well, but follows wrong wings.}
+    \label{fig:lorenz_r3}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_4_param_300_fit.png}
+        \caption{The True Figure vs ESN Prediction}
+        \label{fig:lorenz_r4_fit}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_4_param_300_rmse.png}
+        \caption{RMSE Over Time}
+        \label{fig:lorenz_r4_rmse}
+    \end{subfigure}
+    \caption{Fits model really well, but follows wrong wings.}
+    \label{fig:lorenz_r4}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_5_param_387_fit.png}
+        \caption{The True Figure vs ESN Prediction}
+        \label{fig:lorenz_r5_fit}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_5_param_387_rmse.png}
+        \caption{RMSE Over Time}
+        \label{fig:lorenz_r5_rmse}
+    \end{subfigure}
+    \caption{Diverges after reaching unknown space.}
+    \label{fig:lorenz_r5}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_6_param_222_fit.png}
+        \caption{The True Figure vs ESN Prediction}
+        \label{fig:lorenz_r6_fit}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_6_param_222_rmse.png}
+        \caption{RMSE Over Time}
+        \label{fig:lorenz_r6_rmse}
+    \end{subfigure}
+    \caption{Doesn't fit model smoothy and eventually diverges.}
+    \label{fig:lorenz_r6}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_7_param_73_fit.png}
+        \caption{The True Figure vs ESN Prediction}
+        \label{fig:lorenz_r7_fit}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_7_param_73_rmse.png}
+        \caption{RMSE Over Time}
+        \label{fig:lorenz_r7_rmse}
+    \end{subfigure}
+    \caption{Sporadic behavior shown long after divergence.}
+    \label{fig:lorenz_r7}
 \end{figure}

### 2019-04-17 doc/paper/scripts/AnalyzeDoubPend.py
update scripts

@@ -11,7 +11,7 @@ alphas = [.7, .75, .8, .85, .9, .98]
 Ns = [10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 150, 200, 300]
 T0s = [50, 100, 150, 200, 300, 500]
 params = list(itertools.product(alphas, Ns, T0s))
-experiment = ESNExperiment(DoublePendulumSystem(), params=params,
+experiment = ESNExperiment(DoublePendulumSystem(), params=params, use_diff=False,
                            t_len=50, trials=3, time_steps_per_lce_time=100)
 res = experiment.run(verbose=True)
 with open(fn, 'wb') as f:

### 2019-04-17 doc/paper/scripts/AnalyzeResultsDoubPend.py

@@ -1,13 +1,16 @@
-import pickle
 import os
+import pickle
 import numpy as np
 import matplotlib.pyplot as plt
 from copy import deepcopy
-from echonn.sys import SystemSolver, DoublePendulumAnimator
+
+from echonn.sys import SystemSolver
 
 res = pickle.load(open('doub_pend_results.p', 'rb'))
 run, lce, ts_data, results = res
 
+print('lce:', lce[0])
+print(run['results'].y[:, 0])
 test_rmse = [rmse for _, _, _, (_, _, rmse) in results['best model rmse']]
 test_rmse = np.array(test_rmse)
 
@@ -18,22 +21,31 @@ def rmse(d, y):
 
 
 score = []
+higher_score = []
 for rmse_res in results['best model rmse']:
     ds_test, ys_test, total_rmse = rmse_res[3]
     for sub in range(ds_test.shape[0]):
         err = rmse(ds_test[:sub+1], ys_test[:sub+1])
-        if err > .2:
+        if err > .05:
             score.append(sub)
             break
-
-# doub_pend_anim = os.path.join('..', 'images', 'doub_pend_anim')
-# anim = DoublePendulumAnimator(run)
-# anim.render()
-# anim.save(doub_pend_anim)
+    for sub in range(ds_test.shape[0]):
+        err = rmse(ds_test[:sub+1], ys_test[:sub+1])
+        if err > 1:
+            higher_score.append(sub)
+            break
+dir_pre = os.path.join('..', 'images', 'doub_pend')
+slvr = SystemSolver(run['system'])
+runt = deepcopy(run)
+runt['results'].t = ts_data.t
+runt['results'].y = ts_data.y.T
+fig = slvr.plotnd(runt)
+plt.savefig(os.path.join(dir_pre, 'full_differential.png'))
+# plt.show(True)
 
 sorter = np.flip(np.argsort(score))
 how_many = 10
-for i in sorter[:how_many]:
+for rank, i in enumerate(sorter[:how_many]):
     ds_test, ys_test, total_rmse = results['best model rmse'][i][3]
     print(i, total_rmse, results['params'][i])
 
@@ -41,52 +53,31 @@ for i in sorter[:how_many]:
                    for sub in range(ds_test.shape[0])]
 
     plt.figure()
-    plt.title('Param Config {} RMSE: {}'.format(
+    plt.title('RMSE vs Lyapunov Time\nParam {} ; RMSE {}'.format(
[…]

### 2019-04-17 doc/paper/scripts/AnalyzeResultsLorenz.py

@@ -9,6 +9,8 @@ from echonn.sys import SystemSolver
 res = pickle.load(open('lorenz_results.p', 'rb'))
 run, lce, ts_data, results = res
 
+print('lce:', lce[0])
+print(run['results'].y[:, 0])
 test_rmse = [rmse for _, _, _, (_, _, rmse) in results['best model rmse']]
 test_rmse = np.array(test_rmse)
 

### 2019-04-17 doc/paper/scripts/AnalyzeResultsLorenz.py
cleanup

@@ -79,35 +79,3 @@ for rank, i in enumerate(sorter[:how_many]):
         dir_pre, name))
     plt.close()
     # plt.show(True)
-
-
-def plot_one(ind):
-    ds_test, ys_test, total_rmse = results['best model rmse'][ind][3]
-    plt.show(True)
-
-
-def plot_res():
-    for i, rmse in enumerate(results['best model rmse']):
-        test_rmse = rmse[3]
-        ds_test, ys_test, total_rmse = test_rmse
-
-        rerr = np.mean((ys_test - ds_test) / ds_test, axis=1)
-        rerr = np.sqrt(rerr**2)
-
-        plt.figure()
-        plt.title('Param Config {} RMSE: {}'.format(
-            results['params'][i], total_rmse))
-        plt.plot((ts_data.test_t - ts_data.test_t[0]) / lce[0], rerr)
-
-        slvr = SystemSolver(run['system'])
-        runt = deepcopy(run)
-        runt['results'].t = ts_data.test_t
-        runt['results'].y = ds_test.T
-        slvr.plotnd(runt)
-        runt['results'].t = ts_data.test_t
-        runt['results'].y = ys_test.T
-        slvr.plotnd(runt)
-        slvr.plot3d(runt)
-        plt.title('Param Config {} RMSE: {}'.format(
-            results['params'][i], total_rmse))
-        plt.show(True)

### 2019-04-17 doc/paper/scripts/AnalyzeResultsDoubPend.py
add doub pend results

@@ -29,11 +29,6 @@ for rmse_res in results['best model rmse']:
         if err > .05:
             score.append(sub)
             break
-    for sub in range(ds_test.shape[0]):
-        err = rmse(ds_test[:sub+1], ys_test[:sub+1])
-        if err > 1:
-            higher_score.append(sub)
-            break
 dir_pre = os.path.join('..', 'images', 'doub_pend')
 slvr = SystemSolver(run['system'])
 runt = deepcopy(run)
@@ -43,8 +38,12 @@ fig = slvr.plotnd(runt)
 plt.savefig(os.path.join(dir_pre, 'full_differential.png'))
 # plt.show(True)
 
+runt['results'].t = ts_data.test_t
+runt['results'].y = ds_test.T
+slvr.plotnd(runt)
+plt.savefig(os.path.join(dir_pre, 'test_data.png'))
 sorter = np.flip(np.argsort(score))
-how_many = 10
+how_many = 5
 for rank, i in enumerate(sorter[:how_many]):
     ds_test, ys_test, total_rmse = results['best model rmse'][i][3]
     print(i, total_rmse, results['params'][i])
@@ -63,19 +62,8 @@ for rank, i in enumerate(sorter[:how_many]):
     plt.close()
 
     runt['results'].t = ts_data.test_t
-    runt['results'].y = ds_test.T
-
-    fig = slvr.plotnd(runt)
-    runt['results'].t = ts_data.test_t[:score[i]]
-    runt['results'].y = ys_test[:score[i]].T
-    slvr.plotnd(runt, fig)
-    runt['results'].t = ts_data.test_t[score[i]:higher_score[i]]
-    runt['results'].y = ys_test[score[i]:higher_score[i]].T
-    slvr.plotnd(runt, fig)
-    plt.legend(['actual', 'approximated', 'deviation'])
-    plt.title('Model Approximation\nParam {} ; RMSE {}'.format(
-        results['params'][i], total_rmse))
-    plt.tight_layout()
+    runt['results'].y = ys_test.T
+    slvr.plotnd(runt)
     name = 'rank_{}_param_{}_fit.png'.format(rank, i)
     plt.savefig(os.path.join(
         dir_pre, name))

### 2019-04-17 doc/paper/scripts/AnalyzeResultsDoubPend.py
fix formatting issue

@@ -54,7 +54,7 @@ for rank, i in enumerate(sorter[:how_many]):
     plt.figure()
     plt.title('RMSE vs Lyapunov Time\nParam {} ; RMSE {}'.format(
         results['params'][i], total_rmse))
-    t_adj = (ts_data.test_t - ts_data.test_t[0]) / lce[0]
+    t_adj = (ts_data.test_t - ts_data.test_t[0]) * lce[0]
     plt.plot(t_adj, rmse_over_t, 'o-')
     plt.plot(t_adj, np.zeros_like(t_adj))
     name = 'rank_{}_param_{}_rmse.png'.format(rank, i)
@@ -63,7 +63,7 @@ for rank, i in enumerate(sorter[:how_many]):
 
     runt['results'].t = ts_data.test_t
     runt['results'].y = ys_test.T
-    slvr.plotnd(runt)
+    slvr.plotnd(runt, dims=['θ1', 'θ2', 'ω1', 'ω2'])
     name = 'rank_{}_param_{}_fit.png'.format(rank, i)
     plt.savefig(os.path.join(
         dir_pre, name))

### 2019-04-17 doc/paper/scripts/AnalyzeResultsLorenz.py

@@ -54,7 +54,7 @@ for rank, i in enumerate(sorter[:how_many]):
     plt.figure()
     plt.title('RMSE vs Lyapunov Time\nParam {} ; RMSE {}'.format(
         results['params'][i], total_rmse))
-    t_adj = (ts_data.test_t - ts_data.test_t[0]) / lce[0]
+    t_adj = (ts_data.test_t - ts_data.test_t[0]) * lce[0]
     plt.plot(t_adj, rmse_over_t, 'o-')
     plt.plot(t_adj, np.zeros_like(t_adj))
     name = 'rank_{}_param_{}_rmse.png'.format(rank, i)

### 2019-04-17 doc/paper/paper.tex

@@
 
+To integrate the model, I use the initial conditions:
+$(x_0, y_0, z_0) = (-13.78990503, -11.6069343, 36.17332518)$. The calculated Lyapunov value was $0.9348566$.
+
 One of the issues run into while forecasting with the Lorenz is that the
@@
 the ESN, and due to resource limitations, I may have not been able to find
-the correct parameters to achieve a more perfect performance.
+the correct parameters to achieve a better performance.
 
@@
 \subsubsection{Double Pendulum}
+
 \subsubsection{Restricted Circular 3 Body}

### 2019-04-17 doc/paper/paper.tex

@@
 
+The double pendulum's initial condition was
+
+\begin{equation}
+(\theta_1, \theta_2, \omega_1, \omega_2) 
+= (4.66648088, 4.29820501,
+8.01585663 \times 10^{-4}, 3.71175422 \times 10^{-3}).
+\end{equation}
+In this case, we
+need $\theta_1$ large enough that the system is chaotic, but the rest of
+the parameters were chosen randomly (with rotational velocities close to
+zero). The resulting Lyapunov Exponent is $0.60074$. 
+Like the Lorenz, the differences between time points
+were used to train the ESN since the ESN was showing poor results
+otherwise. As the pendulum kept looping the center, $\theta_1$ and 
+$\theta_2$ kept growing; this shouldn't be an issue, but using
+the differences keeps the system variables in a closer range.
+
+Figure \ref{fig:doub_pend_diff} shows a plot of each
+of the dataset's dimensions. It seems like the dataset's activity is
+fairly sparse. There is only gradual but seemingly random variations until
+a few spots around $t=10, 30, 60, \text{and} 75$.
+The results for the ESN weren't very strong, but it was able to predict
+almost two Lyapunov Times before diverging.
+
+Because there are portions of the dataset which are calm, the ESN then has
+to pickup on signals which lead to the spikes in movement.
+We see the testing graph that while it predicts the testing set well,
+it eventually tries to predict spikes but does so at the wrong time,
+and the spike pattern never ends, has too large of magnitude, and is very 
+rapid. Overall the model performs poorly, and it's reflected in the RMSE.
+The likely reason for this is that the ESN can't learn the complex
+patterns from such a small and sparse dataset.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.5\linewidth]{doc/paper/images/doub_pend/full_differential.png}
+    \caption{The entire difference dataset}
+    \label{fig:doub_pend_diff}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.5\linewidth]{doc/paper/images/doub_pend/test_data.png}
+    \caption{The testing dataset partition}
+    \label{fig:doub_pend_testing}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/doub_pend/rank_0_param_77_fit.png}
+        \caption{Delayed spike prediction}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_0_param_225_rmse.png}
+        \caption{RMSE Over Time}
+    \end{subfigure}
+    \caption{Poor forecast of double pendulum}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/doub_pend/rank_1_param_63_fit.png}
+        \caption{Delayed spike prediction}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_1_param_230_rmse.png}
+        \caption{RMSE Over Time}
+    \end{subfigure}
+    \caption{Poor forecast of double pendulum}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/doub_pend/rank_2_param_305_fit.png}
+        \caption{Delayed spike prediction}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_2_param_458_rmse.png}
+        \caption{RMSE Over Time}
+    \end{subfigure}
+    \caption{Poor forecast of double pendulum}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/doub_pend/rank_3_param_52_fit.png}
+        \caption{Delayed spike prediction}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_3_param_151_rmse.png}
+        \caption{RMSE Over Time}
+    \end{subfigure}
+    \caption{Poor forecast of double pendulum}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/doub_pend/rank_4_param_127_fit.png}
+        \caption{Delayed spike prediction}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_4_param_300_rmse.png}
+        \caption{RMSE Over Time}
+    \end{subfigure}
+    \caption{Poor forecast of double pendulum}
+\end{figure}
+
 \subsubsection{Restricted Circular 3 Body}

### 2019-04-17 doc/paper/paper.tex
line formattings in paper.tex

--- 2019-04-17 doc/paper/paper.tex
@@
 
-Neural Networks are known to be able to approximate any function given 
-enough data and time to train \cite{cybenko1989approximation}.
-In light of this, the hope is that a deep learning based model,
-the ESN, will outperform ARIMA.
+Neural Networks are known to be able to approximate any function given enough
+data and time to train \cite{cybenko1989approximation}. In light of this, the
+hope is that a deep learning based model, the ESN, will outperform ARIMA.
 
@@
 model's \cite{oestreicher2007history}.
-
 One of the defining characteristics of chaos is that the error from
@@
 \subsection{Calculating the Largest Lyapunov Characteristic Exponent}
-
 \label{sec:calc_lya}
+
 In order to find the Lyapunov time, we only need to solve for the largest
@@
 
-The dynamical system will be defined by $\der{\xv}=\fv(\xv)$ where $\fv, \xv \in 
-\mathbb{R}^d$. Given an initial condition $\xv_0$, the largest LCE, denoted 
-$\lambda_1$, is calculated as follows:
+The dynamical system will be defined by $\der{\xv}=\fv(\xv)$ where $\fv, \xv
+\in \mathbb{R}^d$. Given an initial condition $\xv_0$, the largest LCE,
+denoted $\lambda_1$, is calculated as follows:
 
@@
 where $\sigma_1$ is the first singular value of the matrix $\frac{\partial
-\xv(t)}{\partial \xv_0}$ and $\log$ is the natural logarithm. Since we can only 
-numerically approximate the limit for $\lambda_1$, we will need to use a
+\xv(t)}{\partial \xv_0}$ and $\log$ is the natural logarithm. Since we can
+only numerically approximate the limit for $\lambda_1$, we will need to use a
 sufficiently large $t$ until the limit converges to the accuracy we need.
 
-In order to calculate $\frac{\partial \xv(t)}{\partial \xv_0}$, we need to integrate the matrix differential equation:
+In order to calculate $\frac{\partial \xv(t)}{\partial \xv_0}$, we need to
+integrate the matrix differential equation:
 
@@
 
-where $\Mv(t) \in \mathbb{R}^{d \times d}$ and $\frac{\partial \fv}{\partial \xv}$
-is the $d \times d$ Jacobian matrix of $\fv(\xv)$ at $\xv(t)$.
+where $\Mv(t) \in \mathbb{R}^{d \times d}$ and $\frac{\partial \fv}{\partial
+\xv}$ is the $d \times d$ Jacobian matrix of $\fv(\xv)$ at $\xv(t)$.
 
@@
 together. If the dynamical system has dimensions $d$, then the new system of
-differential equations will have dimension $d + d \times d$. For larger dynamical 
-systems, this calculation becomes costly. The dynamical systems analyzed are
-small enough (having dimensions ranging from 3-4) that the calculation can be 
-performed timely and accurately.
+differential equations will have dimension $d + d \times d$. For larger
+dynamical systems, this calculation becomes costly. The dynamical systems
+analyzed are small enough (having dimensions ranging from 3-4) that the
+calculation can be performed timely and accurately.
 
 There are several issues with calculating $\lambda_1$ with this algorithm.
-Already mentioned, we need to choose a sufficiently
-large $t$ for the limit to converge. 
-Unfortunately, for chaotic systems, $\Mv$ will increase rapidly with time. The 
-large floating point values will eventually overflow and $\lambda_1$ will not
-be accurate. I have found empirically that if values of $\Mv$ don't exceed 
-$10^{100}$, then $\lambda_1$ will be accurate.
-
-Implemented in \echonn, the algorithm was tested on the Lorenz system for various 
-parameters and initial conditions. The LCEs match the expected values very 
-accurately (see the Lorenz System section: \ref{sec:lorenz}).
-The 3 dimensional 3-body problem is an 18 dimensional dynamical system, so while 
-the dynamical system is implemented in \echonn, its max LCE can't be calculated 
-accurately since the resultant system has 342 dimensions. The algorithm was 
-also tested on the differential equation which produces a circle:
+Already mentioned, we need to choose a sufficiently large $t$ for the limit
+to converge. Unfortunately, for chaotic systems, $\Mv$ will increase rapidly
+with time. The large floating point values will eventually overflow and
+$\lambda_1$ will not be accurate. I have found empirically that if values of
+$\Mv$ don't exceed $10^{100}$, then $\lambda_1$ will be accurate.
+
+Implemented in \echonn, the algorithm was tested on the Lorenz system for
+various parameters and initial conditions. The LCEs match the expected values
+very accurately (see the Lorenz System section: \ref{sec:lorenz}). The 3
+dimensional 3-body problem is an 18 dimensional dynamical system, so while
+the dynamical system is implemented in \echonn, its max LCE can't be
+calculated accurately since the resultant system has 342 dimensions. The
+algorithm was also tested on the differential equation which produces a
+circle:
 
@@
 
-Given an initial condition $(x_0, y_0)$, it produces the trajectory of a circle
-with radius $r=\sqrt{x_0^2 + y_0^2}$. If perturbed, the radius will change, 
-$r_\epsilon=\sqrt{(x_0+\epsilon_x)^2 + (y_0+\epsilon_y)^2}$, so the new 
-trajectory is
+Given an initial condition $(x_0, y_0)$, it produces the trajectory of a
+circle with radius $r=\sqrt{x_0^2 + y_0^2}$. If perturbed, the radius will
+change, $r_\epsilon=\sqrt{(x_0+\epsilon_x)^2 + (y_0+\epsilon_y)^2}$, so the
+new trajectory is
 
@@
 
-The error doesn't grow at all over time, so the LCE should be less than or equal to zero as the system is not chaotic. The algorithm reports an LCE of $-1.15954 \times 10^{-06}$ as expected.
+The error doesn't grow at all over time, so the LCE should be less than or
+equal to zero as the system is not chaotic. The algorithm reports an LCE of
+$-1.15954 \times 10^{-06}$ as expected.
 
@@
 dynamical systems, and special precaution has to be taken when choosing a
-time interval. Using a larger time will yield a more accurate $\lambda_1$, but 
-it may also cause an overflow if the system is chaotic. Thus, 
-maximum time values need to be chosen such that $\vert \vert \Mv\vert 
-\vert_{\infty}$ doesn't exceed $10^{100}$ (where $\vert \vert \cdot \vert 
-\vert_{\infty}$ is the $\ell^\infty$ norm).
+time interval. Using a larger time will yield a more accurate $\lambda_1$,
+but it may also cause an overflow if the system is chaotic. Thus, maximum
+time values need to be chosen such that $\vert \vert \Mv\vert \vert_{\infty}$
+doesn't exceed $10^{100}$ (where $\vert \vert \cdot \vert \vert_{\infty}$ is
+the $\ell^\infty$ norm).
 
@@
 In order to generate the data to be analyized, the solutions to these
-dynamical systems need to be computed. Each of the systems are defined by 
-a system of differential equations which can be numerically solved. 
-However, due to the chaotic nature, error introduced by the numerical 
-solution will grow exponentially, so precaution need to be taken to avoid 
-having this sort of error introduced into the collected data. When 
-numerically solving chaotic problems, it's possible for the chaotic 
-behavior to be suppressed, or where there shouldn't be chaotic behavior,
-it could be created \cite{corless1994good}.
+dynamical systems need to be computed. Each of the systems are defined by a
+system of differential equations which can be numerically solved. However,
+due to the chaotic nature, error introduced by the numerical solution will
+grow exponentially, so precaution need to be taken to avoid having this sort
+of error introduced into the collected data. When numerically solving chaotic
+problems, it's possible for the chaotic behavior to be suppressed, or where
+there shouldn't be chaotic behavior, it could be created
+\cite{corless1994good}.
 
 However, it's been shown that the Fourth Order Runge-Kutta Method (RK4)
-outperforms other numerical methods such as Euler's Method and the
-Midpoint Method for chaotic systems \cite{zidan2011effect}.
-Not only does RK4 perform better than other methods, it serves as a good
-method for producing accurate numerical solutions to choatic systems with 
-small enough time steps: the Lü chaotic system \cite{mehdi2017using} and
-the Zhou system \cite{roslan2013solving}, for example. One article points
-out that a larger concern 
-regarding these solutions is the accuracy of the initial conditions 
-measured from the physical problem which the mathematical model is based,
-but otherwise, they echo the strength of correctly implemented 
-numerical solutions \cite{corless1994good}. Further, my initial conditions
-are randomly generated, so this is not a concern.
-
-I will be using an improved version of RK4 which
-incorporates the fifth order Runge-Kutta: RK4(5)
-\cite{dormand1980family}. This is the method implemented in
-the Python function \texttt{scipy.integrate.solve\_ivp}, and my
-numerical solver
-wraps this function to integrate the chaotic systems. Using small time
+outperforms other numerical methods such as Euler's Method and the Midpoint
+Method for chaotic systems \cite{zidan2011effect}. Not only does RK4 perform
+better than other methods, it serves as a good method for producing accurate
+numerical solutions to choatic systems with small enough time steps: the Lü
+chaotic system \cite{mehdi2017using} and the Zhou system
+\cite{roslan2013solving}, for example. One article points out that a larger
+concern regarding these solutions is the accuracy of the initial conditions
+measured from the physical problem which the mathematical model is based, but
+otherwise, they echo the strength of correctly implemented numerical
+solutions \cite{corless1994good}. Further, my initial conditions are randomly
+generated, so this is not a concern.
+
+I will be using an improved version of RK4 which incorporates the fifth order
+Runge-Kutta: RK4(5) \cite{dormand1980family}. This is the method implemented
+in the Python function \texttt{scipy.integrate.solve\_ivp}, and my numerical
+solver wraps this function to integrate the chaotic systems. Using small time
 steps will yield accurate samples from the chaotic systems.
@@
 shown in Table \ref{table:lorenz_params} are the values calculated by my
-Python module \echonn are listed. These values are not as accurate,
-but demonstrate the sufficient accuracy of \echonn.
+Python module \echonn are listed. These values are not as accurate, but
+demonstrate the sufficient accuracy of \echonn.
 
@@
 The Lorenz system is named after its creator, but it is also known as the
-Lorenz butterfly due to the shape of the system. The Lorenz
-Butterfly can be seen in Figure \ref{fig:lorenz_butter}.
+Lorenz butterfly due to the shape of the system. The Lorenz Butterfly can be
+seen in Figure \ref{fig:lorenz_butter}.
 
@@
 
-Although the notation in both papers was different, the equations were the same. 
-It's also worth mentioning that Eberle's paper was for his Masters Thesis, and 
-Frnka's paper has no details associated with it, but I was able to find a 
-mathematician with the exact name who achieved his doctorate. With that said, 
-I believe the math is reliable, and the simulations seem to work and at least 
-demonstrate chaotic properties (the only requirement for this project). With my
-lacking physics knowledge, I was unable to derive these equations myself from
-the assumptions made by Henri Poincaré outlined in 
+Although the notation in both papers was different, the equations were the
+same. It's also worth mentioning that Eberle's paper was for his Masters
+Thesis, and Frnka's paper has no details associated with it, but I was able
+to find a mathematician with the exact name who achieved his doctorate. With
+that said, I believe the math is reliable, and the simulations seem to work
+and at least demonstrate chaotic properties (the only requirement for this
+project). With my lacking physics knowledge, I was unable to derive these
+equations myself from the assumptions made by Henri Poincaré outlined in
 \cite{chenciner2000remarkable}.
@@
 the other two, but not in the standard Cartesian coordinates of 3D space.
-Outlined in the both papers, \cite{eberle2007case} and \cite{frnkacase}, the transformation back to
-Cartesian coordinates is given, but we don't need the exact position of the
-bodies since we are only concerned with the chaotic motion, so that isn't
-shown here.
+Outlined in the both papers, \cite{eberle2007case} and \cite{frnkacase}, the
+transformation back to Cartesian coordinates is given, but we don't need the
+exact position of the bodies since we are only concerned with the chaotic
+motion, so that isn't shown here.
 
@@
 (fast initial velocity away from the bodies). The simple divergent spiral
-isn't chaotic, so instead orbital and collision less initial conditions
-will be used. Without divergence and without collisions, the symplectic
-integrator is unnecessary.
+isn't chaotic, so instead orbital and collision less initial conditions will
+be used. Without divergence and without collisions, the symplectic integrator
+is unnecessary.
 
@@
 
-The Echo State Network (ESN) is a type of Recurrent Neural Network (RNN)
-used for time series analysis, but it is trained differently than most
-RNNs and is 
+The Echo State Network (ESN) is a type of Recurrent Neural Network (RNN) used
+for time series analysis, but it is trained differently than most RNNs and is
 characterized by having the echo state property. I implement the design
-detailed by Jaeger in is 2002 book \cite{jaeger2002tutorial} and
-further detailed in his 2007 article \cite{jaeger2007echo}.
-
-A time series data set will contain samples of data which change
-over time. Part of the data at each time point is measured or known
-before, called input data and denoted by $\uv(t)$ with $K$ dimensions.
-The rest of the data will be the goal for prediction, called output data and denoted $\dv(t)$ with $L$ dimensions.
-
-For $T$ time samples, $t_1, t_2, \dots, t_T$, we have a $T \times K$ 
-matrix, $\Uv$, with sample input vectors as rows. We also
-have a $T \times L$ matrix,
-$\Dv$, with sample output vectors as rows. A model will predict the 
-values of $\dv$, and this prediction will be denoted $\yv$.
+detailed by Jaeger in is 2002 book \cite{jaeger2002tutorial} and further
+detailed in his 2007 article \cite{jaeger2007echo}.
+
+A time series data set will contain samples of data which change over time.
+Part of the data at each time point is measured or known before, called input
+data and denoted by $\uv(t)$ with $K$ dimensions. The rest of the data will
+be the goal for prediction, called output data and denoted $\dv(t)$ with $L$
+dimensions.
+
+For $T$ time samples, $t_1, t_2, \dots, t_T$, we have a $T \times K$ matrix,
+$\Uv$, with sample input vectors as rows. We also have a $T \times L$ matrix,
+$\Dv$, with sample output vectors as rows. A model will predict the values of
+$\dv$, and this prediction will be denoted $\yv$.
 
@@
 \label{sec:rnn}
-RNNs are usually designed by having $K$ input units, $L$ output units, 
-and some internal units organized by some network structure. These units
-have evolving values for every time point. The
-structure of networks describes how units are connected. 
-The input units at time $t_i$ are set to $\uv(t_i)$. However, output
-units and
-internal units are calculated from the values of other units. These other 
-units are said to be connected to the resultant unit. Output
-units are calculated from current input units, current internal units,
-and from previous output units. Internal units 
-are calculated from the values of the current input units, 
-previous internal units, and previous output units.
-
-Although, it's rare that the output units are calculated from previous 
-output units \cite{jaeger2002tutorial}. Sometimes for training, the output 
-units are set to the value of $dv$, called teacher forcing, so that the 
-model trains on actual values as opposed to its accumulated errors 
-\cite{jaeger2002tutorial}.
+
+RNNs are usually designed by having $K$ input units, $L$ output units, and
+some internal units organized by some network structure. These units have
+evolving values for every time point. The structure of networks describes how
+units are connected. The input units at time $t_i$ are set to $\uv(t_i)$.
+However, output units and internal units are calculated from the values of
+other units. These other units are said to be connected to the resultant
+unit. Output units are calculated from current input units, current internal
+units, and from previous output units. Internal units are calculated from the
+values of the current input units, previous internal units, and previous
+output units.
+
+Although, it's rare that the output units are calculated from previous output
+units \cite{jaeger2002tutorial}. Sometimes for training, the output units are
+set to the value of $dv$, called teacher forcing, so that the model trains on
+actual values as opposed to its accumulated errors \cite{jaeger2002tutorial}.
 
 To calculate the value of a unit, an activation function is applied to a
-linear combination of the units connected to it \cite{svozil1997introduction}. The constants for the 
-linear combination are called weights, and they are parameters of the 
-model. Which activation function is used is also a parameter of the 
-function. Typically, some non-linear function is used, such as $\tanh$, 
-sigmoid, or rectified linear unit.
-
-During training, we search for
-weights which maximize the model's performance on data
-which the model hasn't seen. A common training method for RNNs
-is to use a form of backpropagation \cite{jaeger2002tutorial}. 
-Backpropagation for RNNs is slightly different from traditional Neural Networks, but the idea is the same.
-
-For example, given some neural network model $\yv(\xv; \wv)$, where
-$\xv$ is known input data, $\wv$ is the model's parameters, and $\yv$ 
-approximates some values, $\tv$, then we define an objective 
-function as the sum of squared error:
+linear combination of the units connected to it
+\cite{svozil1997introduction}. The constants for the linear combination are
+called weights, and they are parameters of the model. Which activation
+function is used is also a parameter of the function. Typically, some
+non-linear function is used, such as $\tanh$, sigmoid, or rectified linear
+unit.
+
+During training, we search for weights which maximize the model's performance
+on data which the model hasn't seen. A common training method for RNNs is to
+use a form of backpropagation \cite{jaeger2002tutorial}. Backpropagation for
+RNNs is slightly different from traditional Neural Networks, but the idea is
+the same.
+
+For example, given some neural network model $\yv(\xv; \wv)$, where $\xv$ is
+known input data, $\wv$ is the model's parameters, and $\yv$ approximates
+some values, $\tv$, then we define an objective function as the sum of
+squared error:
 
@@
 Using steepest descent to solve this minimization problem (as shown in
-\cite{svozil1997introduction}), we can
-reduce $\Ev$ in (\ref{eq:error_eq_rnn}) by
-updating the weights as follows:
+\cite{svozil1997introduction}), we can reduce $\Ev$ in
+(\ref{eq:error_eq_rnn}) by updating the weights as follows:
 
@@
 The implementation design used for this paper is detailed by Herbert Jaeger
-in \cite{jaeger2002tutorial} and \cite{jaeger2007echo}. I explain it here 
-to clarify my exact process (and with a slight change of notation to be 
+in \cite{jaeger2002tutorial} and \cite{jaeger2007echo}. I explain it here to
+clarify my exact process (and with a slight change of notation to be
 consistent with the rest of this paper).
 
-Like RNNs, ESNs have an input layer with $K$ units and with values 
-$\uv(t_1), \uv(t_2), \dots,  \uv(t_T)$ where $\uv(t) \in \mathbb{R}^K$.
-$K$ is zero if there are no input units. There is an internal layer
-with $N$ units, denoted by $\xv(t_i) \in \mathbb{R}^N$ for
-$i = 1, 2, \dots, T$, and 
-there is an output 
-layer with $L$ units, denoted by $\yv(t_i) \in \mathbb{R}^L$ for all
+Like RNNs, ESNs have an input layer with $K$ units and with values $\uv(t_1),
+\uv(t_2), \dots, \uv(t_T)$ where $\uv(t) \in \mathbb{R}^K$. $K$ is zero if
+there are no input units. There is an internal layer with $N$ units, denoted
+by $\xv(t_i) \in \mathbb{R}^N$ for $i = 1, 2, \dots, T$, and there is an
+output layer with $L$ units, denoted by $\yv(t_i) \in \mathbb{R}^L$ for all
 $i = 1, 2, \dots, T$.
@@
 input layer, the previous internal layer, and the previous output layer.
-These connections will have weights described in \ref{sec:rnn}. 
-Respectively, these weights will be denoted by the matrices:
-$\Win$, $\W$, and $\Wback$. The activation function for the internal
-layer is $\tanh$.
-
-The output layer is connected to the current input layer and current 
-internal layer. Additionally, a bias input is connected to this layer.
-The weight for these connections will be denoted $\Wout$ 
-(we only have one matrix since the connected layers are concatenated into a
-single vector, per Jaeger's notation \cite{jaeger2002tutorial}). The 
-activation function for these connections is linear. Although, the 
-recommended choice
-is $\tanh$, the output layer doesn't necessarily lie in the range of 
-$\tanh$, so it can't be used without some form of normalization.
+These connections will have weights described in \ref{sec:rnn}. Respectively,
+these weights will be denoted by the matrices: $\Win$, $\W$, and $\Wback$.
+The activation function for the internal layer is $\tanh$.
+
+The output layer is connected to the current input layer and current internal
+layer. Additionally, a bias input is connected to this layer. The weight for
+these connections will be denoted $\Wout$ (we only have one matrix since the
+connected layers are concatenated into a single vector, per Jaeger's notation
+\cite{jaeger2002tutorial}). The activation function for these connections is
+linear. Although, the recommended choice is $\tanh$, the output layer doesn't
+necessarily lie in the range of $\tanh$, so it can't be used without some
+form of normalization.
 
@@
 
-$\xv$ and $\yv$ are calculated from previous unit values but for the 
-first time step, there are none, so we need to initialize the
-units to zeros \cite{jaeger2002tutorial}. 
-This is done by defining $\xv(t_0)$ and $\yv(t_0)$ to zeros.
+$\xv$ and $\yv$ are calculated from previous unit values but for the first
+time step, there are none, so we need to initialize the units to zeros
+\cite{jaeger2002tutorial}. This is done by defining $\xv(t_0)$ and $\yv(t_0)$
+to zeros.
 
 The weights are initialized such that the model has the defining
-characteristic of echo state networks: the echo state property
-(Section \ref{sec:eprop}).
-During the training phase (Section \ref{sec:etrain}), $\Wout$ is fit to
-the training data.
-
-Once the network has been initialized and trained, it can be used to 
-perform time series analysis.
+characteristic of echo state networks: the echo state property (Section
+\ref{sec:eprop}). During the training phase (Section \ref{sec:etrain}),
+$\Wout$ is fit to the training data.
+
+Once the network has been initialized and trained, it can be used to perform
+time series analysis.
 
@@
 
-The echo state property is defined by the following \cite{jaeger2002tutorial}:
+The echo state property is defined by the following
+\cite{jaeger2002tutorial}:
 
@@
     $\dv(n)$ from compact intervals $\Uv$ and $\Dv$. The network ($\Win$,
-    $\W$, and $\Wback$) has echo states w.r.t. $\Uv$ and $\Dv$, if for 
-    every left-infinite input/output sequence $(\uv(n), \dv(n-1))$, where 
-    $n=\dots, -2, -1, 0$, and for all state sequences $\xv(n)$,
-    $\xv'(n)$ compatible with the teacher sequence, i.e. with
+    $\W$, and $\Wback$) has echo states w.r.t. $\Uv$ and $\Dv$, if for every
+    left-infinite input/output sequence $(\uv(n), \dv(n-1))$, where $n=\dots,
+    -2, -1, 0$, and for all state sequences $\xv(n)$, $\xv'(n)$ compatible
+    with the teacher sequence, i.e. with
     \begin{align}
@@
 networks with differing internal states are run with the same input and
-teacher-forced output, they will eventually converge to each other.
-It's name suggests data fed to the model will echo around but eventually 
-fade as echos do.
-
-Jaeger outlines a process to create weights which will likely have
-echo state properties. The process is posed as conjecture and 
-doesn't guarantee the echo state property. However, the echo state is very 
-likely and can be
-assumed confidently \cite{jaeger2002tutorial} \cite{jaeger2007echo}.
+teacher-forced output, they will eventually converge to each other. It's name
+suggests data fed to the model will echo around but eventually fade as echos
+do.
+
+Jaeger outlines a process to create weights which will likely have echo state
+properties. The process is posed as conjecture and doesn't guarantee the echo
+state property. However, the echo state is very likely and can be assumed
+confidently \cite{jaeger2002tutorial} \cite{jaeger2007echo}.
 
@@
 
-The process to initialize the echo state weights, $\W$, is as follows \cite{jaeger2002tutorial}:
+The process to initialize the echo state weights, $\W$, is as follows
+\cite{jaeger2002tutorial}:
 
@@
 
-If the spectral radius of $\W$ is found to be greater than 1, then it can 
-be shown that the model doesn't demonstrate the echo state property; 
-Jaeger's process uses this fact by ensuring that this isn't the case 
+If the spectral radius of $\W$ is found to be greater than 1, then it can be
+shown that the model doesn't demonstrate the echo state property; Jaeger's
+process uses this fact by ensuring that this isn't the case
 \cite{jaeger2002tutorial}. While it can't guarantee an echo state property,
@@
 
-Applying linear algebra, this makes a lot of sense. Let $A$ be an
-$N \times N$ matrix with spectral radius $r = \rho(A) > 0$ and denote its
-eigenvalues $\lambda_i$ and eigenvectors $\vv_i$ for
-$i = 1, 2, \dots, N$.  Then, define $B = \frac{\alpha}{r} A$.
+Applying linear algebra, this makes a lot of sense. Let $A$ be an $N \times
+N$ matrix with spectral radius $r = \rho(A) > 0$ and denote its eigenvalues
+$\lambda_i$ and eigenvectors $\vv_i$ for $i = 1, 2, \dots, N$. Then, define
+$B = \frac{\alpha}{r} A$.
 
@@
 
-where $\lambda$ is a vector of eigvenvalues and
-$\vert \vert \cdot \vert \vert_\text{max}$ is the $\ell^\infty$ norm.
+where $\lambda$ is a vector of eigvenvalues and $\vert \vert \cdot \vert
+\vert_\text{max}$ is the $\ell^\infty$ norm.
 
 As the model iterates, the matrix transformation of $B$ is repeatedly
-applied. With $\uv$ and teacher-forced $\dv$ as zeros, the internal state
-has the value $\xv(t_{k+1}) = \tanh(B \xv_k)$ for iteration $k$.
+applied. With $\uv$ and teacher-forced $\dv$ as zeros, the internal state has
+the value $\xv(t_{k+1}) = \tanh(B \xv_k)$ for iteration $k$.
 
@@
 
-Different from most training techniques, $\Win$, $\W$, and $\Wback$ are
-never changed after initialization.
+Different from most training techniques, $\Win$, $\W$, and $\Wback$ are never
+changed after initialization.
 
@@
 
-In order to train the echo state network, we need to find good 
-weight values for $\Wout$. Training the echo state network is different
-from the backpropagation approach in that we can solve for
-and calculate directly the weights matrix; more modern methods use
-an iterative approach which converges to good weight values. Part of the
-reason is due to the calculation cost of the iterative methods. Today, that
-it is no longer an issue, but historically, it was a strong quality of
-the echo state network \cite{jaeger2007echo}.
-
-First, the data (input, $\uv$, and output, $\dv$, values) is partitioned into three groups:
+In order to train the echo state network, we need to find good weight values
+for $\Wout$. Training the echo state network is different from the
+backpropagation approach in that we can solve for and calculate directly the
+weights matrix; more modern methods use an iterative approach which converges
+to good weight values. Part of the reason is due to the calculation cost of
+the iterative methods. Today, that it is no longer an issue, but
+historically, it was a strong quality of the echo state network
+\cite{jaeger2007echo}.
+
+First, the data (input, $\uv$, and output, $\dv$, values) is partitioned into
+three groups:
 \begin{enumerate}
@@
 inner state of the model, and the initial state (zeros) is lost due to the
-echo state property. $T_0$ should be chosen such that this is truly the 
-case. Like $\alpha$, $T_0$ is a parameter of the echo state network and
-differs depending on the problem. Good values for $T_0$ are 10 to 500 
-depending on the problem complexity.
-
-Third, the training values are fed into the model to sample the network.
-We can store the input unit values, the internal state values, and the 
-desired output values as matrices $M$ and $T$. $M$ will store the input 
-unit values and internal state value pairs
-$(\uv(t_{T_0}), 1, \xv(t_{T_0})), \dots, (\uv(t_T), 1, \xv(t_T))$
-in its rows. $T$ will store the output unit values
+echo state property. $T_0$ should be chosen such that this is truly the case.
+Like $\alpha$, $T_0$ is a parameter of the echo state network and differs
+depending on the problem. Good values for $T_0$ are 10 to 500 depending on
+the problem complexity.
+
+Third, the training values are fed into the model to sample the network. We
+can store the input unit values, the internal state values, and the desired
+output values as matrices $M$ and $T$. $M$ will store the input unit values
+and internal state value pairs $(\uv(t_{T_0}), 1, \xv(t_{T_0})), \dots,
+(\uv(t_T), 1, \xv(t_T))$ in its rows. $T$ will store the output unit values
 $\dv(t_{T_0}), \dots, \dv(t_T)$.
 
-To train the model, we want $\Wout M$ to approximate accurately
-$g^{-1}(T)$, where $g$ is the activation function for the output layer.
-Since we use an idempotent output layer activation function,
-$g^{-1}(T) = T$.
+To train the model, we want $\Wout M$ to approximate accurately $g^{-1}(T)$,
+where $g$ is the activation function for the output layer. Since we use an
+idempotent output layer activation function, $g^{-1}(T) = T$.
 
 Therefore, $\Wout$ should be found such that $\Wout M = T$
-\cite{jaeger2002tutorial}.
-Solving for $\Wout$,
+\cite{jaeger2002tutorial}. Solving for $\Wout$,
 
@@
 
-Using packages like \texttt{numpy.linalg.eig} and 
-\texttt{numpy.linalg.pinv}, solving for eigenvalues and
-pseudoinverses for $\Wout$ is very 
-easy. After solving for $\Wout$, the echo state network is trained.
-
-To evaluate the network, the testing values are fed into the model. 
-However, this time, the output layer is not teacher-forced since 
-in real application, the output values are unknown. 
-To measure how well the echo state network approximates the desired output
-we define the error to be the root mean squared error (RMSE) of the
-actual output values $\dv$ and
-predicted values $\yv$: $\sqrt{\sum_{i=T+1}^{n} (\dv(t_i) - \yv(t_i))^2}$ 
-where $n \in [T+1 \dots T_f]$. The error represents the predictive power
-of the model for $n$ time steps in the future. If RMSE is low for larger 
-values of $n$, the model has a strong long term predictive power.
+Using packages like \texttt{numpy.linalg.eig} and \texttt{numpy.linalg.pinv},
+solving for eigenvalues and pseudoinverses for $\Wout$ is very easy. After
+solving for $\Wout$, the echo state network is trained.
+
+To evaluate the network, the testing values are fed into the model. However,
+this time, the output layer is not teacher-forced since in real application,
+the output values are unknown. To measure how well the echo state network
+approximates the desired output we define the error to be the root mean
+squared error (RMSE) of the actual output values $\dv$ and predicted values
+$\yv$: $\sqrt{\sum_{i=T+1}^{n} (\dv(t_i) - \yv(t_i))^2}$ where $n \in [T+1
+\dots T_f]$. The error represents the predictive power of the model for $n$
+time steps in the future. If RMSE is low for larger values of $n$, the model
+has a strong long term predictive power.
 
@@
 
-First, data from the models is collected. Using the process described
-above (Section \ref{sec:calc_lya}), the Largest Lyapunov Exponent is
-calculated, $\lambda_1$. Then, the Lyapunov Time is calculated by
-$\frac{1}{\lambda_1}$. We will then integrate the model with an algorithm
-called RK4(5) over the interval
-$[0, \frac{50}{\lambda_1}]$ with time steps of $\frac{1}{100 \lambda_1}$. 
-This will result in 5,000 data points.
-
-Then, the data is partitioned into a training set, a validation set, and 
-a testing set. I use a 90\% split, so testing is the last 10\% of the
-data. The training and validation set are the first 90\%. Of the 90\%,
-training is the first 90\%, and validation is the final 10\%. In total,
-it's a 81\%-9\%-10\% split.
-
-The training set is used to train the model. The validation
-set is used to tune the parameters of the model ($\alpha, N, T_0$). The
-testing set is used to evaluate the performance.
-
-The first data point of the training set is $t_0 = 0$. The parameter
-$T_0$ marks where the model begins training on the data (see Section
-\ref{sec:etrain}). Using slightly different notation to account for the
-extra validation set, we will denote the first validation time point by
-$t = T$. We will denote the first testing time point by $t = T_f$
+First, data from the models is collected. Using the process described above
+(Section \ref{sec:calc_lya}), the Largest Lyapunov Exponent is calculated,
+$\lambda_1$. Then, the Lyapunov Time is calculated by $\frac{1}{\lambda_1}$.
+We will then integrate the model with an algorithm called RK4(5) over the
+interval $[0, \frac{50}{\lambda_1}]$ with time steps of $\frac{1}{100
+\lambda_1}$. This will result in 5,000 data points.
+
+Then, the data is partitioned into a training set, a validation set, and a
+testing set. I use a 90\% split, so testing is the last 10\% of the data. The
+training and validation set are the first 90\%. Of the 90\%, training is the
+first 90\%, and validation is the final 10\%. In total, it's a 81\%-9\%-10\%
+split.
+
+The training set is used to train the model. The validation set is used to
+tune the parameters of the model ($\alpha, N, T_0$). The testing set is used
+to evaluate the performance.
+
+The first data point of the training set is $t_0 = 0$. The parameter $T_0$
+marks where the model begins training on the data (see Section
+\ref{sec:etrain}). Using slightly different notation to account for the extra
+validation set, we will denote the first validation time point by $t = T$. We
+will denote the first testing time point by $t = T_f$
 
@@
 
-where $(\cdot)^2$ is the dot product of the error with itself,
-$\dv(t)$ is the true value at time $t$,
-and $\yv(t)$ is the model's approximation at time $t$.
-
-First, we generate a list of parameter combinations to try. Usually, this
-is the Cartesian product of typical parameter values for the model. 
-For each combination, we train several models on the training data. Then,
-we predict the validation data to measure the general performance of the
-model under those parameters.
-
-$\text{RMSE}_\text{train}$ is calculated on the interval $[T_0, T)$,
-and the prediction uses teacher-forced values on $[0, T_0)$.
-$\text{RMSE}_\text{validation}$ is calculated on the interval $[T, T_f)$,
-and the prediction uses teacher-forced values on $[0, T)$.
+where $(\cdot)^2$ is the dot product of the error with itself, $\dv(t)$ is
+the true value at time $t$, and $\yv(t)$ is the model's approximation at time
+$t$.
+
+First, we generate a list of parameter combinations to try. Usually, this is
+the Cartesian product of typical parameter values for the model. For each
+combination, we train several models on the training data. Then, we predict
+the validation data to measure the general performance of the model under
+those parameters.
+
+$\text{RMSE}_\text{train}$ is calculated on the interval $[T_0, T)$, and the
+prediction uses teacher-forced values on $[0, T_0)$.
+$\text{RMSE}_\text{validation}$ is calculated on the interval $[T, T_f)$, and
+the prediction uses teacher-forced values on $[0, T)$.
 
 Then, the models which performed best on the validation set should also
-perform well on the testing set. The models are retrained on the testing
-and validation set, and the error is calculated:
-$\text{RMSE}_\text{full train}$ is calculated on the interval $[T, T_f)$,
-and the prediction uses teacher-forced values on $[0, T)$.
-
-Finally, the error for the testing set is calculated. 
-The performance on the testing set is 
-an unbias representation of how well the ESN can fit the model.
-
-However, the $\text{RMSE}_\text{test}$ is calculated slightly different
-since we want to know how far the model can predict before deviating too
-far. The $\text{RMSE}_\text{test}(t_f)$ will be calculated on the interval
-$[T_f, t_f)$ and the prediction uses teacher-forced values on $[0, T_f)$.
-Lastly, $\text{RMSE}_\text{test}$ will notate the RMSE for the entire test
-set.
-
-We will look at how far past $T_f$ the ESN model's
-$\text{RMSE}_\text{test}$ grows too large. The best model will be the 
-model with the lowest Testing RMSE for the longest time.
+perform well on the testing set. The models are retrained on the testing and
+validation set, and the error is calculated: $\text{RMSE}_\text{full train}$
+is calculated on the interval $[T, T_f)$, and the prediction uses
+teacher-forced values on $[0, T)$.
+
+Finally, the error for the testing set is calculated. The performance on the
+testing set is an unbias representation of how well the ESN can fit the
+model.
+
+However, the $\text{RMSE}_\text{test}$ is calculated slightly different since
+we want to know how far the model can predict before deviating too far. The
+$\text{RMSE}_\text{test}(t_f)$ will be calculated on the interval $[T_f,
+t_f)$ and the prediction uses teacher-forced values on $[0, T_f)$. Lastly,
+$\text{RMSE}_\text{test}$ will notate the RMSE for the entire test set.
+
+We will look at how far past $T_f$ the ESN model's $\text{RMSE}_\text{test}$
+grows too large. The best model will be the model with the lowest Testing
+RMSE for the longest time.
 
@@
 
-To integrate the model, I use the initial conditions:
-$(x_0, y_0, z_0) = (-13.78990503, -11.6069343, 36.17332518)$. The calculated Lyapunov value was $0.9348566$.
+To integrate the model, I use the initial conditions: $(x_0, y_0, z_0) =
+(-13.78990503, -11.6069343, 36.17332518)$. The calculated Lyapunov value was
+$0.9348566$.
 
 One of the issues run into while forecasting with the Lorenz is that the
-continuous nature of the equation led to the ESN just outputing the
-previous value. For initial values far from the center of the Lorenz
-Butterfly, it would tend towards the center and stop when it does arrive
-there. Most models trained poorly and gave $\text{RMSE}_\text{test}$
-values of 16 or well above 10,000, implying the model diverged.
-
-To fix this issue, we instead have the ESN train on differences between
-time points. This reduces our data set to 4,999 values, and it creates
-a new shape of the data. The Lorenz Butterfly had a similar shape, two rings, but in this case, the rings intersect (See figure \ref{fig:lorenz_diff}).
+continuous nature of the equation led to the ESN just outputing the previous
+value. For initial values far from the center of the Lorenz Butterfly, it
+would tend towards the center and stop when it does arrive there. Most models
+trained poorly and gave $\text{RMSE}_\text{test}$ values of 16 or well above
+10,000, implying the model diverged.
+
+To fix this issue, we instead have the ESN train on differences between time
+points. This reduces our data set to 4,999 values, and it creates a new shape
+of the data. The Lorenz Butterfly had a similar shape, two rings, but in this
+case, the rings intersect (See figure \ref{fig:lorenz_diff}).
 
@@
 After this transformation of the data, the model performed significantly
-better. $\text{RMSE}_\text{test}$ dropped to ranges of 1-2, the
-model exploded (diverged to infinity) less frequently, and it was no
-longer outputting previous values.
-
-Most well performing models couldn't predict past one Lyapunov constant. 
-From the graphs, we can see that most models deviated once reaching the
+better. $\text{RMSE}_\text{test}$ dropped to ranges of 1-2, the model
+exploded (diverged to infinity) less frequently, and it was no longer
+outputting previous values.
+
+Most well performing models couldn't predict past one Lyapunov constant. From
+the graphs, we can see that most models deviated once reaching the
 intersection of the butterfly wings.
 
-Once reaching the intersection,
-several things can happen. First, it can deviate completely and stop
-following either wings. Second, it can follow the wrong wing. Which wing
-the trajectory follows seems almost random, so it makes sense if it
-chooses the wrong one. Third, it will continue to follow the trajectory,
-but slightly perturbed. With every passing of the intersection, any of
-these can happen. Eventually, even the perturbations causes problems. 
-The model was trained on data that existed in a specific range of space. 
-When the perturbations lead the model into space not explored, 
-it performs even worse, and this can cause very sporastic behavior.
-
-The chaotic behavior of the Lorenz model dictates which wing the
-trajectory will follow. This is difficult for the ESN to be able to
-model. However, it is possible that I am not using enough data to train
-the ESN, and due to resource limitations, I may have not been able to find
-the correct parameters to achieve a better performance.
+Once reaching the intersection, several things can happen. First, it can
+deviate completely and stop following either wings. Second, it can follow the
+wrong wing. Which wing the trajectory follows seems almost random, so it
+makes sense if it chooses the wrong one. Third, it will continue to follow
+the trajectory, but slightly perturbed. With every passing of the
+intersection, any of these can happen. Eventually, even the perturbations
+causes problems. The model was trained on data that existed in a specific
+range of space. When the perturbations lead the model into space not
+explored, it performs even worse, and this can cause very sporastic behavior.
+
+The chaotic behavior of the Lorenz model dictates which wing the trajectory
+will follow. This is difficult for the ESN to be able to model. However, it
+is possible that I am not using enough data to train the ESN, and due to
+resource limitations, I may have not been able to find the correct parameters
+to achieve a better performance.
 

### 2019-04-17 doc/paper/paper.tex
formatting

@@
 \end{equation}
-In this case, we
-need $\theta_1$ large enough that the system is chaotic, but the rest of
-the parameters were chosen randomly (with rotational velocities close to
-zero). The resulting Lyapunov Exponent is $0.60074$. 
-Like the Lorenz, the differences between time points
-were used to train the ESN since the ESN was showing poor results
-otherwise. As the pendulum kept looping the center, $\theta_1$ and 
-$\theta_2$ kept growing; this shouldn't be an issue, but using
-the differences keeps the system variables in a closer range.
-
-Figure \ref{fig:doub_pend_diff} shows a plot of each
-of the dataset's dimensions. It seems like the dataset's activity is
-fairly sparse. There is only gradual but seemingly random variations until
-a few spots around $t=10, 30, 60, \text{and} 75$.
-The results for the ESN weren't very strong, but it was able to predict
-almost two Lyapunov Times before diverging.
-
-Because there are portions of the dataset which are calm, the ESN then has
-to pickup on signals which lead to the spikes in movement.
-We see the testing graph that while it predicts the testing set well,
-it eventually tries to predict spikes but does so at the wrong time,
-and the spike pattern never ends, has too large of magnitude, and is very 
-rapid. Overall the model performs poorly, and it's reflected in the RMSE.
-The likely reason for this is that the ESN can't learn the complex
-patterns from such a small and sparse dataset.
+
+In this case, we need $\theta_1$ large enough that the system is chaotic, but
+the rest of the parameters were chosen randomly (with rotational velocities
+close to zero). The resulting Lyapunov Exponent is $0.60074$. Like the
+Lorenz, the differences between time points were used to train the ESN since
+the ESN was showing poor results otherwise. As the pendulum kept looping the
+center, $\theta_1$ and $\theta_2$ kept growing; this shouldn't be an issue,
+but using the differences keeps the system variables in a closer range.
+
+Figure \ref{fig:doub_pend_diff} shows a plot of each of the dataset's
+dimensions. It seems like the dataset's activity is fairly sparse. There is
+only gradual but seemingly random variations until a few spots around $t=10,
+30, 60, \text{and} 75$. The results for the ESN weren't very strong, but it
+was able to predict almost two Lyapunov Times before diverging.
+
+Because there are portions of the dataset which are calm, the ESN then has to
+pickup on signals which lead to the spikes in movement. We see the testing
+graph that while it predicts the testing set well, it eventually tries to
+predict spikes but does so at the wrong time, and the spike pattern never
+ends, has too large of magnitude, and is very rapid. Overall the model
+performs poorly, and it's reflected in the RMSE. The likely reason for this
+is that the ESN can't learn the complex patterns from such a small and sparse
+dataset.
 

### 2019-04-18 src/echonn/ml/esn_experiment.py
random ic for 3bod searchable

@@ -57,7 +57,7 @@ class ESNExperiment:
         if self.verbose:
             print('Collecting data...')
         model = self.model
-        y0 = model.get_rnd_ic()
+        y0 = model.get_rnd_ic(verbose=self.verbose)
         model_solver = SystemSolver(model)
         lce_T = model.best_lce_T
         lce, lce_run = model_solver.get_lce(lce_T, y0)

### 2019-04-18 src/echonn/sys/restricted_circular_3_body.py

@@ -1,14 +1,61 @@
 import numpy as np
 from .system import DynamicalSystem
+from .system_solver import SystemSolver
 
 
 class RestrictedCircular3Body(DynamicalSystem):
-    def __init__(self, body_ratio=0.5, method='BDF'):
+    def __init__(self, body_ratio=0.5, method='RK45', search=False, max_tries=200, T=200, rmin=0.00000000001, rmax=10):
         super().__init__(4, method, 100)  # TODO: verify this
         if body_ratio < 0 or 1 < body_ratio:
             raise ValueError('body_ratio should be in range [0, 1]')
+        self.search = search
+        self.max_tries = max_tries
+        self.T = T
+        self.rmin = rmin
+        self.rmax = rmax
         self.mu = body_ratio
         self.alpha = 1 - body_ratio
+
+    def get_rnd_ic(self, verbose=False, **kwargs):
+        if not self.search:
+            return super().get_rnd_ic()
+        max_tries = self.max_tries
+        T = self.T
+        rmin = self.rmin
+        rmax = self.rmax
+        if verbose:
+            print('Finding good initial condition')
+        slvr = SystemSolver(self)
+        found = False
+        tries = 0
+        while not found:
+            if verbose:
+                print('tries', tries)
+            tries += 1
+            if tries > max_tries:
+                raise Exception('Too many failed tries')
+            y0 = np.random.rand(self.dim)
+            #y0[0] = 1
+            y0[1] = 0
+            y0[2] = 0
+            #y0[3] = 1
+            run = slvr.run([0, T], y0, **kwargs)
+            t = run['results'].t
+            y = run['results'].y
+            found = True
+            for i in range(y.shape[1]):
+                r1 = self.r1(t[i], y[:, i])
+                r2 = self.r2(t[i], y[:, i])
+                if r1 < rmin or r2 < rmin or r1 > rmax or r2 > rmax:
+                    progress = int((i+1) / y.shape[1] * 100)
+                    fstr = 'failed on {:5} / {:5} ({:3} %) with ({:.2}, {:.2})'
+                    if verbose:
+                        print(fstr.format(i, y.shape[1], progress, r1, r2))
+                    found = False
+                    break
+        if verbose:
+            print('soln', y0)
+        return y0
 
     def r1(self, t, v):
         mu = self.mu

### 2019-04-18 src/test/ml/test_experiment.py

@@ -9,7 +9,7 @@ import pickle
 class TestExperiment(unittest.TestCase):
     def setUp(self):
         self.file_io = True
-        self.verbose = True
+        self.verbose = False
         self.alpha = [.7, .8]
         self.N = [3]
         self.T0 = [500]
@@ -54,11 +54,14 @@ class TestExperiment(unittest.TestCase):
     def test3BodyProblem(self):
         if self.verbose:
             print()
-        fn = os.path.join('src', 'test', 'test_data', '3b.p')
+        fn = os.path.join('src', 'test', 'test_data', 'rc3_body_results.p')
         if self.verbose:
             print('Testing 3 Body Problem Model')
-        experiment = ESNExperiment(
-            RestrictedCircular3Body(), self.params, self.trials, 1000)
+        sys = RestrictedCircular3Body(body_ratio=0.5, search=True, max_tries=200,
+                                      T=500, rmin=0.00000000001, rmax=10)
+        experiment = ESNExperiment(sys, params=self.params, use_diff=False,
+                                   t_len=50, trials=self.trials,
+                                   time_steps_per_lce_time=100)
         res = experiment.run(self.verbose)
         if self.file_io:
             with open(fn, 'wb') as f:

### 2019-04-18 src/echonn/sys/system.py
small bug fix

@@ -13,7 +13,7 @@ class DynamicalSystem(ABC):
         return v
 
     def get_rnd_ic(self, **kwargs):
-        return np.random.rand(self.system.dim)
+        return np.random.rand(self.dim)
 
     def Dfun(self, t, v):
         ''' This ignores t '''

### 2019-04-18 src/echonn/ml/esn_experiment.py
enable specified y0

@@ -8,7 +8,8 @@ import matplotlib.pyplot as plt
 
 
 class ESNExperiment:
-    def __init__(self, model, data=None, params=None, use_diff=True, trials=2, t_len=50, time_steps_per_lce_time=100):
+    def __init__(self, model, y0=None, data=None, params=None, use_diff=True, trials=2, t_len=50, time_steps_per_lce_time=100):
+        self.y0 = y0
         self.model = model
         self.t_len = t_len
         self.data = data
@@ -57,7 +58,10 @@ class ESNExperiment:
         if self.verbose:
             print('Collecting data...')
         model = self.model
-        y0 = model.get_rnd_ic(verbose=self.verbose)
+        if self.y0 is None:
+            y0 = model.get_rnd_ic(verbose=self.verbose)
+        else:
+            y0 = self.y0
         model_solver = SystemSolver(model)
         lce_T = model.best_lce_T
         lce, lce_run = model_solver.get_lce(lce_T, y0)

### 2019-04-18 doc/paper/scripts/AnalyzeR3BodyProb.py
good y0 generation

@@ -1,13 +1,39 @@
 import os
 import pickle
+import itertools
 from echonn.sys import RestrictedCircular3Body
 from echonn.ml import ESNExperiment
 
 
-fn = os.path.join('.', 'rc_3_bod_results.p')
+fn = os.path.join('.', 'rc3_body_results.p')
+fn_y0s = os.path.join('.', 'rc3_body_y0s.p')
 print('Testing Restricted Circular 3 Body Model')
-experiment = ESNExperiment(RestrictedCircular3Body())
-res = experiment.run()
+alphas = [.7, .75, .8, .85, .9, .98]
+#alphas = [.7, .8]
+Ns = [1, 3, 5, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 150, 200, 300]
+#Ns = [3]
+T0s = [50, 100, 150, 200, 300, 500]
+#T0s = [30]
+params = list(itertools.product(alphas, Ns, T0s))
+if os.path.exists(fn_y0s):
+    with open(fn_y0s, 'rb') as f:
+        y0s = pickle.load(f)
+else:
+    y0s = []
+sys = RestrictedCircular3Body(search=True, body_ratio=0.5, max_tries=200,
+                              T=500, rmin=0.001, rmax=10)
+y0 = None
+y0 = y0s[-1]
+experiment = ESNExperiment(sys, y0=y0, params=params, use_diff=False,
+                           t_len=50, trials=3,
+                           time_steps_per_lce_time=100)
+res = experiment.run(verbose=True)
+y0 = res[0]['results'].y[:, 0]
+y0s.append(y0)
 with open(fn, 'wb') as f:
     pickle.dump(res, f)
+with open(fn_y0s, 'wb') as f:
+    pickle.dump(y0s, f)
+
 print(res)
+print('y0s', y0s)

### 2019-04-18 doc/paper/scripts/AnalyzeResultsR3Body.py

@@ -1,0 +1,79 @@
+import os
+import pickle
+import numpy as np
+import matplotlib.pyplot as plt
+from copy import deepcopy
+
+from echonn.sys import SystemSolver
+
+res = pickle.load(open('rc3_body_results.p', 'rb'))
+run, lce, ts_data, results = res
+
+print('lce:', lce[0])
+print('ic:', run['results'].y[:, 0])
+print('tf:', run['results'].t[-1])
+test_rmse = [rmse for _, _, _, (_, _, rmse) in results['best model rmse']]
+test_rmse = np.array(test_rmse)
+
+
+#fn_y0s = os.path.join('.', 'rc3_body_y0s.p')
+#y0s = [run['results'].y[:, 0]]
+# with open(fn_y0s, 'wb') as f:
+#pickle.dump(y0s, f)
+
+
+def rmse(d, y):
+    num_samples = d.shape[0]
+    return np.sqrt(np.sum((d-y)**2) / num_samples)
+
+
+score = []
+higher_score = []
+for rmse_res in results['best model rmse']:
+    ds_test, ys_test, total_rmse = rmse_res[3]
+    for sub in range(ds_test.shape[0]):
+        err = rmse(ds_test[:sub+1], ys_test[:sub+1])
+        if err > .05:
+            score.append(sub)
+            break
+dir_pre = os.path.join('..', 'images', 'r3body')
+sys = run['system']
+slvr = SystemSolver(sys)
+runt = deepcopy(run)
+runt['results'].y = run['results'].y[[0, 2], :]
+slvr.plot2d(runt)
+plt.scatter([-sys.alpha, sys.mu], [0, 0])
+plt.show(True)
+plt.savefig(os.path.join(dir_pre, 'full_differential.png'))
+
+runt['results'].t = ts_data.test_t
+runt['results'].y = ds_test[:, :2].T
+slvr.plot2d(runt)
+plt.savefig(os.path.join(dir_pre, 'test_data.png'))
+sorter = np.flip(np.argsort(score))
+how_many = 5
+for rank, i in enumerate(sorter[:how_many]):
+    ds_test, ys_test, total_rmse = results['best model rmse'][i][3]
+    print(i, total_rmse, results['params'][i])
+
+    rmse_over_t = [rmse(ds_test[:sub+1], ys_test[:sub+1])
+                   for sub in range(ds_test.shape[0])]
+
+    plt.figure()
[…]

### 2019-04-18 doc/paper/scripts/RenderR3Bod.py

@@ -11,7 +11,7 @@ if __name__ == "__main__":
     init = np.array(1000*np.random.rand(4), dtype=int)/1000
     slv = SystemSolver(sys)
     #run = slv.run([0, 10], init, max_step=0.001)
-    T = 50
+    T = 100
     lce, run = slv.get_lce(T=T, y0=init)
     #lce, run = slv.get_lce(T=T)
     mat = run['results'].y[4:, -1].reshape(4, 4)

### 2019-04-18 src/echonn/sys/restricted_circular_3_body.py

@@ -35,10 +35,10 @@ class RestrictedCircular3Body(DynamicalSystem):
             if tries > max_tries:
                 raise Exception('Too many failed tries')
             y0 = np.random.rand(self.dim)
-            #y0[0] = 1
+            y0[0] = .5 + y0[0]/2
             y0[1] = 0
             y0[2] = 0
-            #y0[3] = 1
+            y0[3] = .5 + y0[3]/2
             run = slvr.run([0, T], y0, **kwargs)
             t = run['results'].t
             y = run['results'].y

### 2019-04-18 src/echonn/ml/echo_state_network.py
enable split config in experiment

@@ -5,7 +5,7 @@ from .time_series_forecaster import TimeSeriesForecaster
 
 
 class EchoStateNetwork(TimeSeriesForecaster):
-    def __init__(self, K, N, L, T0=50, alpha=.8, noise=0.0, bias=True, f=None, g=None, g_inv=None):
+    def __init__(self, K, N, L, T0=50, alpha=.8, noise=0, bias=True, f=None, g=None, g_inv=None):
         """
         K - input units, u
         N - internal units, x (should not exceed Tf/10 to Tf/2 at risk of overfitting
@@ -16,7 +16,6 @@ class EchoStateNetwork(TimeSeriesForecaster):
         Specifies whether the algorithm should feed noise into the model
         during sampling. Uniform distribution ranging from [-noise, noise].
         """
-        # TODO: implement bias
         bias_K = K
         if bias:
             bias_K += 1
@@ -82,7 +81,7 @@ class EchoStateNetwork(TimeSeriesForecaster):
         x = self.x
         y = self.y
 
-        if self.noise is None:
+        if self.noise == 0:
             noise = 0
         else:
             noise = np.random.uniform(-self.noise, self.noise)

### 2019-04-18 src/echonn/ml/esn_experiment.py

@@ -8,11 +8,12 @@ import matplotlib.pyplot as plt
 
 
 class ESNExperiment:
-    def __init__(self, model, y0=None, data=None, params=None, use_diff=True, trials=2, t_len=50, time_steps_per_lce_time=100):
+    def __init__(self, model, y0=None, data=None, params=None, tsttrn_split=0.9, use_diff=True, trials=2, t_len=50, time_steps_per_lce_time=100):
         self.y0 = y0
         self.model = model
         self.t_len = t_len
         self.data = data
+        self.tsttrn_split = tsttrn_split
         self.use_diff = use_diff
         self.params = params
         self.trials = trials
@@ -87,7 +88,8 @@ class ESNExperiment:
             Use model with best performance, retrain on train and cv
             Run model on testing set and calculate RMSE for every time point
         """
-        ts_data = TSData(run=model_run, use_diff=self.use_diff)
+        ts_data = TSData(run=model_run, use_diff=self.use_diff,
+                         split=self.tsttrn_split)
         ts_analysis = TSAnalysis(ts_data, self.params, self.trials)
         results = ts_analysis.run(self.verbose)
         return ts_data, results

### 2019-04-18 src/test/ml/test_esn.py

@@ -145,6 +145,8 @@ class TestEchoStateNetwork(unittest.TestCase):
         return np.mean((x - y)**2)
 
     def runPaperExperiment(self, noise=0):
+        if noise is None:
+            noise = 0
         esn = EchoStateNetwork(0, 20, 1, T0=100, g=np.tanh,
                                g_inv=np.arctanh, alpha=.99, noise=noise)
         n = np.arange(1, 351)

### 2019-04-18 src/echonn/ml/echo_state_network.py
adjust for None noise

@@ -30,6 +30,8 @@ class EchoStateNetwork(TimeSeriesForecaster):
         self.L = L
         self.T0 = T0  # at T0, x[T0] has stabilized
         self.alpha = alpha
+        if noise is None:
+            noise = 0
         self.noise = noise
         self.w = None
         self.Ws = None

### 2019-04-18 doc/paper/scripts/AnalyzeR3BodyProb.py
change split to .7

@@ -25,7 +25,7 @@ sys = RestrictedCircular3Body(search=True, body_ratio=0.5, max_tries=200,
 y0 = None
 y0 = y0s[-1]
 experiment = ESNExperiment(sys, y0=y0, params=params, use_diff=False,
-                           t_len=50, trials=3,
+                           t_len=50, trials=3, tsttrn_split=.7,
                            time_steps_per_lce_time=100)
 res = experiment.run(verbose=True)
 y0 = res[0]['results'].y[:, 0]

### 2019-04-18 doc/paper/scripts/AnalyzeResultsR3Body.py

@@ -9,17 +9,14 @@ from echonn.sys import SystemSolver
 res = pickle.load(open('rc3_body_results.p', 'rb'))
 run, lce, ts_data, results = res
 
+# print small details
 print('lce:', lce[0])
 print('ic:', run['results'].y[:, 0])
 print('tf:', run['results'].t[-1])
-test_rmse = [rmse for _, _, _, (_, _, rmse) in results['best model rmse']]
-test_rmse = np.array(test_rmse)
 
-
-#fn_y0s = os.path.join('.', 'rc3_body_y0s.p')
-#y0s = [run['results'].y[:, 0]]
-# with open(fn_y0s, 'wb') as f:
-#pickle.dump(y0s, f)
+rank_metric = np.array([test_rmse for _, cv_rmse, _, (_, _, test_rmse)
+                        in results['best model rmse']])
+ranking = np.argsort(rank_metric)
 
 
 def rmse(d, y):
@@ -27,32 +24,29 @@ def rmse(d, y):
     return np.sqrt(np.sum((d-y)**2) / num_samples)
 
 
-score = []
-higher_score = []
-for rmse_res in results['best model rmse']:
-    ds_test, ys_test, total_rmse = rmse_res[3]
-    for sub in range(ds_test.shape[0]):
-        err = rmse(ds_test[:sub+1], ys_test[:sub+1])
-        if err > .05:
-            score.append(sub)
-            break
+# get img dir
 dir_pre = os.path.join('..', 'images', 'r3body')
+
+# extract system and build solver
 sys = run['system']
 slvr = SystemSolver(sys)
+
+# draw full data
 runt = deepcopy(run)
 runt['results'].y = run['results'].y[[0, 2], :]
 slvr.plot2d(runt)
 plt.scatter([-sys.alpha, sys.mu], [0, 0])
-plt.show(True)
 plt.savefig(os.path.join(dir_pre, 'full_differential.png'))
 
+# draw test data
 runt['results'].t = ts_data.test_t
-runt['results'].y = ds_test[:, :2].T
+runt['results'].y = ts_data.test_y[:, [0, 2]].T
 slvr.plot2d(runt)
+plt.scatter([-sys.alpha, sys.mu], [0, 0])
 plt.savefig(os.path.join(dir_pre, 'test_data.png'))
-sorter = np.flip(np.argsort(score))
-how_many = 5
-for rank, i in enumerate(sorter[:how_many]):
+
[…]

### 2019-04-18 doc/paper/scripts/AnalyzeResultsR3Body.py
add 3bod results

@@ -14,9 +14,11 @@ print('lce:', lce[0])
 print('ic:', run['results'].y[:, 0])
 print('tf:', run['results'].t[-1])
 
-rank_metric = np.array([test_rmse for _, cv_rmse, _, (_, _, test_rmse)
-                        in results['best model rmse']])
-ranking = np.argsort(rank_metric)
+rank_metric_cv = np.array([cv_rmse for _, cv_rmse, _, (_, _, test_rmse)
+                           in results['best model rmse']])
+rank_metric_test = np.array([test_rmse for _, cv_rmse, _, (_, _, test_rmse)
+                             in results['best model rmse']])
+ranking = np.argsort(rank_metric_cv)
 
 
 def rmse(d, y):
@@ -45,7 +47,7 @@ slvr.plot2d(runt)
 plt.scatter([-sys.alpha, sys.mu], [0, 0])
 plt.savefig(os.path.join(dir_pre, 'test_data.png'))
 
-how_many = 10
+how_many = 5
 for rank, i in enumerate(ranking[:how_many]):
     ds_test, ys_test, total_rmse = results['best model rmse'][i][3]
     print(i, total_rmse, results['params'][i])
@@ -54,8 +56,7 @@ for rank, i in enumerate(ranking[:how_many]):
                    for sub in range(ds_test.shape[0])]
 
     plt.figure()
-    plt.title('RMSE vs Lyapunov Time\nParam {} ; RMSE {}'.format(
-        results['params'][i], total_rmse))
+    plt.title('Test RMSE vs Lyapunov Time')
     t_adj = (ts_data.test_t - ts_data.test_t[0]) * lce[0]
     plt.plot(t_adj, rmse_over_t, 'o-')
     plt.plot(t_adj, np.zeros_like(t_adj))
@@ -67,7 +68,13 @@ for rank, i in enumerate(ranking[:how_many]):
     runt['results'].y = ys_test[:, [0, 2]].T
     slvr.plot2d(runt)
     plt.scatter([-sys.alpha, sys.mu], [0, 0])
+    title = 'ESN Trajectory\nParam {}\nValidation/Test RMSE {:.4}/{:.4}'
+    param_title = '(α:{}, N:{}, T0:{})'.format(*results['params'][i])
+    plt.title(title.format(param_title,
+                           rank_metric_cv[i],
+                           rank_metric_test[i]))
     name = 'rank_{}_param_{}_fit.png'.format(rank, i)
[…]

### 2019-04-18 doc/paper/paper.tex

@@
 
-Most well performing models couldn't predict past one Lyapunov constant. From
-the graphs, we can see that most models deviated once reaching the
+Most well performing models could predict around one Lyapunov time constant. 
+From the graphs, we can see that most models deviated once reaching the
 intersection of the butterfly wings.
@@
 
+Generating data for the 3 body problem was more difficult since we have to becareful that the trajectory doesn't diverge or collide with the other bodies.
+
+When the bodies collide, then the problem becomes a stiff problem and the
+trajectory might be inaccurate, and when the bodies diverge, the trajectory
+becomes a simple spiral without chaotic properties. Initial conditions which
+don't run into these issues are usually orbital. Such solutions exist
+\cite{frnkacase}, and \texttt{echonn} finds them by randomly generating initial
+conditions, and testing for the conditions. Orbital patterns happen frequently
+enough that they can be found easily with this method.
+
+The case I analyze, the initial conditions were found to be 
+$(x, x', y, y')_{t=0} = (0.79563053, 0, 0, 0.96833226)$
+with a largest Lyapunov Exponent of 0.1759. This case is much different than 
+the other systems since it doesn't have strong chaotic properties. This can
+be seen in the orbital pattern which seems quasi-periodic, and it might be. 
+In light of this, it's also very possible that the measured chaotic properties
+are solely due to error in the numerical calculation. However, the orbiting
+doesn't repeat exactly, so there are still small complexities for the ESN
+to pick up on. 
 %% CONCLUSION %%%

### 2019-04-18 doc/paper/scripts/AnalyzeResultsDoubPend.py
update plotnd generations

@@ -34,13 +34,13 @@ slvr = SystemSolver(run['system'])
 runt = deepcopy(run)
 runt['results'].t = ts_data.t
 runt['results'].y = ts_data.y.T
-fig = slvr.plotnd(runt)
+fig = slvr.plotnd(runt, dims=['θ1', 'θ2', 'ω1', 'ω2'], overlay=False)
 plt.savefig(os.path.join(dir_pre, 'full_differential.png'))
 # plt.show(True)
 
 runt['results'].t = ts_data.test_t
 runt['results'].y = ds_test.T
-slvr.plotnd(runt)
+slvr.plotnd(runt, dims=['θ1', 'θ2', 'ω1', 'ω2'], overlay=False)
 plt.savefig(os.path.join(dir_pre, 'test_data.png'))
 sorter = np.flip(np.argsort(score))
 how_many = 5
@@ -52,8 +52,7 @@ for rank, i in enumerate(sorter[:how_many]):
                    for sub in range(ds_test.shape[0])]
 
     plt.figure()
-    plt.title('RMSE vs Lyapunov Time\nParam {} ; RMSE {}'.format(
-        results['params'][i], total_rmse))
+    plt.title('Test RMSE vs Lyapunov Time')
     t_adj = (ts_data.test_t - ts_data.test_t[0]) * lce[0]
     plt.plot(t_adj, rmse_over_t, 'o-')
     plt.plot(t_adj, np.zeros_like(t_adj))
@@ -63,7 +62,11 @@ for rank, i in enumerate(sorter[:how_many]):
 
     runt['results'].t = ts_data.test_t
     runt['results'].y = ys_test.T
-    slvr.plotnd(runt, dims=['θ1', 'θ2', 'ω1', 'ω2'])
+    title = 'ESN Trajectory | Param {} | Test RMSE {:.4}'
+    param_title = '(α:{}, N:{}, T0:{})'.format(*results['params'][i])
+    title = title.format(param_title, total_rmse)
+    slvr.plotnd(runt, dims=['θ1', 'θ2', 'ω1', 'ω2'],
+                title=title, overlay=False)
     name = 'rank_{}_param_{}_fit.png'.format(rank, i)
     plt.savefig(os.path.join(
         dir_pre, name))

### 2019-04-18 doc/paper/scripts/AnalyzeResultsLorenz.py

@@ -43,7 +43,7 @@ fig = slvr.plot3d(runt)
 plt.savefig(os.path.join(dir_pre, 'full_differential.png'))
 
 sorter = np.flip(np.argsort(score))
-how_many = 10
+how_many = 8
 for rank, i in enumerate(sorter[:how_many]):
     ds_test, ys_test, total_rmse = results['best model rmse'][i][3]
     print(i, total_rmse, results['params'][i])
@@ -52,8 +52,7 @@ for rank, i in enumerate(sorter[:how_many]):
                    for sub in range(ds_test.shape[0])]
 
     plt.figure()
-    plt.title('RMSE vs Lyapunov Time\nParam {} ; RMSE {}'.format(
-        results['params'][i], total_rmse))
+    plt.title('Test RMSE vs Lyapunov Time')
     t_adj = (ts_data.test_t - ts_data.test_t[0]) * lce[0]
     plt.plot(t_adj, rmse_over_t, 'o-')
     plt.plot(t_adj, np.zeros_like(t_adj))
@@ -71,8 +70,9 @@ for rank, i in enumerate(sorter[:how_many]):
     runt['results'].y = ys_test[score[i]:higher_score[i]].T
     slvr.plot3d(runt, fig)
     plt.legend(['actual', 'approximated', 'deviation'])
-    plt.title('Model Approximation\nParam {} ; RMSE {}'.format(
-        results['params'][i], total_rmse))
+    title = 'ESN Trajectory\nParam {}\nTest RMSE {:.4}'
+    param_title = '(α:{}, N:{}, T0:{})'.format(*results['params'][i])
+    plt.title(title.format(param_title, total_rmse))
     plt.tight_layout()
     name = 'rank_{}_param_{}_fit.png'.format(rank, i)
     plt.savefig(os.path.join(

### 2019-04-18 src/echonn/sys/system_solver.py

@@ -174,7 +174,7 @@ class SystemSolver:
         plt.tight_layout()
         return fig
 
-    def plotnd(self, run=None, fig=None, figsize=None, legend=False, dims=None):
+    def plotnd(self, run=None, fig=None, figsize=None, legend=False, dims=None, overlay=True, title=None):
         """ warning legend=True is going to mess up the layout. The legend can be seen by the side of the mini graph. """
         run = self.get_last_run(run)
         sys_num = run['index']
@@ -182,7 +182,9 @@ class SystemSolver:
 
         if fig is None:
             max_rows = 8
-            num_plots = res.y.shape[0] + 1  # +1 for the overlay
+            num_plots = res.y.shape[0]
+            if overlay:
+                num_plots += 1
             rows = min(max_rows, num_plots)
             cols = max(1, math.ceil(num_plots/max_rows))
             if num_plots > max_rows:
@@ -193,8 +195,9 @@ class SystemSolver:
         else:
             fig = plt.figure(fig.number)
         plots = fig.get_axes()
-        overlay = plots[-1]
-        overlay.set_ylabel('overlay')
+        if overlay:
+            overlay_plt = plots[-1]
+            overlay_plt.set_ylabel('overlay')
         for i, y in enumerate(res.y):
             for _ in range(i+1):
                 plots[i].plot(res.t, y)  # TODO: fix this hack
@@ -204,8 +207,13 @@ class SystemSolver:
             else:
                 plots[i].set_ylabel(dims[i])
             label = 'sys {} dim {}'.format(sys_num, i)
-            overlay.plot(res.t, y, label=label)
-        if legend:
+            if overlay:
+                overlay_plt.plot(res.t, y, label=label)
+        if legend and overlay:
             overlay.legend()
-        plt.tight_layout()
+        if title is not None:
+            fig.suptitle(title)
+            fig.tight_layout(rect=[0, 0.03, 1, 0.95])
+        else:
+            plt.tight_layout()
         return fig

### 2019-04-18 doc/paper/paper.tex
Updates from Overleaf

@@
 only gradual but seemingly random variations until a few spots around $t=10,
-30, 60, \text{and} 75$. The results for the ESN weren't very strong, but it
-was able to predict almost two Lyapunov Times before diverging.
+30, 60, \text{and} 75$. The results for the ESN weren't very strong, and none 
+could predict more than one Lyapunov Times before diverging.
 
@@
     \begin{subfigure}[b]{0.45\textwidth}
-        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_0_param_225_rmse.png}
+        \includegraphics[width=\textwidth]{doc/paper/images/doub_pend/rank_0_param_77_rmse.png}
         \caption{RMSE Over Time}
@@
     \begin{subfigure}[b]{0.45\textwidth}
-        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_1_param_230_rmse.png}
+        \includegraphics[width=\textwidth]{doc/paper/images/doub_pend/rank_2_param_305_rmse.png}
         \caption{RMSE Over Time}
@@
     \begin{subfigure}[b]{0.45\textwidth}
-        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_2_param_458_rmse.png}
+        \includegraphics[width=\textwidth]{doc/paper/images/doub_pend/rank_2_param_305_rmse.png}
         \caption{RMSE Over Time}
@@
     \begin{subfigure}[b]{0.45\textwidth}
-        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_3_param_151_rmse.png}
+        \includegraphics[width=\textwidth]{doc/paper/images/doub_pend/rank_3_param_52_rmse.png}
         \caption{RMSE Over Time}
@@
     \begin{subfigure}[b]{0.45\textwidth}
-        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/rank_4_param_300_rmse.png}
+        \includegraphics[width=\textwidth]{doc/paper/images/doub_pend/rank_4_param_127_rmse.png}
         \caption{RMSE Over Time}
@@
 doesn't repeat exactly, so there are still small complexities for the ESN
-to pick up on. 
+to pick up on. In the results, we see that the ESN performs very well, but
+it's possible that the ESN is just converging to the mean of the orbits. 
+The steep increase in error and the subsequent slowing of increase in error
+seen in the graphs below suggests that this might be the case.
+Since the orbits are fairly close, this results in a sufficiently
+accurate approximation.
+
+Most RMSE graphs look very similar. There is a steep increase in RMSE at first
+but then it slows. Overall, the RMSE is really strong. Most don't rise above 
+0.2, and one show below (Figure \ref{fig:3brounded}) shows an RMSE not rising 
+above 0.1 despite its not as elliptical looking shape. Further, this model
+is impressive because it only has 5 internal units. Most other top performers 
+have over 100.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.5\linewidth]{doc/paper/images/r3body/full_differential.png}
+    \caption{The entire difference dataset}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.5\linewidth]{doc/paper/images/r3body/test_data.png}
+    \caption{The testing dataset partition}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/r3body/rank_3_param_15_fit.png}
+        \caption{Predicted Trajectory}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/r3body/rank_3_param_15_rmse.png}
+        \caption{RMSE Over Time}
+    \end{subfigure}
+    \caption{The best validation and testing performance and only 5 internal units}
+    \label{fig:3brounded}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/r3body/rank_0_param_566_fit.png}
+        \caption{Predicted Trajectory}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/r3body/rank_0_param_566_rmse.png}
+        \caption{RMSE Over Time}
+    \end{subfigure}
+    \caption{Best performer on validation set}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/r3body/rank_1_param_366_fit.png}
+        \caption{Predicted Trajectory}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/r3body/rank_1_param_366_rmse.png}
+        \caption{RMSE Over Time}
+    \end{subfigure}
+    \caption{A top performer on validation set}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/r3body/rank_2_param_262_fit.png}
+        \caption{Predicted Trajectory}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/r3body/rank_2_param_262_rmse.png}
+        \caption{RMSE Over Time}
+    \end{subfigure}
+    \caption{A top performer on validation set}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/r3body/rank_4_param_448_fit.png}
+        \caption{Predicted Trajectory}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/r3body/rank_4_param_448_rmse.png}
+        \caption{RMSE Over Time}
+    \end{subfigure}
+    \caption{A top performer on validation set}
+\end{figure}
+
 %% CONCLUSION %%%

### 2019-04-18 doc/paper/scripts/AnalyzeResultsLorenz.py
better formatting of lorenz pictures

@@ -60,16 +60,19 @@ for rank, i in enumerate(sorter[:how_many]):
     plt.savefig(os.path.join(dir_pre, name))
     plt.close()
 
-    runt['results'].t = ts_data.test_t
-    runt['results'].y = ds_test.T
+    #runt['results'].t = ts_data.test_t
+    #runt['results'].y = ds_test.T
+    #fig = slvr.plot3d(runt)
+    runt['results'].t = ts_data.test_t[:higher_score[i]+1]
+    runt['results'].y = ds_test[:higher_score[i]+1].T
     fig = slvr.plot3d(runt)
-    runt['results'].t = ts_data.test_t[:score[i]]
-    runt['results'].y = ys_test[:score[i]].T
+    #runt['results'].t = ts_data.test_t[:score[i]+1]
+    #runt['results'].y = ys_test[:score[i]+1].T
+    #slvr.plot3d(runt, fig)
+    runt['results'].t = ts_data.test_t[:higher_score[i]+1]
+    runt['results'].y = ys_test[:higher_score[i]+1].T
     slvr.plot3d(runt, fig)
-    runt['results'].t = ts_data.test_t[score[i]:higher_score[i]]
-    runt['results'].y = ys_test[score[i]:higher_score[i]].T
-    slvr.plot3d(runt, fig)
-    plt.legend(['actual', 'approximated', 'deviation'])
+    plt.legend(['actual', 'approximation'])
     title = 'ESN Trajectory\nParam {}\nTest RMSE {:.4}'
     param_title = '(α:{}, N:{}, T0:{})'.format(*results['params'][i])
     plt.title(title.format(param_title, total_rmse))
@@ -77,5 +80,5 @@ for rank, i in enumerate(sorter[:how_many]):
     name = 'rank_{}_param_{}_fit.png'.format(rank, i)
     plt.savefig(os.path.join(
         dir_pre, name))
-    plt.close()
     # plt.show(True)
+    plt.close()

### 2019-04-18 doc/paper/paper.tex
Updates from Overleaf

@@
 \newcommand{\echonn}{\texttt{echonn}}
+\newcommand{\echonnsp}{\texttt{echonn}\text{ }}
 \newcommand{\der}[2][t]{\frac{\mathrm{d}#2}{\mathrm{d}#1}}
@@
 the computer used 6 digit accuracy, it only output 3 digits, so when the
-simulation was continued by Lorenz with the less accurate measurements, he
+simulation was continued with the less accurate measurements, Lorenz
 found the results to be very different \cite{oestreicher2007history}.
@@
 
-In terms of forecasting, it's useful to know how chaotic a dynamical system
-is. Chaos will be measured by the exponential rate at which nearby
+It's useful to know how chaotic a dynamical system
+is. Chaos is usually measured by the exponential rate at which nearby
 trajectories diverge. Constants describing the divergence are known as the
-Lyapunov Characteristic Exponents (LCEs) \cite{sandri1996numerical}. There is
-an LCE for each dimension, and since they all represent exponential growth.
-The error growing at the rate of the largest LCE will dominate over the
+Lyapunov Characteristic Exponents (LCEs) \cite{sandri1996numerical}.
+There is
+an LCE for each dimension, and since they all represent exponential growth,
+the error growing at the rate of the largest LCE will dominate over the 
 others.
-
 Given a small perturbation $\epsilon$ to a dynamical system in the direction
@@
 \lambda$. This is known as the Lyapunov time \cite{bezruchko2010extracting}.
-For our solar system, the Lyapunov time is 10,000,000 years
-\cite{oestreicher2007history}. Researchers have used it to evaluate the
-effectiveness of forecasting models on chaotic problems
-\cite{pathak2018model}. Predicting the solution of a chaotic model over an
+For example, the Lyapunov time of the orbits of our solar system is
+10,000,000 years
+\cite{oestreicher2007history}.
+
+Researchers have used it to evaluate the
+effectiveness of forecasting models on chaotic problems in the past.
+Predicting the solution of a chaotic model over an
 interval of time which is multiple times the size of the Lyapunov time shows
-the predictability of a model.
+the predictability of a model. One group of researchers studied the 
+Kuramoto-Sivashinsky chaotic model near an attractor and found that with a max
+LCE of 0.1, they were able to predict 8 Lyapunov times before their forecast 
+diverged \cite{pathak2018model}.
 
@@
 In order to find the Lyapunov time, we only need to solve for the largest
-LCE. To find this, we will use the following method detailed in
+LCE. To find this, we will use the following method
 \cite{viswanath1998lyapunov}.
@@
 
-where $\Mv(t) \in \mathbb{R}^{d \times d}$ and $\frac{\partial \fv}{\partial
-\xv}$ is the $d \times d$ Jacobian matrix of $\fv(\xv)$ at $\xv(t)$.
+where $\Mv(t) \in \mathbb{R}^{d \times d}$ and
+$\frac{\partial \fv}{\partial \xv}$
+is the $d \times d$ Jacobian matrix of $\fv(\xv)$ at $\xv(t)$ and with 
+$\Mv(0)$ equal to the identity matrix.
 
@@
 the dynamical system is implemented in \echonn, its max LCE can't be
-calculated accurately since the resultant system has 342 dimensions. The
+calculated accurately since the resulting system has 342 dimensions. The
 algorithm was also tested on the differential equation which produces a
@@
 
-Empirically shown, \echonn can be used to identify the max LCE of smaller
+Empirically shown, \echonnsp can be used to identify the max LCE of smaller
 dynamical systems, and special precaution has to be taken when choosing a
@@
 due to the chaotic nature, error introduced by the numerical solution will
-grow exponentially, so precaution need to be taken to avoid having this sort
+grow exponentially, so precaution needs to be taken to avoid having this sort
 of error introduced into the collected data. When numerically solving chaotic
@@
 shown in Table \ref{table:lorenz_params} are the values calculated by my
-Python module \echonn are listed. These values are not as accurate, but
-demonstrate the sufficient accuracy of \echonn.
+Python module \echonnsp are listed. These values 
+demonstrate the accuracy of \echonn.
 
@@
 
+% TODO: Graph of converging LCE
+
 The Lorenz system is named after its creator, but it is also known as the
@@
 to a much more useful form of the problem. The mass of one of the objects is
-assumed to be negligible. This is the case for Sun-Earth-Moon orbits. The
+assumed to be negligible. This is the case for Sun-Earth-Moon orbits: the
 moon has little effect on the orbit of the sun. Another aspect is the
@@
 approach each other, the distance between them approaches zero. If collision
-occurs, then this approaches infinity. There is no collision detection in
+occurs, then the diverging trajectory approaches infinity. There is
+no collision detection in
 this model, and each body has a radius of zero, so this case is not handled.
@@
 Under the right initial conditions, the third body can diverge regardless
-(fast initial velocity away from the bodies). The simple divergent spiral
+(e.g., fast initial velocity away from the bodies).
+The simple divergent spiral
 isn't chaotic, so instead orbital and collision less initial conditions will
@@
 
-Although, it's rare that the output units are calculated from previous output
+Although, it's rare for the ESN that the output units are calculated
+directly from previous output
 units \cite{jaeger2002tutorial}. Sometimes for training, the output units are
-set to the value of $dv$, called teacher forcing, so that the model trains on
+set to the value of $\dv$, called teacher forcing, so that the model trains on
 actual values as opposed to its accumulated errors \cite{jaeger2002tutorial}.
@@
 How the units are connected can vary, but the suggested way
-\cite{jaeger2002tutorial} is to connect the internal layer to the current
-input layer, the previous internal layer, and the previous output layer.
+\cite{jaeger2002tutorial} is to connect the current
+input layer, the previous internal layer, and the previous output layer to
+the internal layer.
 These connections will have weights described in \ref{sec:rnn}. Respectively,
@@
 
-The output layer is connected to the current input layer and current internal
+Connected to the output layer is the current input layer and current internal
 layer. Additionally, a bias input is connected to this layer. The weight for
@@
 form of normalization.
+
+The dimensions of the weight matrices are as follows:
+$\Win \in \mathbb{R}^{N \times K}$,
+$\W \in \mathbb{R}^{N \times N}$,
+$\Wback \in \mathbb{R}^{N \times L}$, and
+$\Wout \in \mathbb{R}^{N \times (K+1+N)}$,
 
@@
 \end{align}
+
+"Feeding data" into the model means iterating using values for $\uv$. If the 
+feeding is teacher-forced, then the output layer's values are set to $\dv$. 
+Otherwise, the calculated output values are used.
 
@@
 This suggests that as the model iterates, the internal state converges to
-zero.
+zero. Also, $\tanh$ is contractive further supporting this idea.
 
@@
 
-Second, the initialization data is fed into the model. This initializes the
+Second, the initialization data is fed into the model with teacher-forced 
+output. This initializes the
 inner state of the model, and the initial state (zeros) is lost due to the
@@
 
-Third, the training values are fed into the model to sample the network. We
-can store the input unit values, the internal state values, and the desired
-output values as matrices $M$ and $T$. $M$ will store the input unit values
-and internal state value pairs $(\uv(t_{T_0}), 1, \xv(t_{T_0})), \dots,
-(\uv(t_T), 1, \xv(t_T))$ in its rows. $T$ will store the output unit values
-$\dv(t_{T_0}), \dots, \dv(t_T)$.
-
-To train the model, we want $\Wout M$ to approximate accurately $g^{-1}(T)$,
-where $g$ is the activation function for the output layer. Since we use an
-idempotent output layer activation function, $g^{-1}(T) = T$.
+Third, the training values are fed into the model to sample the network with 
+teacher-forced output. We define a matrix $M$ to store the input unit values,
+bias, and internal state value pairs, $(\uv(t_{T_0}), 1, \xv(t_{T_0})), \dots,
+(\uv(t_T), 1, \xv(t_T))$, in its rows. We define a matrix $T$ to store the 
+output unit values: $\dv(t_{T_0}), \dots, \dv(t_T)$.
+
+To train the model, we want $g(\Wout M)$ to approximate accurately 
+$T$, where $g$ is the activation function for the output layer.
+Since we use an
+idempotent output layer activation function, $g^{-1}(X) = g(X) = X$, so
+
+\begin{align*}
+    g(\Wout M) &= T \\
+    \Wout M &= g^{-1}(T) \\
+    \Wout M &= T
+\end{align*}
 
@@
 $\lambda_1$. Then, the Lyapunov Time is calculated by $\frac{1}{\lambda_1}$.
-We will then integrate the model with an algorithm called RK4(5) over the
+We will then integrate the model with the algorithm RK4(5) over the
 interval $[0, \frac{50}{\lambda_1}]$ with time steps of $\frac{1}{100
@@
 
-The first data point of the training set is $t_0 = 0$. The parameter $T_0$
+The first data point of the training set is $t = 0$. The parameter $T_0$
 marks where the model begins training on the data (see Section
@@
 we want to know how far the model can predict before deviating too far. The
-$\text{RMSE}_\text{test}(t_f)$ will be calculated on the interval $[T_f,
-t_f)$ and the prediction uses teacher-forced values on $[0, T_f)$. Lastly,
-$\text{RMSE}_\text{test}$ will notate the RMSE for the entire test set.
+$\text{RMSE}_\text{test}(t)$ will be calculated on the interval $[T_f,
+t)$ and the prediction uses teacher-forced values on $[0, T_f)$. Lastly,
+$\text{RMSE}_\text{test}$ as a value instead of a function
+will notate the RMSE for the entire test set.
 
 We will look at how far past $T_f$ the ESN model's $\text{RMSE}_\text{test}$
-grows too large. The best model will be the model with the lowest Testing
-RMSE for the longest time.
+grows too large. The best model will be the model with the lowest validation 
+RMSE as well as a low testing RMSE.
 
 \subsection{Results}
+
+Overall, the systems proved too difficult to forecast; however, this could 
+be due to lacking data. While more data can generated, numerical errors might 
+grow too large. Since we have about 5,000 data points, we can't have an 
+internal state dimension of more than 500 or we risk over-fitting 
+\cite{jaeger2002tutorial}.
+
+The Lorenz model showed strong forecasting results, but due to the shape of 
+the attractor, it was easy for the small perturbations to cause quick 
+divergence as the trajectories followed the incorrect "wing" of the butterfly.
+
+The double pendulum model showed very poor results. Occasionally, the pendulum
+would move much quicker causing spikes in the data. The ESN tried to predict 
+when these spikes occurred, but was unable. Even the forecasted shape of the 
+rapid movement was inaccurate. This is likely due to the lack of data.
+
+The restricted circular three body problem performed much better than the
+other two, but this is due to the existence of a simple attractor: a single 
+circular orbit as opposed to two rings in the Lorenz system. Also, the max LCE
+was much smaller for this system (0.1759 vs 0.9 and 0.6) suggesting that it 
+was less chaotic. While
+it's possible that the chaotic behavior was created by the numerical
+solution, it suggests the strong predictability of lesser chaotic systems.
+
+In the case of researchers performing a similar analysis mentioned before 
+\cite{pathak2018model}, they too had strong performance for a less chaotic 
+problem. I wasn't able to predict 8 Lyapunov times, but there might be 
+reasons why my results weren't as strong other than the fact that
+the ESN might need more data that I used.
+The lorenz attractor is more chaotic than the attractor they used. The double 
+pendulum lacks an attractor and shows a higher level of chaos. The 3 body
+problem has a lower LCE and a simple attractor, and showed strong results as 
+well.
+
 \subsubsection{Lorenz}
 
-To integrate the model, I use the initial conditions: $(x_0, y_0, z_0) =
+To integrate the model, I use the initial conditions: $(x, y, z)_{t=0} =
 (-13.78990503, -11.6069343, 36.17332518)$. The calculated Lyapunov value was
-$0.9348566$.
+$\lambda_1 = 0.9348566$.
 
@@
     \end{subfigure}
-    \caption{The model diverges when entering never seen region.}
+    \caption{Diverges when entering space outside training data.}
     \label{fig:lorenz_r0}
@@
     \end{subfigure}
-    \caption{The model continues in wrong direction after intersection.}
+    \caption{Continues in wrong direction at intersection.}
     \label{fig:lorenz_r1}
@@
     \end{subfigure}
-    \caption{The model has completely deviated after an intersection.}
+    \caption{Deviates after an intersection.}
     \label{fig:lorenz_r2}
@@
     \end{subfigure}
-    \caption{Fits model really well, but follows wrong wings.}
+    \caption{Fits model well but follows wrong wings.}
     \label{fig:lorenz_r3}
@@
     \end{subfigure}
-    \caption{Fits model really well, but follows wrong wings.}
+    \caption{Fits model well but trajectory is ahead in time.}
     \label{fig:lorenz_r4}
@@
     \end{subfigure}
-    \caption{Doesn't fit model smoothy and eventually diverges.}
+    \caption{Diverges after intersection.}
     \label{fig:lorenz_r6}
@@
     \end{subfigure}
-    \caption{Sporadic behavior shown long after divergence.}
+    \caption{Sporadic behavior shown after divergence.}
     \label{fig:lorenz_r7}
@@
 \subsubsection{Double Pendulum}
-
-The double pendulum's initial condition was
-
-\begin{equation}
-(\theta_1, \theta_2, \omega_1, \omega_2) 
-= (4.66648088, 4.29820501,
-8.01585663 \times 10^{-4}, 3.71175422 \times 10^{-3}).
-\end{equation}
 
@@
 the rest of the parameters were chosen randomly (with rotational velocities
-close to zero). The resulting Lyapunov Exponent is $0.60074$. Like the
+close to zero). The double pendulum's initial condition was
+
+\begin{equation}
+(\theta_1, \theta_2, \omega_1, \omega_2)_{t=0} 
+= (4.66648088, 4.29820501,
+8.01585663 \times 10^{-4}, 3.71175422 \times 10^{-3}),
+\end{equation}
+
+and the resulting Lyapunov Exponent is $\lambda_1 = 0.60074$. Like the
 Lorenz, the differences between time points were used to train the ESN since
@@
 dimensions. It seems like the dataset's activity is fairly sparse. There is
-only gradual but seemingly random variations until a few spots around $t=10,
-30, 60, \text{and} 75$. The results for the ESN weren't very strong, and none 
+only gradual but seemingly random variations until a few spots around
+$t=10, 30, 60, \text{and } 75$. The results
+for the ESN weren't very strong, and none 
 could predict more than one Lyapunov Times before diverging.
@@
 Because there are portions of the dataset which are calm, the ESN then has to
-pickup on signals which lead to the spikes in movement. We see the testing
-graph that while it predicts the testing set well, it eventually tries to
+pickup on signals which lead to the spikes in movement. We see in the 
+graphs that while it predicts the testing set well, it eventually tries to
 predict spikes but does so at the wrong time, and the spike pattern never
 ends, has too large of magnitude, and is very rapid. Overall the model
-performs poorly, and it's reflected in the RMSE. The likely reason for this
+performs poorly, and it's reflected in the RMSE which quickly diverges well above 3. The likely reason for this
 is that the ESN can't learn the complex patterns from such a small and sparse
-dataset.
+dataset. With more data or more complex internals, maybe the ESN could learn 
+to forecast this dataset.
 
@@
 
-Generating data for the 3 body problem was more difficult since we have to becareful that the trajectory doesn't diverge or collide with the other bodies.
+Generating data for the 3 body problem is more difficult since we have to be 
+careful that the trajectory doesn't diverge or collide with the other bodies.
 
@@
 don't run into these issues are usually orbital. Such solutions exist
-\cite{frnkacase}, and \texttt{echonn} finds them by randomly generating initial
-conditions, and testing for the conditions. Orbital patterns happen frequently
+\cite{frnkacase}, and \echonnsp finds them by randomly generating initial
+conditions, and testing for the conditions in the numerical solution.
+Orbital patterns happen frequently
 enough that they can be found easily with this method.
@@
 $(x, x', y, y')_{t=0} = (0.79563053, 0, 0, 0.96833226)$
-with a largest Lyapunov Exponent of 0.1759. This case is much different than 
+with a largest Lyapunov Exponent of $\lambda_1 = 0.1759$. This case is
+much different than 
 the other systems since it doesn't have strong chaotic properties. This can
-be seen in the orbital pattern which seems quasi-periodic, and it might be. 
+be seen in the orbital pattern which seems quasi-periodic (which it may be). 
 In light of this, it's also very possible that the measured chaotic properties
@@
 it's possible that the ESN is just converging to the mean of the orbits. 
-The steep increase in error and the subsequent slowing of increase in error
-seen in the graphs below suggests that this might be the case.
+The steep increase in error and the subsequent slowing of the increase
+in error, seen in the graphs below, suggests that this might be the case.
 Since the orbits are fairly close, this results in a sufficiently
@@
 but then it slows. Overall, the RMSE is really strong. Most don't rise above 
-0.2, and one show below (Figure \ref{fig:3brounded}) shows an RMSE not rising 
-above 0.1 despite its not as elliptical looking shape. Further, this model
+0.2, and one shown below (Figure \ref{fig:3brounded}) shows
+an RMSE not rising 
+above 0.1 despite its not as elliptical looking shape. Maybe this model
+is truly capturing the complexity of the system. Further, this model
 is impressive because it only has 5 internal units. Most other top performers 
-have over 100.
+have over 100 internal units.
 

### 2019-04-23 doc/paper/scripts/AnalyzeResultsDoubPend.py
add lce convergence and figure 8 soln to 3 body

@@ -8,8 +8,29 @@ from echonn.sys import SystemSolver
 
 res = pickle.load(open('doub_pend_results.p', 'rb'))
 run, lce, ts_data, results = res
+sys = run['system']
+slvr = SystemSolver(sys)
+dir_pre = os.path.join('..', 'images', 'doub_pend')
 
+# print small details
 print('lce:', lce[0])
+# LCE
+lce_val, lce_run = lce
+T0 = 0
+t = lce_run['results'].t[T0:]
+y = lce_run['results'].y[:, T0:]
+lces = []
+for i, t_val in enumerate(t):
+    Df_y0 = y[sys.dim:, i].reshape(sys.dim, sys.dim)
+    lces.append(slvr.calc_lce(Df_y0, t_val))
+plt.figure()
+plt.title(f'Double Pendulum LCE ({lce_val:.3}) vs t')
+plt.xlabel('t')
+plt.ylabel('LCE')
+plt.plot(t, lces)
+plt.savefig(os.path.join(dir_pre, 'lce_converge.png'))
+exit()
+
 print(run['results'].y[:, 0])
 test_rmse = [rmse for _, _, _, (_, _, rmse) in results['best model rmse']]
 test_rmse = np.array(test_rmse)
@@ -29,7 +50,6 @@ for rmse_res in results['best model rmse']:
         if err > .05:
             score.append(sub)
             break
-dir_pre = os.path.join('..', 'images', 'doub_pend')
 slvr = SystemSolver(run['system'])
 runt = deepcopy(run)
 runt['results'].t = ts_data.t

### 2019-04-23 doc/paper/scripts/AnalyzeResultsLorenz.py

@@ -8,8 +8,30 @@ from echonn.sys import SystemSolver
 
 res = pickle.load(open('lorenz_results.p', 'rb'))
 run, lce, ts_data, results = res
+sys = run['system']
+slvr = SystemSolver(sys)
+dir_pre = os.path.join('..', 'images', 'lorenz')
 
 print('lce:', lce[0])
+
+# LCE
+lce_val, lce_run = lce
+T0 = 0
+t = lce_run['results'].t[T0:]
+y = lce_run['results'].y[:, T0:]
+lces = []
+for i, t_val in enumerate(t):
+    Df_y0 = y[sys.dim:, i].reshape(sys.dim, sys.dim)
+    lces.append(slvr.calc_lce(Df_y0, t_val))
+plt.figure()
+plt.title(f'Lorenz LCE ({lce_val:.3}) vs t')
+plt.xlabel('t')
+plt.ylabel('LCE')
+plt.plot(t, lces)
+plt.savefig(os.path.join(dir_pre, 'lce_converge.png'))
+exit()
+
+# results
 print(run['results'].y[:, 0])
 test_rmse = [rmse for _, _, _, (_, _, rmse) in results['best model rmse']]
 test_rmse = np.array(test_rmse)
@@ -34,8 +56,6 @@ for rmse_res in results['best model rmse']:
         if err > 1:
             higher_score.append(sub)
             break
-dir_pre = os.path.join('..', 'images', 'lorenz')
-slvr = SystemSolver(run['system'])
 runt = deepcopy(run)
 runt['results'].t = ts_data.t
 runt['results'].y = ts_data.y.T

### 2019-04-23 doc/paper/scripts/AnalyzeResultsR3Body.py

@@ -8,9 +8,31 @@ from echonn.sys import SystemSolver
 
 res = pickle.load(open('rc3_body_results.p', 'rb'))
 run, lce, ts_data, results = res
+sys = run['system']
+slvr = SystemSolver(sys)
+
+# get img dir
+dir_pre = os.path.join('..', 'images', 'r3body')
 
 # print small details
 print('lce:', lce[0])
+# LCE
+lce_val, lce_run = lce
+T0 = 0
+t = lce_run['results'].t[T0:]
+y = lce_run['results'].y[:, T0:]
+lces = []
+for i, t_val in enumerate(t):
+    Df_y0 = y[sys.dim:, i].reshape(sys.dim, sys.dim)
+    lces.append(slvr.calc_lce(Df_y0, t_val))
+plt.figure()
+plt.title(f'Restricted 3 Body LCE ({lce_val:.3}) vs t')
+plt.xlabel('t')
+plt.ylabel('LCE')
+plt.plot(t, lces)
+plt.savefig(os.path.join(dir_pre, 'lce_converge.png'))
+exit()
+
 print('ic:', run['results'].y[:, 0])
 print('tf:', run['results'].t[-1])
 
@@ -25,9 +47,6 @@ def rmse(d, y):
     num_samples = d.shape[0]
     return np.sqrt(np.sum((d-y)**2) / num_samples)
 
-
-# get img dir
-dir_pre = os.path.join('..', 'images', 'r3body')
 
 # extract system and build solver
 sys = run['system']

### 2019-04-23 doc/paper/scripts/ThreeBodyFigEight.py

@@ -1,0 +1,72 @@
+import os
+from echonn.sys import NBodySystem, SystemSolver
+import numpy as np
+import matplotlib.pyplot as plt
+
+dir_pre = os.path.join('..', 'images', '3body')
+sys = NBodySystem(body_masses=[1, 1, 1], G=1)
+solver = SystemSolver(sys)
+tspan = [0, 1.5]
+y0 = np.zeros(2*sys.body_dim*len(sys.body_masses), dtype=np.float64)
+
+x1 = np.array([0.97000436, -0.24308753, 0])
+x3p = np.array([-0.93240737, -0.86473146, 0])
+
+y0[0:3] = x1
+y0[3:6] = -x1
+y0[6:9] = 0
+y0[9:12] = -x3p / 2
+y0[12:15] = -x3p / 2
+y0[15:18] = x3p
+# print(sys.fun(np.zeros_like(y0), y0).reshape(6, -1))
+lce, run = solver.get_lce(tspan[1], y0)
+t = run['results'].t[1:]
+y = run['results'].y[sys.dim:, 1:].reshape(sys.dim, sys.dim, -1)
+#print(y[:, :, -1])
+lces = []
+for i, t_val in enumerate(t):
+    Df_y = y[:, :, i]
+    lces.append(solver.calc_lce(Df_y, t_val))
+print(lces[-1])
+print(np.mean(lces[-5:]))
+
+plt.figure()
+plt.title(f'Incomplete LCE ({lces[-1]:.2}) Convergence for 3 Body Figure 8')
+plt.plot(t, lces)
+plt.ylabel('LCE')
+plt.ylabel('t')
+plt.savefig(os.path.join(dir_pre, 'lce_converge.png'))
+
+tspan = [0, 100]
+# print(sys.fun(np.zeros_like(y0), y0).reshape(6, -1))
+run = solver.run(tspan, y0)
+# solver.plotnd(run)
+# print(run['results'].y[:, -1].reshape(6, -1))
+y_act = run['results'].y[:9]
+run['results'].y = y_act[:3]
+fig = solver.plot3d(run)
+run['results'].y = y_act[3:6]
+fig = solver.plot3d(run, fig=fig)
+run['results'].y = y_act[6:9]
+fig = solver.plot3d(run, fig=fig)
+plt.title('Long Run of 3 Body Figure 8')
+plt.tight_layout()
+plt.savefig(os.path.join(dir_pre, 'fig8_long.png'))
+
+
[…]

### 2019-04-23 doc/paper/scripts/ThreeBodyFigEight.py
fix 3 body fig 8 soln graph

@@ -6,7 +6,7 @@ import matplotlib.pyplot as plt
 dir_pre = os.path.join('..', 'images', '3body')
 sys = NBodySystem(body_masses=[1, 1, 1], G=1)
 solver = SystemSolver(sys)
-tspan = [0, 1.5]
+
 y0 = np.zeros(2*sys.body_dim*len(sys.body_masses), dtype=np.float64)
 
 x1 = np.array([0.97000436, -0.24308753, 0])
@@ -18,24 +18,6 @@ y0[6:9] = 0
 y0[9:12] = -x3p / 2
 y0[12:15] = -x3p / 2
 y0[15:18] = x3p
-# print(sys.fun(np.zeros_like(y0), y0).reshape(6, -1))
-lce, run = solver.get_lce(tspan[1], y0)
-t = run['results'].t[1:]
-y = run['results'].y[sys.dim:, 1:].reshape(sys.dim, sys.dim, -1)
-#print(y[:, :, -1])
-lces = []
-for i, t_val in enumerate(t):
-    Df_y = y[:, :, i]
-    lces.append(solver.calc_lce(Df_y, t_val))
-print(lces[-1])
-print(np.mean(lces[-5:]))
-
-plt.figure()
-plt.title(f'Incomplete LCE ({lces[-1]:.2}) Convergence for 3 Body Figure 8')
-plt.plot(t, lces)
-plt.ylabel('LCE')
-plt.ylabel('t')
-plt.savefig(os.path.join(dir_pre, 'lce_converge.png'))
 
 tspan = [0, 100]
 # print(sys.fun(np.zeros_like(y0), y0).reshape(6, -1))

### 2019-04-23 doc/paper/scripts/ThreeBodyFigEightLCE.py

@@ -1,0 +1,38 @@
+import os
+from echonn.sys import NBodySystem, SystemSolver
+import numpy as np
+import matplotlib.pyplot as plt
+
+dir_pre = os.path.join('..', 'images', '3body')
+sys = NBodySystem(body_masses=[1, 1, 1], body_dim=2, G=1)
+solver = SystemSolver(sys)
+tspan = [0, 200]
+y0 = np.zeros(2*sys.body_dim*len(sys.body_masses), dtype=np.float64)
+
+x1 = np.array([0.97000436, -0.24308753])
+x3p = np.array([-0.93240737, -0.86473146])
+
+y0[0:2] = x1
+y0[2:4] = -x1
+# y0[4:6] = zero
+y0[6:8] = -x3p / 2
+y0[8:10] = -x3p / 2
+y0[10:12] = x3p
+
+# print(sys.fun(np.zeros_like(y0), y0).reshape(6, -1))
+lce, run = solver.get_lce(tspan[1], y0)
+t = run['results'].t[1:]
+y = run['results'].y[sys.dim:, 1:].reshape(sys.dim, sys.dim, -1)
+#print(y[:, :, -1])
+lces = []
+for i, t_val in enumerate(t):
+    Df_y = y[:, :, i]
+    lces.append(solver.calc_lce(Df_y, t_val))
+
+plt.figure()
+plt.title(f'LCE ({lces[-1]:.2}) Convergence for 3 Body Figure 8')
+plt.plot(t, lces)
+plt.ylabel('LCE')
+plt.xlabel('t')
+plt.savefig(os.path.join(dir_pre, 'lce_converge.png'))
+# plt.show(True)

### 2019-04-23 doc/paper/paper.tex
Updates from Overleaf

@@
    F_g^{(i)} = m_i \frac{\mathrm{d}^2 \rv_i}{\mathrm{dt}^2} =
-   \sum_{j \neq i} G \frac{m_i m_j \hat{\rv}_{ij}}{\norm{\rv_{ij}}^3}
+   \sum_{j \neq i} G \frac{m_i m_j \rv_{ij}}{\norm{\rv_{ij}}^3}
    \label{eq:nbody}
@@
 known input data, $\wv$ is the model's parameters, and $\yv$ approximates
-some values, $\tv$, then we define an objective function as the sum of
+some values, $\dv$, then we define an objective function as the sum of
 squared error:
@@
 \begin{equation}
-    \Ev(\xv; \wv) = \vert \vert \yv(\xv; \wv) - \tv \vert \vert_2^2.
+    \Ev(\xv; \wv) = \vert \vert \yv(\xv; \wv) - \dv \vert \vert_2^2.
     \label{eq:error_eq_rnn}
@@
 Third, the training values are fed into the model to sample the network with 
-teacher-forced output. We define a matrix $M$ to store the input unit values,
+teacher-forced output. We define a matrix
+$M \in \mathbb{R}^{(K + 1 + N) \times (T-T_0+1)}$
+to store the input unit values,
 bias, and internal state value pairs, $(\uv(t_{T_0}), 1, \xv(t_{T_0})), \dots,
-(\uv(t_T), 1, \xv(t_T))$, in its rows. We define a matrix $T$ to store the 
+(\uv(t_T), 1, \xv(t_T))$, in its columns. We define a matrix
+$T \in \mathbb{R}^{L \times (T-T_0+1)}$ to store the 
 output unit values: $\dv(t_{T_0}), \dots, \dv(t_T)$.

### 2019-04-24 doc/paper/paper.tex
first draft complete

@@
 
-Neural Networks are known to be able to approximate any function given enough
-data and time to train \cite{cybenko1989approximation}. In light of this, the
-hope is that a deep learning based model, the ESN, will outperform ARIMA.
-
-This is a paper about chaos.
-% TODO: Intro
-% TODO: mention \texttt{echonn}
+Chaotic dynamical systems have been an important area of study
+especially in fields of atmospheric sciences. However, mathematical
+models are not always available, and if they are, very small errors in
+measurements can amplify very quickly making the prediction useless.
+
+Deep learning techniques are known to be able to approximate any
+function given enough data and resources
+\cite{cybenko1989approximation}. 
+In light of this, the hope is that a deep learning based
+model, the Echo State Network, will be able to capture the 
+complexity of chaotic systems and forecast accurately.
+
+Using a tool created by me to perform the analysis, \echonn, I will
+explore this problem and apply the Echo State Network to three classic
+chaotic problems. The python module can numerically solve these 
+dynamical systems, provides an echo state network implemented by me,
+and has classes to help with experimentation and visualization.
+
+One of the first chaotic systems is
+Henri Poincaré's solution to the restricted circular three body
+problem, a very useful simplification of the popular three body
+problem \cite{chenciner2015poincare}. The second system will be the
+Lorenz equation, researched by the father of chaos, Edward Lorenz
+\cite{lorenz1963deterministic}. 
+The system is commonly known as the Lorenz Butterfly for its shape, and
+it's the origin of the butterfly effect commonly used in pop culture.
+The last dynamical system is the double pendulum. This system is very
+commonly cited for its very random looking behavior despite being a
+completely deterministic system.
+
+Chaos of the systems will be measured using the systems' maximum
+Lyapunov Characteristic Exponents, and data for the systems is
+collected from numerical solutions of the dynamical system's governing
+ODE equations and randomly generated initial conditions.
 
@@
 
-% TODO: Graph of converging LCE
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.6\linewidth]{doc/paper/images/lorenz/lce_converge.png}
+    \caption{Convergence of LCE with large $t$}
+    \label{fig:lce_conv_lorenz}
+\end{figure}
 
@@
     \centering
-    \begin{subfigure}[b]{0.6\textwidth}
+    \begin{subfigure}[b]{0.45\textwidth}
         \includegraphics[width=\textwidth]{images/lorenz_nd_plot.png}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.6\textwidth}
+    \begin{subfigure}[b]{0.45\textwidth}
         \includegraphics[width=\textwidth]{images/lorenz_3d_plot.png}
@@
 which many have analyzed; most notably, Herni Poincaré wrote many papers
-developing the area significantly \cite{chenciner2000remarkable}. One of
+developing the area significantly \cite{chenciner2015poincare}. One of
 Poincaré's solutions, known as the circular restricted 3-body problem,
@@
 
-The N body problem is defined as follows \cite{chenciner2000remarkable}:
+The N body problem is defined as follows \cite{chenciner2015poincare}:
 
@@
 the gravitational constant.
+
+An interesting solution (originally calculated by Carles Simó) can be
+found where the trajectories of the bodies
+are periodic and form a figure 8 \cite{chenciner2000remarkable}.
+
+The gravitational constant, $G$ was redefined to equal 1, the masses of the 
+objects were equal to 1, and the initial conditions are as follows:
+
+\begin{align*}
+    \rv_1 &= (0.97000436, -0.24308753) \\
+    \rv_2 &= -\rv_1 \\
+    \rv_3 &= (0, 0) \\
+    \rv_1' &= -\rv_3 / 2 \\
+    \rv_2' &= -\rv_3 / 2 \\
+    \rv_3' &= (-0.93240737, -0.86473146) \\
+\end{align*}
+
+However, the resulting solution doesn't demonstrate chaotic properties. The LCE 
+convergence chart (Figure \ref{fig:lce_fig_8}) shows the LCE converging to zero.
+
+\begin{figure}[H]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/3body/fig8_short.png}
+        \caption{Figure 8 solution to 3 body problem}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\linewidth]{doc/paper/images/3body/lce_converge.png}
+        \caption{Nonchaotic LCE convergence}
+        \label{fig:lce_fig_8}
+    \end{subfigure}
+    \caption{Nonchaotic Figure 8 Solution to 3 Body}
+\end{figure}
 
@@
 equations myself from the assumptions made by Henri Poincaré outlined in
-\cite{chenciner2000remarkable}.
+\cite{chenciner2015poincare}.
 
@@
     \centering
-    \includegraphics[width=.5\linewidth]{doc/paper/images/lorenz/full_differential.png}
-    \caption{A plot of the differences between time steps of an approximation of the Lorenz Butterfly}
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\linewidth]{doc/paper/images/lorenz/full_differential.png}
+        \caption{A plot of the differences between time steps}
         \label{fig:lorenz_diff}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/lorenz/lce_converge.png}
+        \caption{LCE converges to 0.935 for the Lorenz system}
+    \end{subfigure}
+    \caption{The difference butterfly and LCE convergence}
 \end{figure}
@@
     \centering
-    \includegraphics[width=.5\linewidth]{doc/paper/images/doub_pend/full_differential.png}
-    \caption{The entire difference dataset}
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\linewidth]{doc/paper/images/doub_pend/full_differential.png}
+        \caption{A plot of the differences between time steps}
         \label{fig:doub_pend_diff}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/doub_pend/lce_converge.png}
+        \caption{LCE converges to 0.601 for the Double Pendulum system}
+    \end{subfigure}
+    \caption{The difference graph and LCE convergence}
 \end{figure}
@@
     \centering
-    \includegraphics[width=.5\linewidth]{doc/paper/images/r3body/full_differential.png}
-    \caption{The entire difference dataset}
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\linewidth]{doc/paper/images/r3body/full_differential.png}
+        \caption{A plot of the trajectory of third body}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{doc/paper/images/r3body/lce_converge.png}
+        \caption{The LCE is 0.176 for the Restricted 3 Body Problem}
+    \end{subfigure}
+    \caption{The orbital pattern of the restricted 3 body problem}
 \end{figure}
@@
 \section{Conclusion}
+
+The degree of chaotic behavior was measured by the largest Lyapunov
+Characteristic Exponent of the system, and three dynamical systems were
+used: Lorenz's convection equation, the Double Pendulum system,
+and Henri Poincaré's solution to the Restricted Circular Three Body
+Problem. The performance of the Echo State Network was determined from
+RMSE of the trajectory after some time. The number of the Lyapunov
+times that the model was able to predict accurately determined strength
+of the ESN.
+
+The Lorenz equation's trajectory was reported to have an LCE of 0.935,
+and the predictability of the model was around 1 Lyapunov time for the
+best performers. In general, the ESN forecasted poorly on the Lorenz
+model due to the shape of the attractor. The ESN struggled to predict
+accurately which "wing" to follow and usually diverged at the
+intersection point.
+
+The double pendulum performed the worst with an LCE of 0.601.
+Its best predictors were
+unable to predict at least one Lyapunov time. Much of the pendulums'
+motion was small with occasionally very large spikes of motion. There
+wasn't enough data for the ESN to capture the complexity of the signals
+which trigger the large spikes of motion. When the ESN begins to
+predict spikes (often incorrectly), the RMSE increases significantly.
+It did however produce the correlation found between dimensions, so
+some of the pendulum's patterns were captured successfully.
+
+The restricted three body problem performed very well with a much
+smaller LCE value of 0.176. Similar to the Lorenz, there was an
+apparent pattern, but in this case, only a single point for the
+trajectory to orbit around. Due to this, the performance of the ESN was
+much better. Unbias testing RMSE would remain below 0.1 for several 
+Lyapunov times, similar to the results of another paper with a small 
+maximum LCE value.
+
+It's possible that the ESN model needs more data to train on, or 
+other methods need to be applied to improve the model's performance,
+but it seems that highly chaotic systems can't be predicted very
+easily, but lesser chaotic systems with attractor patterns can be
+forecasted with more success.
 

### 2019-04-24 doc/paper/paper.tex
add abstract

@@
 \begin{abstract}
-This is a paper about chaos.
+
+The predictive power of the recurrent neural networks known as the Echo
+State Network is not as strong for chaotic problems. Measured by
+Lyapunov Characteristic Exponents (LCE), a time series forecasting 
+analysis was applied to highly chaotic systems: Lorenz System, Double
+Pendulum, and Poincaré's solution to the restricted circular three body
+problem. The Lorenz system had an LCE of 0.935 with a predictability
+length of 1 Lyapunov time. The double pendulum had an LCE of 0.601, and
+a predictability length of less than 1 Lyapunov time. The restricted
+circular three body problem solution had an LCE of 0.176 and a
+predictability power as strong as 10 Lyapunov times. 
+Other methods may be applied to improve the results, but in general
+it seems the predictability is very low for very chaotic systems.
+
 \end{abstract}

### 2019-05-01 doc/paper/paper.tex
paper in final version

@@
 
-The predictive power of the recurrent neural networks known as the Echo
-State Network is not as strong for chaotic problems. Measured by
+The predictive power of the recurrent neural network known as the Echo
+State Network is not very strong for very chaotic problems. Measured by
 Lyapunov Characteristic Exponents (LCE), a time series forecasting 
@@
 a predictability length of less than 1 Lyapunov time. The restricted
-circular three body problem solution had an LCE of 0.176 and a
-predictability power as strong as 10 Lyapunov times. 
+circular three body problem solution had an LCE of 0.176, much less
+chaotic, and a predictability power as strong as 10 Lyapunov times. 
 Other methods may be applied to improve the results, but in general
-it seems the predictability is very low for very chaotic systems.
+it seems the predictability is very low for more chaotic systems.
 
@@
 where $\sigma_1$ is the first singular value of the matrix $\frac{\partial
-\xv(t)}{\partial \xv_0}$ and $\log$ is the natural logarithm. Since we can
+\xv(t)}{\partial \xv_0}$ and $\log$ is the natural logarithm. The notation of 
+$\frac{\partial \xv(t)}{\partial \xv_0}$, to me, is unfamiliar. I
+believe $\xv_0$ is a vector representing the initial conditions,
+but not a specific point in $\mathbb{R}^n$, since that
+would leave the partial derivative meaningless. Regardless, it will
+be defined as the matrix obtained from the calculation outlined below.
+Since we can
 only numerically approximate the limit for $\lambda_1$, we will need to use a
@@
 \begin{equation}
-    \der{\Mv(t)} = \frac{\partial \fv}{\partial \xv} \Mv(t)
+    \der{\Mv(t)} = \frac{\partial \fv }{\partial \xv}(\xv(t)) \Mv(t)
     \label{eq:lce_mat_diffeq}
